---
title: "NEPA Clean Energy Analysis: Deliverable 2"
subtitle: "Programmatic & Tiered Reviews"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    toc-location: left
    code-fold: true
    fig-align: center
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(tidyverse)
library(knitr)
library(kableExtra)
library(arrow)
library(here)

reviews_path <- here("data", "analysis", "projects_reviews.parquet")
reviews <- read_parquet(reviews_path) %>% as_tibble()
```

## Executive Summary

::: {.callout-tip}
## Key Findings
Programmatic and tiered reviews are **rare** in the clean energy EA/EIS subset, but they are detectable with high confidence. Any statistical comparison (e.g., timelines) will rely on a **small sample**, so results should be framed as exploratory.
:::

**Snapshot (EA/EIS clean energy subset; n = 1,416):**
```{r}
#| label: tbl-review-type-snapshot
#| tbl-cap: "Review type snapshot within the EA/EIS clean energy subset"

reviews %>%
  count(project_review_type, name = "Count") %>%
  mutate(Share = Count / sum(Count)) %>%
  arrange(desc(Count)) %>%
  mutate(
    Share = scales::percent(Share, accuracy = 0.1),
    project_review_type = str_to_sentence(project_review_type)
  ) %>%
  kable(align = c("l", "r", "r")) %>%
  kable_styling(full_width = FALSE, position = "left") %>%
  footnote(
    general = "“Standard” means a typical, stand‑alone EA/EIS review that is not programmatic and does not explicitly tier from a programmatic review.",
    general_title = "Note:"
  )
```

**Detection quality:**
- Confidence is **high** for 1,413 projects; **medium** for 3
- Detection source is primarily **text regex** (1,413) with a small number via **title** (3)

**Interpretation for Thursday (included here for quick briefing):**
Programmatic reviews appear mainly in large, multi‑project planning efforts (solar/wind/nuclear programs). Tiered reviews are uncommon but present, often tied to a specific programmatic EIS/EA. Because sample sizes for tiered reviews are small, any timeline comparison should be framed as exploratory until the timeline data are cleaned.

---

# Examples

**Programmatic review examples (titles pulled from data):**

```{r}
#| label: tbl-programmatic-examples
#| tbl-cap: "Programmatic review examples (from project titles)"

reviews %>%
  filter(project_review_type == "programmatic") %>%
  distinct(project_title) %>%
  slice_head(n = 8) %>%
  mutate(project_title = str_wrap(project_title, width = 80)) %>%
  rename(`Project title` = project_title) %>%
  kable(align = "l") %>%
  kable_styling(full_width = FALSE, position = "left")
```

**Tiered review examples (project, tiers‑from label, and evidence text):**

```{r}
#| label: tbl-tiered-examples
#| tbl-cap: "Tiered review examples with evidence text from the source document"

reviews %>%
  filter(project_review_type == "tiered") %>%
  mutate(
    tiers_from_clean = if_else(
      str_detect(project_review_tiers_from, regex("PEIS|Programmatic|EIS|EA", ignore_case = TRUE)),
      project_review_tiers_from,
      NA_character_
    ),
    evidence_text = coalesce(project_review_tiers_from_context, project_review_match_text)
  ) %>%
  filter(!is.na(tiers_from_clean)) %>%
  transmute(
    `Project title` = str_wrap(project_title, 70),
    `Tiers from` = str_wrap(tiers_from_clean, 60),
    `Evidence text` = str_wrap(evidence_text, 80)
  ) %>%
  slice_head(n = 8) %>%
  kable(align = c("l", "l", "l")) %>%
  kable_styling(full_width = FALSE, position = "left")
```

---

# Summary Analyses to Present

Distribution by agency (where programmatic/tiered appear most):

```{r}
#| label: tbl-agency-programmatic-tiered
#| tbl-cap: "Top agencies associated with programmatic or tiered reviews"
reviews %>%
  filter(project_review_type %in% c("programmatic", "tiered")) %>%
  count(lead_agency, project_review_type, name = "Count") %>%
  group_by(lead_agency) %>%
  summarise(Total = sum(Count), .groups = "drop") %>%
  arrange(desc(Total)) %>%
  slice_head(n = 6) %>%
  kable(align = c("l", "r")) %>%
  kable_styling(full_width = FALSE, position = "left")
```

---

# Data Source

Counts and examples are derived from `data/analysis/projects_reviews.parquet` for clean energy EA/EIS projects.
