---
title: "NEPA Clean Energy Analysis: Deliverable 2"
subtitle: "Programmatic & Tiered Reviews"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    toc-location: left
    code-fold: true
    fig-align: center
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(tidyverse)
library(knitr)
library(kableExtra)
library(arrow)
library(here)

reviews_path <- here("data", "analysis", "projects_reviews.parquet")
reviews <- read_parquet(reviews_path) %>% as_tibble()
```

::: {.callout-important}
## Key Decisions
- [ ] Evaluate whether 26 captured programmatic and tiered reviews are indeed captured correctly (see For Manual Validation below)
- [ ] "Programmatic Collaborations" will refer to grants/agreements with state energy offices, and are not programmatic environmental reviews (PEIS/PEA)?

:::


## Executive Summary

::: {.callout-note}
## Key Findings
Programmatic and tiered reviews are **rare** in the clean energy EA/EIS subset, but they are detectable with high confidence. Any statistical comparison (e.g., timelines) will rely on a **small sample**, so results should be framed as exploratory.
:::

::: {.callout-tip}
## Next steps
- [ ] Add one potential project ("3621210fbd086bddbbf6fbedc1d6a488") that was captured using regex on file names. 
- [ ] Merge with timeline data once completed to determine speed of completion. 
:::


This report aims to create the following set of deliverables: 

> Data on programmatic and tiered reviews: how many tiered reviews are there compared to total and are they completed faster



## What I've done so far

We developed a multi-tier extraction approach to identify programmatic and tiered reviews within the NEPATEC 2.0 clean energy EA/EIS corpus (n = 1,416 projects):

1. **Title-based detection**: Scanned project titles for explicit programmatic indicators ("programmatic", "PEIS", "PEA")

2. **Regex extraction from documents**: Searched the first 30 pages of each project's documents for tiering language patterns (e.g., "this EA tiers from...", "pursuant to the... Programmatic EIS"). We included false-positive filtering to exclude common non-NEPA uses of "tier" (EPA engine standards, road classifications, pricing tiers)

3. **LLM classification (optional)**: For ambiguous cases with medium-confidence regex matches, an LLM can refine the classification (not used in this initial extraction since there are only a few so far)

### Search Patterns Used

The following regex patterns were used to identify programmatic and tiered reviews:

**Programmatic Review Detection (in titles):**

| Pattern | Matches |
|---------|---------|
| `programmatic` | Explicit "programmatic" keyword |
| `program-wide` or `programwide` | Program-wide reviews |
| `PEIS` | Programmatic Environmental Impact Statement |
| `PEA` | Programmatic Environmental Assessment |

**Tiered Review Detection (in document text):**

| Pattern Type | Example Match |
|--------------|---------------|
| Direct tiering statement | "this EA tiers from the..." |
| Tiering to programmatic | "tiers to the...PEIS" |
| Pursuant to | "pursuant to the...Programmatic EIS" |
| Site-specific tiering | "site-specific EA that tiers from..." |
| PEIS/PEA reference with tiering | "2012 Solar PEIS...tier" |

**False Positive Exclusions:**

| Excluded Pattern | Reason |
|------------------|--------|
| `EPA Tier 1-4` | EPA engine emission standards |
| `Tier 1-4 engine/equipment` | Equipment classifications |
| `first-tier`, `top-tier` | Ranking language |
| `Tier 1-3 Roads` | Road classifications |
| `tiered pricing/rate/system` | Non-NEPA tiering concepts |

---

**Snapshot (EA/EIS clean energy subset; n = 1,416):**
```{r}
#| label: tbl-review-type-snapshot
#| tbl-cap: "Review type snapshot within the EA/EIS clean energy subset"

reviews %>%
  count(project_review_type, name = "Count") %>%
  mutate(Share = Count / sum(Count)) %>%
  arrange(desc(Count)) %>%
  mutate(
    Share = scales::percent(Share, accuracy = 0.1),
    project_review_type = str_to_sentence(project_review_type)
  ) %>%
  kable(align = c("l", "r", "r")) %>%
  kable_styling(full_width = FALSE, position = "left") %>%
  footnote(
    general = "“Standard” means a typical, stand‑alone EA/EIS review that is not programmatic and does not explicitly tier from a programmatic review.",
    general_title = "Note:"
  )
```

**Detection quality:**

- Confidence is **high** for 1,413 projects; **medium** for 3
- Detection source is primarily **text regex** (1,413) with a small number via **title** (3)


---

# Summary Analyses

## Review Type Distribution

@fig-review-type-distribution shows the overall distribution of review types. The vast majority of clean energy EA/EIS projects (98%) are standard reviews—standalone environmental assessments or impact statements that do not tier from or constitute a programmatic review. This is expected: programmatic reviews are by design rare umbrella documents, and tiered reviews require a programmatic review to exist first.

```{r}
#| label: fig-review-type-distribution
#| fig-cap: "Distribution of review types across clean energy EA/EIS projects"

reviews %>%
  count(project_review_type) %>%
  mutate(project_review_type = fct_reorder(project_review_type, n)) %>%
  ggplot(aes(x = project_review_type, y = n, fill = project_review_type)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = scales::comma(n)), hjust = -0.1, size = 3.5) +
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  scale_fill_manual(values = c(
    "standard" = "gray70",
    "programmatic" = "#2171b5",
    "tiered" = "#6baed6"
  )) +
  labs(x = NULL, y = "Number of projects") +
  theme_minimal() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  )
```

## Review Type by NEPA Process

@fig-review-type-by-process breaks down the review types by whether the project went through an Environmental Assessment (EA) or Environmental Impact Statement (EIS). Programmatic reviews appear in both EA and EIS processes, while tiered reviews in this dataset are predominantly EAs—consistent with the idea that tiered reviews leverage prior programmatic analysis to streamline site-specific assessments.

```{r}
#| label: fig-review-type-by-process
#| fig-cap: "Review type distribution by NEPA process (EA vs EIS)"

reviews %>%
  count(dataset_source, project_review_type) %>%
  ggplot(aes(x = dataset_source, y = n, fill = project_review_type)) +
  geom_col(position = "dodge") +
  geom_text(
    aes(label = n),
    position = position_dodge(width = 0.9),
    vjust = -0.3,
    size = 3
  ) +
  scale_fill_manual(
    values = c(
      "standard" = "gray70",
      "programmatic" = "#2171b5",
      "tiered" = "#6baed6"
    ),
    name = "Review Type"
  ) +
  labs(x = "Process Type", y = "Number of projects") +
  theme_minimal()
```

## Distribution by Agency

@tbl-agency-programmatic-tiered shows which agencies have the most programmatic or tiered reviews in the clean energy subset.

```{r}
#| label: tbl-agency-programmatic-tiered
#| tbl-cap: "Top agencies associated with programmatic or tiered reviews"
reviews %>%
  filter(project_review_type %in% c("programmatic", "tiered")) %>%
  count(lead_agency, project_review_type, name = "Count") %>%
  group_by(lead_agency) %>%
  summarise(Total = sum(Count), .groups = "drop") %>%
  arrange(desc(Total)) %>%
  slice_head(n = 6) %>%
  kable(align = c("l", "r")) %>%
  kable_styling(full_width = FALSE, position = "left")
```

---

# Examples

## Programmatic Reviews

@tbl-programmatic-examples shows 5 sample of projects classified as programmatic reviews, with the evidence text that triggered the classification. 

```{r}
#| label: tbl-programmatic-examples
#| tbl-cap: "Programmatic review examples with evidence text"

reviews %>%
  filter(project_review_type == "programmatic") %>%
  slice_head(n = 5) %>%
  transmute(
    `Project title` = str_wrap(project_title, 60),
    `Evidence text` = paste0("\"", str_wrap(project_review_match_text, 50), "\"")
  ) %>%
  kable(align = c("l", "l")) %>%
  kable_styling(full_width = FALSE, position = "left")
```

## Tiered Reviews

@tbl-tiered-examples shows 5 sample of projects classified as tiered reviews, including which programmatic review they tier from and the evidence text.

```{r}
#| label: tbl-tiered-examples
#| tbl-cap: "Tiered review examples with evidence text"

reviews %>%
  filter(project_review_type == "tiered") %>%
  slice_head(n = 5) %>%
  transmute(
    `Project title` = str_wrap(project_title, 50),
    `Tiers from` = str_wrap(project_review_tiers_from, 40),
    `Evidence text` = paste0("\"", str_wrap(project_review_match_text, 50), "\"")
  ) %>%
  kable(align = c("l", "l", "l")) %>%
  kable_styling(full_width = FALSE, position = "left")
```

### For Manual Validation

I've have prepared a Google Sheet for manual review of the 26 identified programmatic and tiered reviews:

**[NEPA Reviews Validation Sheet](https://docs.google.com/spreadsheets/d/1d25Arj2IFR3SLcgv8B6tPdLbs2cDZtC_IjMhZUcueBA/edit?usp=sharing)**

The sheet contains two tabs:

- **Programmatic** (16 projects): Reviews that ARE programmatic EIS/EAs
- **Tiered** (10 projects): Reviews that tier FROM a programmatic review


In the programmatic tab, there appears to be two sets of duplicates. While they have the same project title, they have unique `project_ids`. So, I'll have to manually look at them to determine whether they are actually duplicates or not. Same the tiered tab -- except there is just one set of duplicates.  

### Key Variables for Reviewers

| Variable | Description | What to check |
|----------|-------------|---------------|
| `evidence_text` | The actual text that triggered our classification | Does this text clearly indicate a programmatic or tiered review? |
| `tiers_from` | (Tiered tab only) The programmatic review being referenced | Is this a real PEIS/PEA reference? |
| `evidence_context` | (Tiered tab only) Surrounding text for context | Does the context support the classification? |
| `correct` | Your assessment | Enter Yes / No / Unsure |
| `notes` | Any additional comments | Note false positives or edge cases |

::: {.callout-note}
## Note on Programmatic Tab
The **Programmatic** tab currently does not include the `evidence_context` variable (surrounding text). This will be added in the next extraction run. For now, the `evidence_text` column shows the matched text that triggered classification.
:::

---

# Data Source

Counts and examples are derived from `data/analysis/projects_reviews.parquet` for clean energy EA/EIS projects.
