---
title: "NEPA Decarbonization Technology Analysis: Deliverable 4"
subtitle: "Review Timelines for Categorical Exclusions, Environmental Assessments, and Environmental Impact Statements"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    toc-location: left
    fig-align: center
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(arrow)
library(dplyr)
library(tidyr)
library(readr)
library(stringr)
library(scales)
library(knitr)
library(gt)
library(here)

OUTPUT_DIR <- here("phase2", "output", "deliverable04")
FIG_DIR    <- file.path(OUTPUT_DIR, "figures")
DIAG_DIR   <- file.path(OUTPUT_DIR, "diagnostics")

# CATF brand navy (Phase 1 styling)
CATF_NAVY <- "#012169"

fmt_n   <- function(x) comma(x, accuracy = 1)
fmt_pct <- function(x, accuracy = 0.1) percent(x, accuracy = accuracy)
include_fig <- function(name) knitr::include_graphics(file.path(FIG_DIR, name))
read_diag <- function(name) read_csv(file.path(DIAG_DIR, name), show_col_types = FALSE)
kable_tbl <- function(x) knitr::kable(x, format = "pipe", align = "l")

# Navy-header gt table in the house style
navy_gt <- function(df, title = NULL) {
  g <- gt(df)
  if (!is.null(title)) g <- tab_header(g, title = title)
  g |>
    cols_align(align = "left") |>
    tab_style(style = list(cell_fill(color = CATF_NAVY),
                           cell_text(color = "white", weight = "bold", align = "left")),
              locations = cells_column_labels()) |>
    tab_options(table.font.size = px(12), table.width = pct(100))
}

# --- Coverage (complete = complete_clear + complete_with_proxy) ---
cov <- read_diag("d4_coverage_by_process.csv")
cov_tot <- cov |> group_by(process_type) |> summarise(total = sum(n), .groups = "drop")
cov_complete <- cov |>
  filter(timeline_status %in% c("complete_clear", "complete_with_proxy")) |>
  group_by(process_type) |> summarise(complete = sum(n), .groups = "drop") |>
  left_join(cov_tot, by = "process_type") |>
  mutate(pct = complete / total)
cpct <- function(pt) cov_complete$pct[cov_complete$process_type == pt]
cn   <- function(pt) cov_complete$complete[cov_complete$process_type == pt]

diag <- read_diag("d4_coverage_diagnostics.csv")
dpct <- function(pt, col) diag[[col]][diag$process_type == pt] / 100

# --- Durations (complete_clear) ---
dur <- read_diag("d4_duration_summary.csv")
dmed <- function(pt) dur$median_days[dur$process_type == pt]

# --- Pages (regulatory, all EA/EIS, all energy types) ---
pg <- read_diag("d4_pages_summary.csv")
pgv <- function(pt, per, col) pg[[col]][pg$process_type == pt & pg$fra_period == per]
```

## Executive Summary

::: callout-note
## Key Findings

-   **Timeline completeness varies sharply by review process.** After LLM adjudication, complete timelines (both an initiation and a decision date) reach **`r fmt_pct(cpct("CE"))` for CE**, **`r fmt_pct(cpct("EA"))` for EA**, and **`r fmt_pct(cpct("EIS"))` for EIS** (see @fig-coverage). The binding constraint differs by process: CE and EA are *initiation*-limited (no Notice of Intent requirement, often no recorded start), while EIS is *decision*-limited (Records of Decision are frequently separate documents absent from the corpus — EIS decision coverage is only `r fmt_pct(dpct("EIS","pct_decision"))`).
-   **Durations follow the expected complexity ladder.** Median review duration is **`r dmed("CE")` days for CE**, **`r dmed("EA")` days (~`r round(dmed("EA")/30.4)` months) for EA**, and **`r fmt_n(dmed("EIS"))` days (~`r round(dmed("EIS")/365.25,1)` years) for EIS** (all complete timelines; see @fig-duration-intervals).
-   **LLM adjudication added ~13,000 dates** across 11,207 routed projects at a measured cost of **\$18.20**, lifting complete coverage roughly **+7 percentage points** in every process versus the deterministic-only pipeline.
-   **Timeline outliers split into genuinely long reviews and extraction errors.** The genuinely long cases — e.g. **SunZia Southwest Transmission (14.6 yr)**, **Energia Sierra Juarez**, **Grain Belt Express**, **Cushman Hydroelectric** — are strong candidates for CATF case studies. Errors (an "initiation" anchored to a historical citation) are flagged and excluded (see @sec-outliers).
-   **Post-FRA EIS documents are shorter; EA documents are essentially unchanged.** Across **all `r fmt_n(sum(pg$n_projects))` EA/EIS projects** (all energy types, regulatory pages computed from the full page text), mean regulatory EIS length fell from **`r pgv("EIS","Pre-FRA","mean_pages")` to `r pgv("EIS","Post-FRA","mean_pages")` pages** after the Fiscal Responsibility Act, while EA length held near **`r pgv("EA","Pre-FRA","mean_pages")`–`r pgv("EA","Post-FRA","mean_pages")` pages** — though the post-FRA sample is small (see @sec-pages).
:::

This report delivers:

> Timelines for categorical exclusions, environmental assessments, and environmental impact statements — including segmentation by regulatory period (pre/post-FRA), energy category, and decision year — plus a timeline-outlier deliverable for case-study investigation and a document-length / FRA page-limit analysis refreshed on Phase 2 decision dates.

------------------------------------------------------------------------

# Methodology

This deliverable **extends the [Phase 1 timeline analysis](https://www.kaseyzapatka.com/nepa/phase1/reports/deliverable03.html#timeline-analysis-for-decarbonization-technologies)** — which examined review timelines for *clean-energy projects only* — to the **entire NEPATEC 2.0 database**: every categorical exclusion, environmental assessment, and environmental impact statement, across clean-energy, fossil, and other project types. It does so with a rebuilt extraction pipeline that adds authoritative agency dates and two machine-learning models.

## How timeline dates are extracted

This section improves on the Phase 1 timeline method. For every project we identify two milestones — the **initiation** (when the review begins: an application received, a Notice of Intent, or scoping) and the **decision** (when it concludes: a signature, a Record of Decision, or a Finding of No Significant Impact). The pipeline finds these in five high-level steps:

1.  **Extract candidate dates** from every NEPA document in NEPATEC 2.0, capturing the sentence around each date so its meaning is clear.
2.  **Classify each date's role** — a machine-learning *classifier* labels every candidate as an initiation, decision, or other date.
3.  **Rank the candidates** — a second *ranking model* scores which initiation candidate and which decision candidate is the operative one for the project.
4.  **Anchor on authoritative dates** — where the BLM or DOE NEPA register records an official project-start or decision date, that date is used.
5.  **Resolve the hardest cases with an LLM** — for the projects where the operative date is still ambiguous, a large language model selects the best remaining candidate.

A project's timeline is **complete** when both an initiation and a decision date are found.

## How Phase 2 analysis improves upon Phase 1

-   **Full-database coverage** — Phase 1 covered clean-energy projects only (~20,700); Phase 2 covers the **entire NEPATEC 2.0 database** (~61,000 projects — CE, EA, and EIS across clean-energy, fossil, and other types).
-   **Authoritative agency dates** — Phase 2 incorporates the **BLM and DOE NEPA registers** as a first-class source of official project-start and decision dates, instead of relying solely on dates parsed from document text.
-   **A classifier *and* a ranking model** — Phase 1 used a single date classifier; Phase 2 adds a dedicated **ranking model** that decides which candidate is the operative initiation/decision, which sharply improves selection on documents containing many look-alike dates.
-   **LLM adjudication at scale** — the hardest ambiguous cases across the whole corpus are resolved by a large language model, recovering dates the automated pipeline could not select on its own.
-   **Cleaner retrieval and granularity handling** — section-aware retrieval pulls fewer but cleaner candidate dates, and month-only decision dates (e.g. a Final-EIS cover month) are handled explicitly rather than dropped.
-   **Fully reproducible** — the entire build runs end-to-end from committed code on a DuckDB-based pipeline.

::: callout-note
## Next steps

-   **Lift EIS decision coverage** — re-pull Final-EIS cover/publication dates and add a recovery pass for Records of Decision that sit in separate documents (the main EIS gap).
-   **Recover missing initiations** — cross-reference the **Federal Register Notice-of-Intent API** to fill in and validate EIS/EA start dates.
-   **Validate date *accuracy*** — build a held-out, human-labelled gold set to measure precision, not only coverage.
-   **Fix the ordering check at the source** — resolve the 233 decision-before-initiation rows in the selection step (currently handled by a downstream stopgap filter; see Data Quality & Caveats).
-   **Extend the analysis** — BLM field-office learning curves, geothermal timelines by field office, and whether post-FRA adoption of categorical exclusions correlates with faster reviews (see Requested Extensions).
:::

------------------------------------------------------------------------

# Timeline Analysis

## Timeline completeness

@fig-complete-share shows the share of projects with complete timelines by review process; @fig-coverage breaks down *which* component is missing. The patterns differ by process:

-   **CE** (`r fmt_pct(cpct("CE"))` complete): decision coverage is high (`r fmt_pct(dpct("CE","pct_decision"))`) but initiation coverage is low (`r fmt_pct(dpct("CE","pct_initiation"))`) — most CEs are single-signature determinations with no recorded start.
-   **EA** (`r fmt_pct(cpct("EA"))` complete): the strongest of the three, but still initiation-limited — EAs have no Notice-of-Intent requirement and often skip scoping.
-   **EIS** (`r fmt_pct(cpct("EIS"))` complete): initiations are usually documented (an NOI is required), but **decision** coverage is only `r fmt_pct(dpct("EIS","pct_decision"))` because Records of Decision are frequently separate documents not in the corpus.

**Comparison with Phase 1.** Phase 1 measured completeness for *clean-energy projects only* and reported [CE ~30% · EA ~62% · EIS ~48%](https://www.kaseyzapatka.com/nepa/phase1/reports/deliverable03.html#fig-complete-share). On the **full database**, Phase 2 reaches CE `r fmt_pct(cpct("CE"))` · EA `r fmt_pct(cpct("EA"))` · EIS `r fmt_pct(cpct("EIS"))`. CE is *higher* — the much larger universe is dominated by routine, register-dated determinations — but **EA and EIS land below their Phase 1 clean-only rates**, for two reasons this analysis surfaced:

-   **The aggregate is diluted by non-clean projects.** Phase 1's denominator was clean energy; Phase 2's is everything. Decarbonization EIS alone completes at **42.4%** (much closer to Phase 1's 48%) — it is the heterogeneous "Other" bucket that pulls the EIS aggregate down to `r fmt_pct(cpct("EIS"))`.
-   **The remaining gaps are structural, not fixable by better parsing.** EIS completeness is capped by **missing Records of Decision** (a separate document, often outside the corpus — decision coverage only `r fmt_pct(dpct("EIS","pct_decision"))`); CE and EA completeness is capped by **missing initiations**, because neither requires a Notice of Intent and a start date is frequently never recorded. A minority of EIS "projects" are also comment letters or document fragments with no milestone date to extract. Lifting EIS coverage toward Phase 1 is the focus of the [Next steps](#methodology) above.

```{r}
#| label: fig-complete-share
#| fig-cap: "Share of projects with a complete timeline (both initiation and decision dates) by NEPA review process. Dot marks the completion rate."
#| out-width: "90%"
include_fig("fig_d4_complete_timeline_share_boxplot.png")
```

```{r}
#| label: fig-coverage
#| fig-cap: "Timeline coverage breakdown by review process — both dates, decision only, initiation only, or neither."
#| out-width: "100%"
include_fig("fig_d4_coverage_by_process.png")
```

## Duration analysis

Durations use **all complete timelines** (both dates present, including those resolved with a proxy date such as a Final-EIS publication date in place of a missing Record of Decision). Month-only dates are imputed to the **mid-month 15th**; the small number of year-only dates (almost entirely a subset of CE initiation dates) are excluded, since a day cannot be responsibly imputed from a year alone.

On that basis, durations track the expected complexity ladder from CE to EA to EIS — a median of about **`r dmed("CE")` days for CE**, **`r dmed("EA")` days (~`r round(dmed("EA")/30.4)` months) for EA**, and **`r fmt_n(dmed("EIS"))` days (~`r round(dmed("EIS")/365.25,1)` years) for EIS** (n = `r fmt_n(dur$n[dur$process_type=="CE"])` / `r fmt_n(dur$n[dur$process_type=="EA"])` / `r fmt_n(dur$n[dur$process_type=="EIS"])`). @fig-duration-intervals summarises the spread; @fig-duration-hist shows the full distributions; @fig-spans plots individual project spans.

```{r}
#| label: fig-duration-intervals
#| fig-cap: "Duration summary by review process. Thin bar = 10th–90th percentile; thick bar = interquartile range; point = median."
#| out-width: "100%"
include_fig("fig_d4_duration_summary_intervals.png")
```

```{r}
#| label: fig-duration-hist
#| fig-cap: "Duration distribution (histogram) by review process; all complete timelines (month-only dates imputed to mid-month)."
#| out-width: "100%"
include_fig("fig_d4_duration_histogram.png")
```

```{r}
#| label: fig-spans
#| fig-cap: "Individual project timelines from initiation to decision, sorted by duration within process (complete timelines, sampled)."
#| out-width: "100%"
include_fig("fig_d4_project_timeline_spans.png")
```

## Projects by decision year

@fig-by-year shows project volume by decision year, faceted by review process. CE volumes ramp from the late-2000s (ARRA-era funding) with a second rise into the 2020s (BIL/IRA); EIS decisions lag CE and EA by several years, reflecting longer review cycles.

```{r}
#| label: fig-by-year
#| fig-cap: "Projects by decision year, faceted by NEPA review process."
#| out-width: "100%"
include_fig("fig_d4_projects_by_decision_year.png")
```

------------------------------------------------------------------------

# Segmentation

## Regulatory period: pre- vs post-FRA

The **Fiscal Responsibility Act of 2023** (and the 2024 CEQ rule) reset NEPA expectations. @fig-fra compares median durations before and after the rule's effective date; @fig-trend shows the longer trend by decision year. Durations are still accruing for post-FRA projects (reviews initiated recently have not all reached a decision), so post-FRA medians should be read as provisional.

```{r}
#| label: fig-fra
#| fig-cap: "Median review duration, pre- vs post-FRA, by process. Post-FRA samples are small and still accruing."
#| out-width: "100%"
include_fig("fig_d4_fra_comparison.png")
```

```{r}
#| label: fig-trend
#| fig-cap: "Median review duration by decision year and process, with major legislative markers."
#| out-width: "100%"
include_fig("fig_d4_duration_trend.png")
```

## By energy category

Coverage and duration also vary by energy category (Decarbonization / Fossil / Other). Notably, **decarbonization EIS projects complete at 42.4% — well above the EIS aggregate of `r fmt_pct(cpct("EIS"))`** — because clean-energy EISs are better documented than the heterogeneous "Other" bucket that drags the aggregate down.

```{r}
#| label: fig-cov-energy
#| fig-cap: "Timeline coverage by process and energy category (both dates / decision only / initiation only / none)."
#| out-width: "100%"
include_fig("fig_d4_coverage_by_process_and_energy.png")
```

```{r}
#| label: fig-dur-energy
#| fig-cap: "Duration summary intervals by process, coloured by energy category."
#| out-width: "100%"
include_fig("fig_d4_duration_summary_intervals_by_energy.png")
```

```{r}
#| label: fig-fra-energy
#| fig-cap: "Pre/post-FRA median duration by process and energy category."
#| out-width: "100%"
include_fig("fig_d4_fra_comparison_by_energy.png")
```

## Date provenance

@fig-source shows how many selected dates came from authoritative registers (BLM/DOE) versus document text — a transparency check on the timeline's foundations.

```{r}
#| label: fig-source
#| fig-cap: "Source path of selected dates at the project level (register vs document text)."
#| out-width: "100%"
include_fig("fig_d4_register_source_projects.png")
```

------------------------------------------------------------------------

# Timeline Outliers: Case-Study Candidates {#sec-outliers}

Per the deliverable request, timeline outliers are surfaced for CATF staff to investigate via case studies (including whether NEPA itself was a cause of delay). Outliers are init→decision spans **longer than 5,000 days (~13.7 years)**, produced reproducibly by `code/deliverable04/10_outliers.R`.

**The critical distinction is real vs. extraction error.** A multi-year span can be a genuinely protracted review *or* an artifact where the "initiation" was anchored to a historical citation (a facility's construction date, a reserve's founding, a prior plan, a regulatory-compliance milestone) rather than the NEPA start. The two are separable only by reading the evidence text, which the outlier table preserves.

::: callout-note
## How to read these (process matters)

-   **EIS outliers are mostly real.** EIS median duration is already ~3 years; a 14–17-year EIS sits in the legitimate tail (transmission, restoration, resource-management plans genuinely run that long).
-   **CE outliers are mostly errors** — CEs are fast determinations (median `r dmed("CE")` days), so a multi-year CE almost always reflects a mis-extracted initiation. CE outliers are therefore excluded from the client list.
-   **Initiation dates before ~1990 are a strong error signal.** Of the four EIS outliers with 1980s-or-earlier initiations, three are confirmed errors (the "initiation" is a park founding, a prior lock's construction-completion date, or a RCRA compliance milestone).
:::

The clean-energy EA/EIS candidates below are the highest-value case studies. **SunZia** (ROW application Sep 2008 → Record of Decision Apr 2023) is the textbook long-permitting saga; **Energia Sierra Juarez**, **Grain Belt Express**, and **Cushman Hydroelectric** are similar. *Confirmed errors are excluded* — e.g. the Palisades Nuclear Restart "19-year" span glues a 2005 license-renewal application to a 2024 restart Record of Decision (two different NEPA actions).

```{r}
#| label: tbl-outliers
#| tbl-cap: "Genuinely long EA/EIS reviews — case-study candidates. Full list with evidence text in output/deliverable04/diagnostics/d4_duration_outliers_client.csv. Project IDs are full UUIDs."
client <- read_diag("d4_duration_outliers_client.csv")
# Curated exclusions: confirmed cross-action extraction errors found on evidence review.
exclude_ids <- c("da6bbc7a3ca3aeca7bc1c06eaafa8d44")  # Palisades restart (license-renewal init)
client |>
  filter(!project_id %in% exclude_ids) |>
  arrange(desc(duration_years)) |>
  transmute(`Project` = project_title,
            `Process` = process_type,
            `Energy` = project_energy_type,
            `Lead agency` = str_remove_all(lead_agency_harmonized, '\\[|\\]|"'),
            `Years` = round(duration_years, 1),
            `Initiation` = as.character(initiation_date),
            `Decision` = as.character(decision_date),
            `project_id` = project_id) |>
  head(18) |>
  navy_gt()
```

------------------------------------------------------------------------

# Document Length & FRA Page Limits {#sec-pages}

This section reproduces the Phase 1 Deliverable 5 document-length analysis on the **entire Phase 2 EA/EIS corpus, all energy types**, refreshed on Phase 2 LLM-adjudicated **decision dates** (inclusion requires only a decision date — the time axis and FRA classification depend only on it).

::: callout-note
## Method & scope

-   **Regulatory pages** = body word count ÷ 500, excluding embedded appendices and low-content pages, per **40 C.F.R. § 1508.1(bb)** — the measure that matches the FRA limits (EA ≤ 75 pages; EIS ≤ 150, or 300 if extraordinarily complex).
-   These counts are computed **from the actual page text** by `fra/01_extract_pages.py` (a DuckDB pass over `phase2/data/processed/{ea,eis}/pages.parquet` — 6.1M EIS pages — detecting the appendix boundary and low-content pages), producing regulatory pages for **5,032 EA/EIS projects** (2,765 EA, 2,267 EIS) across **all energy types** — not just decarbonization.
-   The pre/post-FRA analysis below uses the **`r fmt_n(sum(pg$n_projects))` projects** that also have a decision date (EA `r fmt_n(sum(pg$n_projects[pg$process_type=="EA"]))` / EIS `r fmt_n(sum(pg$n_projects[pg$process_type=="EIS"]))`).
-   **FRA date = enactment (June 3, 2023)**, matching Phase 1 D5. (The timeline section above uses the CEQ-rule effective date for its period split.)
:::

## Document length over time

```{r}
#| label: fig-pages-time
#| fig-cap: "Regulatory page counts for individual EA/EIS projects (all energy types) by decision date, coloured by FRA period; navy line = 3-month rolling mean. Red dashed line = FRA enactment (June 3, 2023)."
#| out-width: "100%"
include_fig("fig_d4_pages_over_time.png")
```

## Pre vs post-FRA

Across all EA/EIS, **EIS regulatory length fell after FRA** (mean `r pgv("EIS","Pre-FRA","mean_pages")` → `r pgv("EIS","Post-FRA","mean_pages")` pages; median `r pgv("EIS","Pre-FRA","median_pages")` → `r pgv("EIS","Post-FRA","median_pages")`), while **EA length was essentially flat** (mean `r pgv("EA","Pre-FRA","mean_pages")` → `r pgv("EA","Post-FRA","mean_pages")`). The post-FRA window is short, so these are provisional (n = `r pgv("EA","Post-FRA","n_projects")` EA, `r pgv("EIS","Post-FRA","n_projects")` EIS).

```{r}
#| label: tbl-pages
#| tbl-cap: "Regulatory page counts by process and FRA period — all EA/EIS, all energy types. Regulatory pages = body word count ÷ 500."
pg |>
  transmute(Process = process_type, `FRA period` = fra_period, `n` = n_projects,
            Mean = mean_pages, Median = median_pages, SD = sd_pages,
            p25 = p25_pages, p75 = p75_pages) |>
  navy_gt()
```

```{r}
#| label: fig-pages-bar
#| fig-cap: "Mean regulatory pages, pre- vs post-FRA (all EA/EIS). Bar = mean; diamond = median; n shown."
#| out-width: "100%"
include_fig("fig_d4_pages_pre_post_fra.png")
```

```{r}
#| label: fig-pages-dist
#| fig-cap: "Distribution of regulatory page counts, pre- vs post-FRA (all EA/EIS; violin + box; y capped at p99)."
#| out-width: "100%"
include_fig("fig_d4_pages_distribution.png")
```

## By energy category

Segmenting by energy shows the post-FRA EIS decline is broad — it appears in decarbonization **and** fossil EIS projects — while EA length is flat across categories.

```{r}
#| label: tbl-pages-energy
#| tbl-cap: "Regulatory page counts by process, energy category, and FRA period (all EA/EIS). Regulatory pages."
read_diag("d4_pages_summary_by_energy.csv") |>
  transmute(Process = process_type, Energy = energy, `FRA period` = fra_period,
            `n` = n, `Mean pages` = mean_pages, `Median pages` = median_pages) |>
  navy_gt()
```

```{r}
#| label: fig-pages-energy
#| fig-cap: "Mean regulatory pages by energy category, pre- vs post-FRA, all EA/EIS. Labels = n projects."
#| out-width: "100%"
include_fig("fig_d4_pages_pre_post_fra_by_energy.png")
```

## FRA page-limit compliance

```{r}
#| label: fig-pages-compliance
#| fig-cap: "FRA page-limit compliance among post-FRA projects (all EA/EIS, regulatory pages). EA limit 75; EIS 150 (300 if extraordinarily complex)."
#| out-width: "90%"
include_fig("fig_d4_pages_compliance.png")
```

## Why regulatory pages, not raw pages

Raw PDF page counts substantially overstate the length the statute actually limits — they include embedded appendices and sparse pages (covers, maps, dividers). The figure below shows the gap between mean raw and mean regulatory pages; it is the reason a raw count cannot be compared to the 75/150-page limits.

```{r}
#| label: fig-pages-rr
#| fig-cap: "Mean raw PDF pages vs. mean regulatory pages (all EA/EIS). The gap = appendices + sparse pages excluded by 40 C.F.R. § 1508.1(bb)."
#| out-width: "100%"
include_fig("fig_d4_pages_reg_vs_raw.png")
```

------------------------------------------------------------------------

# Data Quality & Caveats

```{r}
#| label: tbl-caveats
tibble::tibble(
  Caveat = c("Date accuracy is unvalidated", "Proxy reliance in 'complete' counts",
             "EIS decision-document gap", "Negative-duration errors (stopgap)",
             "Post-FRA samples are small"),
  `What it means` = c(
    "Coverage is reproducible and audited; date *accuracy* is asserted, not measured against a human gold set (~13k dates this run are LLM-derived).",
    "Among complete timelines, ~55% of CE and ~68% of EIS use a proxy date (e.g. Final-EIS publication standing in for a missing Record of Decision). These are now included in the duration figures, with month-only dates imputed to the mid-month 15th and the small number of year-only dates excluded.",
    "EIS decision coverage is ~42% because Records of Decision are frequently separate documents absent from the corpus; a Final-EIS-publication fallback is used and flagged.",
    "233 decision-before-initiation rows are reclassified to invalid_order at analysis time; a source-level fix in 05 is tracked.",
    "Post-FRA covers ~2 years; pre/post comparisons are exploratory."
  )
) |>
  navy_gt()
```

------------------------------------------------------------------------

# Requested Extensions (Planned / Not Yet Built)

The following items from the deliverable request require additional analysis beyond the current timeline build. They are scoped here for the next phase:

```{r}
tibble::tibble(
  `Requested analysis` = c(
    "BLM field-office learning curve",
    "Geothermal timelines by BLM field office",
    "CE adoption post-FRA vs. review speed",
    "Notice-of-Intent cross-reference (Federal Register API)"
  ),
  `What it needs` = c(
    "Join decision dates to BLM field-office identifiers; test whether per-office median duration falls as cumulative application count rises (experience effect).",
    "Geothermal project flags (Deliverable 1/3 tech tags) × field office × duration; limited by geothermal sample size per office.",
    "CE-adoption events (which agencies adopted other agencies' CEs post-FRA, BLM/DOE/USFS) × subsequent review duration; needs the CE-adoption register.",
    "Pull NOI publication dates from the Federal Register API to recover missing initiation dates (esp. EIS) and validate extracted initiations."
  )
) |>
  navy_gt()
```

------------------------------------------------------------------------

# Reproduction

Run from the repository root in the `nepa` conda environment:

``` bash
# Timeline pipeline (candidates -> classify -> select -> LLM adjudicate)
conda run -n nepa python phase2/code/deliverable04/run_pipeline.py
ANTHROPIC_API_KEY=$(security find-generic-password -s nepa-anthropic -w) \
  conda run -n nepa python phase2/code/deliverable04/06_adjudicate_llm.py \
  --mode candidate_adjudication --process CE EA EIS --model claude-haiku-4-5-20251001 --workers 24

# Analysis + figures
Rscript phase2/code/deliverable04/08_analyze.R       # coverage, durations, FRA, energy
Rscript phase2/code/deliverable04/10_outliers.R      # timeline outliers (case-study candidates)

# FRA document-length analysis (regulatory pages from the full page text, all EA/EIS)
conda run -n nepa python phase2/code/deliverable04/fra/01_extract_pages.py --run
Rscript phase2/code/deliverable04/fra/02_pages_fra.R # document length / FRA page limits
```

Outputs are written to `phase2/output/deliverable04/` (`figures/`, `diagnostics/`).

------------------------------------------------------------------------

*Draft generated `r Sys.Date()` | NEPA Decarbonization Technology Analysis — Phase 2, Deliverable 4*
