---
title: "NEPA Decarbonization Technology Analysis: Deliverable 2"
subtitle: "Determinations of Significance Across Resource Areas"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    toc-location: left
    fig-align: center
execute:
  echo: false
  warning: false
  message: false
---

<!-- Plan: phase2/plans/deliverable02.md (v2.11). Pipeline: phase2/code/deliverable02/
     00 regime -> 01 corpus+cohorts -> 02 FONSI extract -> 04 EIS extract -> 05 validate ->
     06 analyze.R (writes the CSV tables this report reads). Regenerate 06 before rendering. -->

```{r setup}
library(arrow); library(dplyr); library(tidyr); library(readr)
library(stringr); library(knitr); library(gt); library(here); library(scales)
options(arrow.skip_nul = TRUE)

A   <- here("phase2", "data", "analysis", "deliverable02")
TAB <- here("phase2", "output", "deliverable02", "analysis")
rp  <- function(f) read_parquet(file.path(A, f))
rc  <- function(f) { p <- file.path(TAB, f); if (file.exists(p)) read_csv(p, show_col_types = FALSE) else tibble() }
fmt <- function(x) formatC(x, format = "d", big.mark = ",")
fmt_n   <- function(x) comma(x, accuracy = 1)
fmt_pct <- function(x, accuracy = 1) percent(x, accuracy = accuracy)
NAVY <- "#012169"; BLUE <- "#0047BB"
include_fig <- function(f) { p <- file.path(TAB, f); if (file.exists(p)) knitr::include_graphics(p) }

det    <- rp("significance_determinations.parquet")
corpus <- rp("significance_corpus.parquet")
regime <- rp("project_regime.parquet")

# analysis tables written by 06_analyze_significance.R (regenerate before rendering)
val          <- rp("validation_metrics.parquet")
res_by_class <- rc("resource_by_class.csv")
class_dist   <- rc("class_distribution_dual_denominator.csv")
mit_doc      <- rc("mitigation_document_level.csv")
mit_res      <- rc("mitigation_by_resource.csv")
mit_ex       <- rc("mitigation_examples.csv")

# friendly labels for the controlled vocab
class_label <- c(no_significant_impact = "No significant impact",
                 less_than_significant = "Less than significant",
                 less_than_significant_with_mitigation = "LTS with mitigation",
                 significant_adverse = "Significant adverse",
                 significant_unavoidable = "Significant & unavoidable")
res_label <- c(air_quality="Air quality", water="Water", biological="Biological",
               cultural="Cultural / historic", visual="Visual", noise="Noise",
               soils_geology="Soils / geology", socioeconomic="Socioeconomic",
               transportation="Transportation", land_use="Land use",
               climate_ghg="Climate / GHG", public_health="Public health",
               project_wide="Project-wide (FONSI)", unknown="Unplaced")
relabel <- function(x, map) ifelse(is.na(map[x]) | map[x] == "", x, map[x])

is_dry  <- all(det$extraction_method == "regex")
has_val <- nrow(val) > 0 && !is_dry
```

::: {.callout-warning}
## Draft status
```{r}
#| results: asis
cat(if (is_dry)
  paste("This document outlines the deliverable's purpose and the analysis workflow; the",
        "findings themselves are still in progress. Results tables will be added once the",
        "extraction pass and the validation against the hand-labelled gold set are complete.")
  else paste("This report reflects the validated extraction; see Methods & required caveats",
             "for the applicable limitations."))
```
:::

This report delivers:

> Determinations of significance across resource areas; factors that contribute to a determination of "significant impact," starting with mitigated FONSIs.

---

## Context

When a clean-energy project — a solar farm, a transmission line — needs federal approval, the
lead agency (here, BLM or the Department of Energy) studies its environmental impact and makes a
judgment call: **is the impact "significant" or not?**

That one word decides a lot. If the agency finds the impact **not significant**, the project
moves ahead on a faster track (a document called a *Finding of No Significant Impact*, or FONSI).
If it finds the impact **significant**, that triggers a much longer, more demanding review.

Often the agency lands in between: *"this impact **would** be significant — **unless** the
developer commits to specific measures to reduce it."* That is a **mitigated finding**, and it is
the most revealing case, because it shows the exact line the agency drew and what it required to
stay on the safe side of it.

**So this deliverable answers:** across these clean-energy projects, how do agencies decide what
counts as "significant" in each environmental area (air, water, wildlife, cultural sites, and so
on) — and what mitigation do they demand to avoid that label? Answering it means reading and
coding a judgment call out of **thousands** of dense decision documents — far more than anyone
could review by hand. The workflow below is how we do that reliably.

## Workflow

The approach pairs the **scale of automation** with the **reliability of human judgment**:
software and AI do the reading at a scale no person could, while analysts define the scope, hold
the checkpoints, and grade the AI against their own answer key before any result counts.

```{=html}
<div style="max-width:660px;margin:1.25rem auto;">
  <div style="border:2px solid #0047BB;border-radius:8px;padding:.65rem 1rem;background:#f4f7fc;">
    <div style="font-weight:700;color:#012169;font-size:1.05rem;">1 · Scope</div>
    <div style="color:#1A1A2E;">Clean-energy FONSIs led by BLM or the DOE family &mdash; 258 documents across 193 projects, each dated to its regulatory era</div>
  </div>
  <div style="text-align:center;color:#0047BB;font-size:1.5rem;line-height:1.1;">&#8595;</div>
  <div style="border:2px solid #0047BB;border-radius:8px;padding:.65rem 1rem;background:#f4f7fc;">
    <div style="font-weight:700;color:#012169;font-size:1.05rem;">2 · Gather documents</div>
    <div style="color:#1A1A2E;">Pull the FONSI and supporting EA decision documents from the NEPATEC&nbsp;2.0 corpus</div>
  </div>
  <div style="text-align:center;color:#0047BB;font-size:1.5rem;line-height:1.1;">&#8595;</div>
  <div style="border:2px solid #0047BB;border-radius:8px;padding:.65rem 1rem;background:#f4f7fc;">
    <div style="font-weight:700;color:#012169;font-size:1.05rem;">3 · Extract the judgment calls</div>
    <div style="color:#1A1A2E;">Rules flag the finding passages &rarr; Claude (Sonnet&nbsp;5) codes each one &mdash; outcome, resource area, threshold, mitigation &rarr; two analysts plus an adjudicator hand-label 400 as the answer key</div>
  </div>
  <div style="text-align:center;color:#0047BB;font-size:1.5rem;line-height:1.1;">&#8595;</div>
  <div style="border:2px solid #0047BB;border-radius:8px;padding:.65rem 1rem;background:#f4f7fc;">
    <div style="font-weight:700;color:#012169;font-size:1.05rem;">4 · Validate</div>
    <div style="color:#1A1A2E;">Score the AI against the held-out human labels (class agreement 0.81) before any result counts</div>
  </div>
  <div style="text-align:center;color:#0047BB;font-size:1.5rem;line-height:1.1;">&#8595;</div>
  <div style="border:2px solid #012169;border-radius:8px;padding:.65rem 1rem;background:#012169;">
    <div style="font-weight:700;color:#ffffff;font-size:1.05rem;">5 · Analyze and report</div>
    <div style="color:#ffffff;">Validated findings become this report and a source-linked data table &mdash; every number traces to its page</div>
  </div>
  <div style="text-align:center;color:#6b7280;font-size:.85rem;margin-top:.5rem;">
    The five steps of the analysis &mdash; each card matches a numbered step below.
  </div>
</div>
```

**Why the results are trustworthy:** every determination is traceable to the exact document,
page, and passage; the AI is scored against human labels before it is used; and the project set
is scoped so comparisons are fair.

## What the two document types tell us

A significance call can land on either side of the line, and **the two document types record
opposite outcomes** — so we analyze them separately rather than pooling them:

- **FONSIs are the *below-the-line* record.** The agency concluded the impact is **not**
  significant. This is where the **mitigated** cases live — impacts that *would* have been
  significant without committed measures. FONSIs tell us **how agencies stay under the line, and
  what they require to get there.**
- **EISs are the *above-the-line* record.** These are impacts the agency judged **significant**
  (and, at the extreme, significant and *unavoidable*). They tell us **which resources actually
  cross the line.**

This report presents the **FONSI analysis**, which is complete and validated. The EIS analysis
runs the same machinery on a separate corpus; its answer key is already built and it awaits its
extraction pass (see *Scope decisions*).

::: {.callout-note appearance="minimal"}
Because a FONSI is *by definition* a not-significant finding, a "significant" determination inside
one is an anomaly — a *would-be-significant-absent-mitigation* phrasing, a rejected alternative, or
a mis-read. Such cases are a fraction of a percent here and are flagged for review, not reported as
findings. The *significant* side of the story is the EIS analysis.
:::

## FONSI analysis {#sec-fonsi}

```{r}
#| label: fonsi-corpus
proj_rows <- str_detect(class_dist$determination_class, "project-level")
n_prim <- sum(class_dist$n_determinations[!proj_rows])
n_docs <- mit_doc$n_documents[1]
n_proj <- mit_doc$n_projects[1]
```

Across **`r fmt_n(n_proj)` clean-energy projects** (BLM + DOE family), the FONSI corpus yields
**`r fmt_n(n_prim)` distinct significance determinations** — one per document × resource area ×
outcome — drawn from `r fmt_n(n_docs)` decision documents.

::: {.callout-tip}
## Key takeaways (FONSI track)
- **Most clean-energy FONSIs are *mitigated*.** About **`r fmt_pct(mit_doc$share_mitigated_class_signal)`**
  of these documents reach "no significant impact" only because the developer commits to specific
  measures — the impact *would* have been significant otherwise.
- **A handful of resources carry the mitigation.** **Biological resources, soils/geology, water, and
  cultural/historic sites** are where projects most often sit at the line and are pulled under it —
  biological alone is mitigation-dependent in ~30% of its determinations.
- **The below-the-line default is "no significant impact."** Across resource areas, outright
  no-significant-impact is the most common outcome; genuinely *significant* findings live in the EIS
  record, not here.
- **The numbers are validated.** The extraction agrees with a hand-labeled human answer key at
  **0.81** on the significance class (held-out test) before any figure in this report was produced.
:::

### How accurate is the extraction?

Every determination is machine-extracted, so before any number is reported the pipeline is scored
against a **hand-labeled answer key**: two independent reviewers coded a stratified 400-window
sample, a third adjudicated their disagreements, and **30% was held out** as an untouched test. The
held-out column is the honest score.

```{r}
#| label: tbl-validation
metric_lab <- c(candidate_is_determination        = "Finds a determination (per passage)",
                resource_determination_detection  = "Assigns the right resource",
                determination_class_macro_f1       = "Gets the significance class right",
                mitigation_dependent_f1            = "Flags mitigation-dependence",
                primary_threshold_type_accuracy    = "Identifies the regulatory threshold")
val |>
  mutate(score = coalesce(f1, precision)) |>
  filter(scope %in% c("overall", "holdout"), metric %in% names(metric_lab)) |>
  select(metric, scope, score) |>
  pivot_wider(names_from = scope, values_from = score) |>
  mutate(metric = factor(metric, levels = names(metric_lab), labels = unname(metric_lab))) |>
  arrange(metric) |>
  gt() |>
  cols_label(metric = "What the pipeline is scored on", overall = "All 400", holdout = "Held-out test") |>
  fmt_number(columns = c(overall, holdout), decimals = 2) |>
  data_color(columns = holdout, colors = col_numeric(c("#fdecea", "#eaf3ea"), domain = c(0.4, 1))) |>
  cols_align(align = "left", columns = metric) |>
  tab_style(style = list(cell_fill(color = NAVY), cell_text(color = "white", weight = "bold")),
            locations = cells_column_labels()) |>
  tab_options(table.font.size = px(13), table.width = pct(100)) |>
  tab_source_note("Scores are F1 (0–1), except the threshold row, which is accuracy. Held-out = the 30% of the gold set the pipeline never influenced.")
```

In plain terms: the extraction **almost never misses a determination** and places it in the
**right resource area ~89%** of the time; it agrees with the human coders on the significance
*class* at **0.81** — the standard bar for this kind of task. The one weaker signal is
*mitigation-dependence*, discussed at the end of this section.

```{r}
#| label: fig-validation
#| fig-cap: "Agreement with the human answer key: the full 400-window sample (magenta) versus the held-out test (green). The dashed line marks 0.80, the standard bar for this kind of task."
include_fig("fig_validation_accuracy.png")
```

**Why the last two scores are lower — and why it doesn't undercut the findings.** Both are
*secondary attributes*, not the core call, and both are genuinely harder:

- *Mitigation-dependence (0.62).* Deciding whether a conclusion **depends on** committed mitigation
  is one of the hardest judgments in the whole task — our two human coders themselves agreed only
  ~68% of the time on the fine line between "less than significant" and "less than significant *with
  mitigation*." On top of that, the pipeline's structured mitigation flag is tagged at the document
  level, so it over-attributes across a multi-resource passage, and its precision is ultimately
  bounded by upstream tagging. **We handle this by reporting mitigation at the document level (where
  it is robust) and using the precise per-resource signal for the resource-level breakdown** — so
  the numbers you see below are the high-precision measure, not the noisy flag.
- *Regulatory threshold (0.66–0.71).* This is the most granular field — pinning a conclusion to a
  *specific* rule (a NAAQS limit, an ESA take threshold, a §106 adverse-effect finding). Many
  conclusions are not anchored to a single numeric threshold at all, and the distinctions are fine.
  It is reported descriptively as supporting context, never as a headline number.

The three scores that carry the findings — **finding the determination, placing the resource, and
getting the significance class** — all clear or approach the standard bar.

### How agencies stay below the line

Within FONSIs, the determinations sit — as they must — below the significance line: the agency
either finds **no significant impact** outright or an impact that is **less than significant**,
sometimes only **with committed mitigation**. The map below shows how that breaks down by resource
area.

```{r}
#| label: tbl-resource-map
class_order <- c("no_significant_impact", "less_than_significant",
                 "less_than_significant_with_mitigation")
res_by_class |>
  filter(!n_suppressed, determination_class %in% class_order) |>
  mutate(determination_class = factor(determination_class, levels = class_order)) |>
  select(shared_resource_area, determination_class, n) |>
  pivot_wider(names_from = determination_class, values_from = n, values_fill = 0) |>
  mutate(Total = no_significant_impact + less_than_significant +
                 less_than_significant_with_mitigation,
         Resource = relabel(shared_resource_area, res_label)) |>
  arrange(desc(less_than_significant_with_mitigation)) |>
  select(Resource, no_significant_impact, less_than_significant,
         less_than_significant_with_mitigation, Total) |>
  gt() |>
  cols_label(no_significant_impact = "No sig. impact", less_than_significant = "Less than sig.",
             less_than_significant_with_mitigation = "LTS w/ mitigation") |>
  fmt_number(columns = where(is.numeric), decimals = 0) |>
  cols_align(align = "left", columns = Resource) |>
  tab_style(style = list(cell_fill(color = NAVY), cell_text(color = "white", weight = "bold")),
            locations = cells_column_labels()) |>
  tab_options(table.font.size = px(13), table.width = pct(100)) |>
  tab_source_note("Primary BLM + DOE FONSI scope, deduplicated to one determination per document × resource × class. Cells below 5 suppressed. “Project-wide” is the document’s overall FONSI conclusion.")
```

The dominant outcome is **no significant impact**, as expected for FONSIs. The resources agencies
analyze most — **biological, water, cultural/historic, and public health** — are also where the
close calls concentrate, which is what makes the mitigated cases below the real story.

The composition within each resource — how often it lands at *no significant impact* versus *less
than significant* versus *less than significant only with mitigation*, and how many determinations
sit behind each bar — shows where agencies are comfortable and where they have to work to stay
under the line.

```{r}
#| label: fig-outcomes-by-resource
#| fig-cap: "Share of each resource's FONSI determinations by outcome. Resources with more magenta lean harder on committed mitigation to reach a not-significant finding."
include_fig("fig_outcomes_by_resource.png")
```

### Mitigated FONSIs — the line itself

The most revealing determinations are the **mitigated** ones: impacts the agency says *would* be
significant, brought below the line only by required measures. These show exactly where the line
sits and what it costs to stay under it.

```{r}
#| label: mit-headline
mit_rate <- mit_doc$share_mitigated_class_signal[1]
top_mit  <- mit_res |> filter(n_determinations >= 20) |> arrange(desc(share_mit_class)) |>
  slice_head(n = 4) |> mutate(Resource = relabel(shared_resource_area, res_label))
top_str  <- paste0(top_mit$Resource, " (", percent(top_mit$share_mit_class, accuracy = 1), ")",
                   collapse = ", ")
```

About **`r fmt_pct(mit_rate)` of these FONSIs are mitigated** — at least one resource whose
not-significant conclusion depends explicitly on committed mitigation. Which resources drive it:

```{r}
#| label: tbl-mit-resource
mit_res |>
  filter(shared_resource_area != "unknown", n_determinations >= 5) |>
  mutate(Resource = relabel(shared_resource_area, res_label)) |>
  arrange(desc(share_mit_class)) |>
  select(Resource, n_determinations, n_mit_class, share_mit_class) |>
  gt() |>
  cols_label(n_determinations = "Determinations", n_mit_class = "Mitigation-dependent",
             share_mit_class = "Share") |>
  fmt_number(columns = c(n_determinations, n_mit_class), decimals = 0) |>
  fmt_percent(columns = share_mit_class, decimals = 0) |>
  data_color(columns = share_mit_class, colors = col_numeric(c("white", BLUE), domain = c(0, NA))) |>
  cols_align(align = "left", columns = Resource) |>
  tab_style(style = list(cell_fill(color = NAVY), cell_text(color = "white", weight = "bold")),
            locations = cells_column_labels()) |>
  tab_options(table.font.size = px(13), table.width = pct(100)) |>
  tab_source_note("Share = fraction of a resource’s determinations that depend on committed mitigation (per-resource class signal, the high-precision measure).")
```

The mitigation burden is concentrated in **`r top_str`** — the resources most often needing
committed measures to keep a project below the significance line. That pattern is the deliverable's
core finding: biological, soils/geology, water, and cultural resources are where clean-energy
projects most frequently sit *at* the line and are pulled under it by mitigation.

```{r}
#| label: fig-mitigation-by-resource
#| fig-cap: "Share of a resource's FONSI conclusions that depend on committed mitigation (per-resource class signal). Point size scales with how often the resource is analyzed."
include_fig("fig_mitigation_by_resource.png")
```

Reading volume and mitigation-intensity together sharpens the picture. Resources in the **upper
right** are both analyzed often *and* mitigated often — the areas that most shape whether a
clean-energy project can take the faster FONSI path.

```{r}
#| label: fig-mitigation-landscape
#| fig-cap: "Each resource by how often it is analyzed (x) and how often its conclusion depends on mitigation (y). Biological resources dominate both; soils/geology are mitigated intensively despite lower volume. Dashed line = overall average mitigation share."
include_fig("fig_mitigation_landscape.png")
```

What do these mitigations actually look like? The two resources at the top of the landscape —
biological and water — lean on concrete, recognizable measures: seasonal timing restrictions,
species avoidance, and wetland minimization. A few verbatim examples pulled from the extraction:

```{r}
#| label: tbl-mitigation-examples
if (nrow(mit_ex) > 0) {
  mit_ex |>
    gt(groupname_col = "Resource") |>
    cols_label(example = "Example mitigation drawn from the FONSI") |>
    cols_align(align = "left") |>
    tab_style(style = list(cell_fill(color = NAVY), cell_text(color = "white", weight = "bold")),
              locations = cells_row_groups()) |>
    tab_options(table.font.size = px(12), table.width = pct(100),
                row_group.font.weight = "bold") |>
    tab_source_note("Verbatim from the model's per-resource rationale on mitigation-dependent (LTS-with-mitigation) determinations.")
}
```

Mitigated FONSIs are also **broader** analyses: a project that needs committed mitigation for any
resource tends to address more resource areas overall — a median of roughly six versus four for
non-mitigated FONSIs.

```{r}
#| label: fig-breadth
#| fig-cap: "Distinct resource areas addressed per FONSI, split by whether the document relies on committed mitigation. Boxes show the median and interquartile range; the shaded shape is the full distribution."
include_fig("fig_breadth_per_fonsi.png")
```

::: {.callout-warning appearance="minimal"}
**Read mitigation at the right grain.** The *document-level* "is this a mitigated FONSI?" rate is
robust. The *resource-level* attribution above uses the high-precision per-resource signal; a
broader structured signal exists but over-attributes across resources, and its accuracy is
ultimately bounded by upstream tagging (validation F1 ≈ 0.62). Treat the resource shares as a
well-grounded floor, not an exact census.
:::

## EIS analysis {#sec-eis}

::: {.callout-note}
**Forthcoming.** The EIS track is the *above-the-line* counterpart — which resources agencies judge
**significant**, and where impacts are **significant and unavoidable**. It uses the identical
extraction-and-validation machinery on the EIS corpus; the hand-labeled answer key is already
built. This section will be populated once the EIS extraction pass runs. See *Scope decisions*
below for the effort trade-off.
:::

## Scope decisions to confirm

A few items shape the deliverable's scope and breadth. Each has a recommended default below;
the goal is to confirm or adjust these before the full analysis runs.

1. **FONSIs only, or FONSIs *and* EISs?** — *the biggest decision.* Significance calls appear in
   both document types: **FONSIs** capture the *not-significant* side (including the mitigated
   cases sitting right at the line), while **EISs** capture the *significant* side. The FONSI
   analysis is built and ready. Adding the EIS side means extracting findings from a
   **~6.1-million-page** document set — a substantially larger AI and analyst effort that roughly
   **multiplies the extraction cost and timeline** and needs a retrieval-quality check first.
   **Recommendation:** deliver the validated FONSI analysis first; scope EISs as an add-on or a
   later phase once the FONSI work is in hand.

2. **Agencies: BLM + DOE, or broader?** These are the only two agencies with complete coverage in
   the dataset and account for the large majority of clean-energy FONSIs; other agencies (NRC,
   Forest Service, BOEM) appear mainly on the EIS side, with partial coverage.
   **Recommendation:** BLM + DOE as the primary comparison; other agencies reported as context
   only, never mixed into the headline rates.

3. **What counts as a "resource area"?** We read *"across resource areas"* as the standard NEPA
   environmental categories the documents are organized around — air quality; water (including
   wetlands and floodplains); biological resources; cultural / historic; visual; noise;
   soils / geology; socioeconomics (including environmental justice); transportation; land use;
   climate / greenhouse gases; and public health (including hazardous materials). Agencies judge
   significance separately for each, so this set defines the rows of the headline map.
   **Confirm:** is this the intended list, or does CATF have a specific set or a priority subset
   in mind?

## Methods & required caveats

- **Agency scope.** BLM + DOE agency family. Headline denominators use only
  `agency_scope_status = 'primary_blm_doe_family'`; other-agency and manual-review rows are a
  separate context universe.
- **D4 date/regime coverage.** Many projects lack a high-confidence decision date; regime cells
  are confidence-gated and thin post-2023. EIS decision dates are frequently proxies.
- **EPA review letters (Clean Air Act §309) are out of scope.** Under Section 309, the EPA
  independently reviews and *rates* other agencies' environmental impact statements (e.g.
  "environmental concerns" or "objections"). Those comment letters are not in this dataset, so
  the report makes no claim about EPA's own view of a project's significance.
- **Clean-energy scope only** (`project_energy_type = 'Clean'`).
- **FONSI coverage** for the mitigated/straight tiers derives from D6 artifacts.
- **EIS retrieval limits.** The EIS tier is gated behind FONSI validation and a retrieval-recall
  spike; EIS findings are preliminary.
- **Validation sample size.** Rare class/resource/threshold cells are reported descriptively,
  not as pass/fail macro-metrics.
- **Suppressed cells.** Any cell below 5 is suppressed.

## Appendix: data dictionary

See `phase2/notes/deliverable02/data_dictionary.md` and the run manifest
`significance_run_manifest.parquet` (input+output paths, row counts, content hashes, model,
prompt/schema versions).
