---
title: "NEPA Decarbonization Technology Analysis: Deliverable 5"
subtitle: "Categorical Exclusion Spikes After Major Infrastructure Legislation"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    toc-location: left
    fig-align: center
execute:
  echo: false
  warning: false
  message: false
---

<!-- Render phase2/reports/deliverable05.qmd for automatic table and inline-value updates.
     Figures are produced by phase2/code/deliverable05/03_create_figures.R; the law-citation and
     CE-category inputs by 01_extract_law_citations.py and 02_build_ce_categories.py. -->

```{r setup}
library(arrow)
library(dplyr)
library(tidyr)
library(readr)
library(stringr)
library(lubridate)
library(scales)
library(knitr)
library(gt)
library(here)

options(arrow.skip_nul = TRUE)

TIMELINE <- here("phase2", "data", "analysis", "timeline", "timeline_project_dates.parquet")
PROJECTS <- here("phase1", "data", "analysis", "projects_combined.parquet")
CITES    <- here("phase2", "data", "analysis", "deliverable05", "law_citations.parquet")
FIG_DIR  <- here("phase2", "output", "deliverable05", "figures")
DIAG_DIR <- here("phase2", "output", "deliverable05", "diagnostics")

# --- helpers ---------------------------------------------------------------
fmt_n   <- function(x) comma(round(x), accuracy = 1)
pct1    <- function(x) paste0(number(x, accuracy = 0.1), "%")        # x already on 0-100 scale
ratio1  <- function(x) number(x, accuracy = 0.01)
first0  <- function(x) if (length(x) == 0 || all(is.na(x))) 0 else x[[1]]
include_fig <- function(name) knitr::include_graphics(file.path(FIG_DIR, name))

# --- recompute CE-by-year by agency (live, mirrors 03_create_figures.R) -----
dates <- read_parquet(TIMELINE)
meta  <- read_parquet(PROJECTS, col_select = c("project_id", "project_energy_type",
                                               "project_department", "lead_agency_harmonized"))
ce <- dates |>
  filter(process_type == "CE") |>
  left_join(meta, by = "project_id") |>
  mutate(
    decision_date   = as.Date(decision_date),
    initiation_date = as.Date(initiation_date),
    year_date = coalesce(decision_date, initiation_date),
    year = year(year_date),
    agency2 = case_when(
      str_detect(coalesce(lead_agency_harmonized, ""), "Department of Energy") ~ "DOE",
      str_detect(coalesce(lead_agency_harmonized, ""), "Bureau of Land Management") ~ "BLM",
      TRUE ~ "Other")
  )

n_ce_total     <- nrow(ce)
n_ce_placeable <- sum(!is.na(ce$year))
pct_ce_placeable <- 100 * n_ce_placeable / n_ce_total

doe_year <- ce |> filter(agency2 == "DOE", !is.na(year)) |> count(year, name = "n")
blm_year <- ce |> filter(agency2 == "BLM", !is.na(year)) |> count(year, name = "n")
yr <- function(df, y) first0(df$n[df$year == y])
doe_peak_year <- doe_year$year[which.max(doe_year$n)]
doe_peak_n    <- max(doe_year$n)

# --- diagnostic tables (written by 03) -------------------------------------
spike      <- read_csv(file.path(DIAG_DIR, "d5_spike_summary.csv"), show_col_types = FALSE)
citerate   <- read_csv(file.path(DIAG_DIR, "d5_citation_rates.csv"), show_col_types = FALSE)
catshift   <- read_csv(file.path(DIAG_DIR, "d5_category_shift.csv"), show_col_types = FALSE)
techshift  <- read_csv(file.path(DIAG_DIR, "d5_technology_shift.csv"), show_col_types = FALSE)

cite_rate <- function(law_, scope_, period_) first0(
  citerate$pct_cited[citerate$law == law_ & citerate$scope == scope_ & citerate$period == period_])
cat_pct <- function(law_, scope_, period_, code_) first0(
  catshift$pct[catshift$law == law_ & catshift$scope == scope_ &
               catshift$period == period_ & catshift$code_norm == code_])
spike_ratio_val <- function(subset_, law_) first0(
  spike$spike_ratio[spike$subset == subset_ & spike$law == law_])

# headline scalars
arra_cite_ce  <- cite_rate("ARRA", "All CE", "spike window")
b51_spike     <- cat_pct("ARRA", "All CE", "spike window", "B5.1")
b51_base      <- cat_pct("ARRA", "All CE", "baseline", "B5.1")
bil_doe_ratio <- spike_ratio_val("DOE", "BIL")
ira_eis_win   <- cite_rate("IRA", "All EIS", "spike window")
ira_eis_base  <- cite_rate("IRA", "All EIS", "baseline")
bil_eis_win   <- cite_rate("BIL", "All EIS", "spike window")
```

# Executive Summary

This deliverable examines whether the use of **categorical exclusions (CEs)** — the lightest form of
NEPA review — spikes after the passage of major infrastructure legislation, and if so, whether those
actions are demonstrably tied to the new law. We anchor every CE to the date its determination was
issued (from the Deliverable 4 timeline), detect explicit citations to the legislation in the CE
documents, and read the specific categorical-exclusion *category* each action invoked.

The three named laws are the **American Recovery and Reinvestment Act (ARRA**, Feb 2009), the
**Bipartisan Infrastructure Law / Infrastructure Investment and Jobs Act (BIL/IIJA**, Nov 2021), and
the **Inflation Reduction Act (IRA**, Aug 2022).

::: callout-note
## Key Findings

1. **There is a large, unmistakable CE spike after ARRA — and it is almost entirely a Department of
   Energy phenomenon.** DOE categorical exclusions jumped from `r fmt_n(yr(doe_year, 2008))` in 2008
   to `r fmt_n(yr(doe_year, 2009))` in 2009 and `r fmt_n(yr(doe_year, 2010))` in 2010, while the
   other major CE user, the Bureau of Land Management, stayed flat
   (`r fmt_n(yr(blm_year, 2010))` in 2010). The agency that administered ARRA's energy money is
   exactly the agency whose CEs surged.

2. **The spiking actions are demonstrably tied to the law.** `r pct1(arra_cite_ce)` of CEs issued in
   the ARRA window explicitly cite the Recovery Act in their documents — direct attribution, not a
   temporal coincidence.

3. **The *type* of CE used shifted decisively toward the law's purpose.** Among DOE categorical
   exclusions in the ARRA window, category B5.1, "Actions to conserve energy or water," rose to
   `r pct1(b51_spike)` of all DOE CEs (from `r pct1(b51_base)` at baseline) — the energy-efficiency
   stimulus showing up directly in the categorical-exclusion mix.

4. **BIL and IRA show a real but far more muted pattern.** DOE CE activity rose after BIL
   (`r ratio1(bil_doe_ratio)`× the pre-law monthly baseline), but explicit CE citations to BIL and IRA
   are rare — those laws are referenced far more often in larger reviews (EISs) than in CEs
   (IRA is cited in `r pct1(ira_eis_win)` of post-IRA EISs vs a negligible share of CEs).

5. **Read raw counts with care.** The dataset's coverage thins before 2009 and is incomplete for
   2024–2025, so aggregate counts overstate the ARRA step-change. The DOE-vs-BLM split and the
   citation evidence isolate the genuine policy signal from this coverage ramp.
:::

# Methodology

## The question, in three parts

The deliverable asks three distinct questions, each answered by a distinct analysis:

| Question | Analysis |
|---|---|
| **Is there a spike** in CE use after each law? | CE counts by determination year, with legislative markers — overall, by review type, by department, and DOE-vs-BLM. |
| **Are the actions associated** with the law? | Detection of explicit law citations in the CE document text. A citation can only post-date the law, so it is coverage-robust attribution evidence. |
| **What *types* of CEs** were used? | The specific categorical-exclusion category invoked (e.g. DOE's B5.1), plus the technology mix of the projects, in the spike window vs a baseline. |

The pipeline runs in four stages:

```{=html}
<div style="max-width:660px;margin:1.25rem auto;">
  <div style="border:2px solid #0047BB;border-radius:8px;padding:.65rem 1rem;background:#f4f7fc;">
    <div style="font-weight:700;color:#012169;font-size:1.05rem;">1 · Anchor CEs in time</div>
    <div style="color:#1A1A2E;">Place each CE by its <b>determination (decision) date</b> from the Deliverable&nbsp;4 timeline, falling back to the <b>initiation date</b> as a same-year proxy where the decision is absent (safe for CEs, whose median duration is ≈3 weeks). This locates <b>52,089 of 54,668</b> CEs (95.3%) in time — far broader than the complete-timeline base used for duration in Deliverable&nbsp;4</div>
  </div>
  <div style="text-align:center;color:#0047BB;font-size:1.5rem;line-height:1.1;">&#8595;</div>
  <div style="border:2px solid #0047BB;border-radius:8px;padding:.65rem 1rem;background:#f4f7fc;">
    <div style="font-weight:700;color:#012169;font-size:1.05rem;">2 · Detect law citations</div>
    <div style="color:#1A1A2E;"><code>01_extract_law_citations.py</code> scans CE / EA / EIS document pages for explicit <b>ARRA</b>, <b>BIL</b>, and <b>IRA</b> citations, with context-based disambiguation of the ambiguous IRA and BIL acronyms. A citation can only post-date the law, so it is coverage-robust attribution evidence</div>
  </div>
  <div style="text-align:center;color:#0047BB;font-size:1.5rem;line-height:1.1;">&#8595;</div>
  <div style="border:2px solid #0047BB;border-radius:8px;padding:.65rem 1rem;background:#f4f7fc;">
    <div style="font-weight:700;color:#012169;font-size:1.05rem;">3 · Read CE categories</div>
    <div style="color:#1A1A2E;"><code>02_build_ce_categories.py</code> parses the document-level <code>ce_category</code> metadata into normalized CFR codes (DOE 10 CFR 1021, DOI 516 DM 11, EPAct 2005 §390), so the categorical-exclusion mix can be compared across periods</div>
  </div>
  <div style="text-align:center;color:#0047BB;font-size:1.5rem;line-height:1.1;">&#8595;</div>
  <div style="border:2px solid #0047BB;border-radius:8px;padding:.65rem 1rem;background:#f4f7fc;">
    <div style="font-weight:700;color:#012169;font-size:1.05rem;">4 · Compare spike windows vs baselines</div>
    <div style="color:#1A1A2E;"><code>03_create_figures.R</code> joins dates, citations, and categories, then produces the counts, citation rates, and category- and technology-mix comparisons — spike window versus baseline — read by this report</div>
  </div>
  <div style="text-align:center;color:#0047BB;font-size:1.5rem;line-height:1.1;">&#8595;</div>
  <div style="border:2px solid #012169;border-radius:8px;padding:.65rem 1rem;background:#012169;">
    <div style="font-weight:700;color:#ffffff;font-size:1.05rem;">Deliverable 5 report</div>
    <div style="color:#ffffff;">Whether CE use spikes after major legislation, whether the spike is tied to the law, and which categorical exclusions were used</div>
  </div>
</div>
```

The base population is every CE placeable in time. Anchoring on the determination date with an
initiation-date fallback yields a base of `r fmt_n(n_ce_placeable)` of `r fmt_n(n_ce_total)` CE
projects (`r pct1(pct_ce_placeable)`) — far broader than the complete-timeline base used for the
duration analysis in Deliverable 4, because placing a CE in a year requires only one date, not two. How
the analysis separates a genuine policy signal from the growth of the dataset's coverage over time is
detailed on the [Coverage & Limitations page](../notes/deliverable05/coverage_limitations.html).

# The Spike

Categorical exclusions rise sharply around ARRA and again, more modestly, in the BIL/IRA period.

```{r fig-ce-all}
#| label: fig-ce-all
#| fig-cap: "Categorical exclusions by determination year. Dashed lines mark ARRA (2009), BIL (2021), and IRA (2022)."
include_fig("fig_d5_ce_counts_by_year_all.png")
```

The spike is **specific to categorical exclusions**. Faceting the same series by review type shows that
the more demanding Environmental Assessment and Environmental Impact Statement processes do not exhibit
the post-ARRA jump — consistent with a surge of fast, low-burden actions rather than a general increase
in NEPA activity.

```{r fig-byprocess}
#| label: fig-byprocess
#| fig-cap: "Reviews by year and review type (CE / EA / EIS). The post-ARRA spike appears only in the CE panel."
include_fig("fig_d5_counts_by_year_byprocess.png")
```

## The spike is a DOE phenomenon

The single most important result is that the post-ARRA CE surge belongs almost entirely to the
**Department of Energy** — the agency that administered ARRA's energy grants, loan guarantees, and
weatherization funds. DOE CEs went from `r fmt_n(yr(doe_year, 2008))` (2008) to
`r fmt_n(yr(doe_year, 2009))` (2009) to a peak of `r fmt_n(doe_peak_n)` in `r doe_peak_year`. Over
the same years the Bureau of Land Management — the only other agency that issues CEs at scale, and one
drawing on the same dataset — barely moved (`r fmt_n(yr(blm_year, 2009))`, `r fmt_n(yr(blm_year, 2010))`,
`r fmt_n(yr(blm_year, 2011))` in 2009–2011). BLM's own growth comes later (2016 onward) and is unrelated
to the legislation. Because both agencies are subject to the identical coverage ramp, this contrast is
the cleanest available evidence that the spike is a policy effect, not a data artifact.

```{r fig-doe-blm}
#| label: fig-doe-blm
#| fig-cap: "Categorical exclusions by year, DOE vs BLM. The ARRA spike is a DOE phenomenon; BLM is flat through the ARRA window."
include_fig("fig_d5_ce_counts_by_year_doe_blm.png")
```

```{r spike-table}
spike_tbl <- spike |>
  mutate(
    Subset = subset, Law = law,
    `Window (CEs/mo)` = number(window_mean_monthly, accuracy = 0.1),
    `Baseline (CEs/mo)` = ifelse(is.na(baseline_mean_monthly), "—",
                                 number(baseline_mean_monthly, accuracy = 0.1)),
    `Spike ratio` = ifelse(is.na(spike_ratio), "—", number(spike_ratio, accuracy = 0.01))
  ) |>
  select(Subset, Law, `Window (CEs/mo)`, `Baseline (CEs/mo)`, `Spike ratio`)
spike_tbl |>
  gt() |>
  tab_header(title = "Mean monthly CE volume: spike window vs pre-law baseline") |>
  tab_source_note("ARRA has no usable pre-law baseline (the dataset is sparse before 2009); its spike is established by the DOE-vs-BLM contrast and citation evidence rather than a window/baseline ratio.") |>
  tab_options(table.font.size = px(13))
```

## By energy type and department

Splitting the CE series by energy type (decarbonization, fossil, other) and by lead department provides
the all-data and breakdown views requested for every analysis.

```{r fig-byenergy}
#| label: fig-byenergy
#| fig-cap: "CE counts by year, by energy type."
include_fig("fig_d5_ce_counts_by_year_byenergy.png")
```

```{r fig-bydept}
#| label: fig-bydept
#| fig-cap: "CE counts by year, by lead department."
include_fig("fig_d5_ce_counts_by_year_bydept.png")
```

```{r fig-byprocess-byenergy}
#| label: fig-byprocess-byenergy
#| fig-cap: "Reviews by year, faceted by review type and stacked by energy type."
include_fig("fig_d5_counts_by_year_byprocess_byenergy.png")
```

# Association With the Law

A temporal coincidence is not attribution. The strongest test is whether the spiking documents *say* the
action stems from the law. We detect explicit citations to each statute in the document text (with
context disambiguation for the ambiguous "IRA" and "BIL" acronyms).

```{r fig-cites-year}
#| label: fig-cites-year
#| fig-cap: "Reviews whose documents explicitly cite each law, by year. A citation cannot precede the law's passage."
include_fig("fig_d5_citations_by_year_byprocess.png")
```

For **ARRA the attribution is overwhelming**: `r pct1(arra_cite_ce)` of CEs issued in the ARRA window
cite the Recovery Act by name, versus a negligible rate outside it. The Recovery Act citations themselves
peak in 2010 and decay over the following years as stimulus obligations wound down — the citation curve
tracks the disbursement curve.

**BIL and IRA tell a more nuanced story.** Their citation rates rise after passage relative to baseline,
confirming a real association, but the absolute rate within CEs stays low — these laws are seldom named
in a one-page categorical exclusion. They surface far more often in larger reviews: IRA is cited in
`r pct1(ira_eis_win)` of post-IRA EISs (vs `r pct1(ira_eis_base)` before), and BIL in
`r pct1(bil_eis_win)` of post-BIL EISs. In other words, the *citation* evidence for BIL/IRA lives in the
EIS record, even though the *volume* signal (more DOE CEs) is real.

```{r fig-citerate}
#| label: fig-citerate
#| fig-cap: "Share of CEs citing each law, spike window vs baseline."
include_fig("fig_d5_citation_rate_window_vs_baseline.png")
```

```{r citerate-table}
citerate |>
  mutate(`% citing` = pct1(pct_cited), N = fmt_n(n)) |>
  transmute(Law = law, Scope = scope, Period = period, `% citing`, N) |>
  filter(Scope %in% c("All CE", "All EIS")) |>
  gt() |>
  tab_header(title = "Law-citation rate by scope and period") |>
  tab_options(table.font.size = px(12))
```

# What Types of CEs Were Used

NEPATEC records the specific categorical-exclusion category each action invoked. For DOE these are the
codes of 10 C.F.R. 1021 (e.g. `B5.1`, `A9`, `B1.3`); for the Interior/BLM they are the `516 DM 11`
series; oil-and-gas drilling uses the Energy Policy Act of 2005 Section 390 exclusion. Comparing the mix
inside the ARRA window against a stable baseline shows the categorical-exclusion profile shifting
directly toward the stimulus's purpose.

```{r fig-catshift}
#| label: fig-catshift
#| fig-cap: "DOE categorical-exclusion category mix, ARRA window (2009-2011) vs 2016-2019 baseline."
include_fig("fig_d5_ce_category_shift_arra.png")
```

The standout is **B5.1, "Actions to conserve energy or water,"** which accounts for `r pct1(b51_spike)`
of DOE categorical exclusions in the ARRA window versus only `r pct1(b51_base)` at baseline — a roughly
forty-fold concentration. Information-gathering and technical-assistance categories (A9, A11, A1) are
also elevated, consistent with a wave of grant- and study-related actions. By contrast, the baseline
period is dominated by routine-maintenance exclusions (B1.3), which recede during the stimulus. The
categorical-exclusion mix, in short, reorganizes around energy efficiency exactly when ARRA's efficiency
money arrives.

```{r catshift-table}
catshift |>
  filter(law == "ARRA", scope == "All CE", period == "spike window") |>
  slice_max(pct, n = 8) |>
  transmute(
    Code = code_norm,
    Description = ifelse(is.na(code_description) | code_description == code_norm, "—", code_description),
    `ARRA window %` = pct1(pct),
    `Baseline %` = pct1(sapply(code_norm, function(c) cat_pct("ARRA", "All CE", "baseline", c)))
  ) |>
  gt() |>
  tab_header(title = "Top DOE categorical-exclusion codes in the ARRA window") |>
  tab_options(table.font.size = px(13))
```

## Technology mix

By project technology, the ARRA-window CEs are dominated by DOE environmental-management and
research-and-development categories alongside the expected clean-energy build-out (solar, transmission).

```{r fig-techshift}
#| label: fig-techshift
#| fig-cap: "Technology-tag mix of ARRA-window CEs (a project may carry multiple tags)."
include_fig("fig_d5_technology_shift_arra.png")
```

```{r techshift-table}
techshift |>
  filter(law == "ARRA") |>
  slice_max(pct, n = 8) |>
  transmute(`Technology tag` = pt, `Share of ARRA-window CEs` = pct1(pct), N = fmt_n(n)) |>
  kable(format = "pipe", align = "l")
```

# Reproducibility

All figures and tables in this report regenerate from committed code. Run from the repository root in the `nepa` conda environment:

``` bash
python phase2/code/deliverable05/02_build_ce_categories.py
python phase2/code/deliverable05/01_extract_law_citations.py --source all
Rscript phase2/code/deliverable05/03_create_figures.R
```

Outputs are written to `phase2/output/deliverable05/{figures,diagnostics}/`; the intermediate
citation and category datasets to `phase2/data/analysis/deliverable05/`.
