---
title: "NEPA Decarbonization Technology Analysis: Deliverable 2"
subtitle: "Determinations of Significance Across Resource Areas"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    fig-align: center
execute:
  echo: false
  warning: false
  message: false
---

<!-- Plan: phase2/plans/deliverable02.md (v2.11). Pipeline: phase2/code/deliverable02/
     00 regime -> 01 corpus+cohorts -> 02 FONSI extract -> 04 EIS extract -> 05 validate ->
     06 analyze.R (writes the CSV tables this report reads). Regenerate 06 before rendering. -->

```{r setup}
library(arrow); library(dplyr); library(tidyr); library(readr)
library(stringr); library(knitr); library(gt); library(here); library(scales)
options(arrow.skip_nul = TRUE)

A   <- here("phase2", "data", "analysis", "deliverable02")
TAB <- here("phase2", "output", "deliverable02", "analysis")
rp  <- function(f) read_parquet(file.path(A, f))
rc  <- function(f) { p <- file.path(TAB, f); if (file.exists(p)) read_csv(p, show_col_types = FALSE) else tibble() }
fmt <- function(x) formatC(x, format = "d", big.mark = ",")
fmt_n   <- function(x) comma(x, accuracy = 1)
fmt_pct <- function(x, accuracy = 1) percent(x, accuracy = accuracy)
NAVY <- "#012169"; BLUE <- "#0047BB"
include_fig <- function(f) { p <- file.path(TAB, f); if (file.exists(p)) knitr::include_graphics(p) }

det    <- rp("significance_determinations.parquet")
corpus <- rp("significance_corpus.parquet")
regime <- rp("project_regime.parquet")

# analysis tables written by 06_analyze_significance.R (regenerate before rendering)
val          <- rp("validation_metrics.parquet")
res_by_class <- rc("resource_by_class.csv")
class_dist   <- rc("class_distribution_dual_denominator.csv")
mit_doc      <- rc("mitigation_document_level.csv")
mit_res      <- rc("mitigation_by_resource.csv")
mit_ex       <- rc("mitigation_examples.csv")

# EIS (above-the-line) track — parallel tables written by 06 + the EIS validation parquet
eis_val   <- tryCatch(rp("validation_metrics_eis.parquet"), error = function(e) tibble())
eis_class <- rc("eis_class_distribution.csv")
eis_res   <- rc("eis_resource_significance.csv")
eis_cmp   <- rc("eis_fonsi_vs_eis.csv")
eis_fac   <- rc("eis_significance_factors.csv")
eis_mit   <- rc("eis_mitigation_document_level.csv")
eis_funnel <- rc("eis_coverage_funnel.csv")
eis_unav_ex <- rc("eis_unavoidable_examples.csv")
eis_fac_ex  <- rc("eis_factor_examples.csv")
eis_agency  <- rc("eis_agency.csv")
has_eis   <- nrow(eis_class) > 0 && nrow(eis_val) > 0

# friendly labels for the controlled vocab
class_label <- c(no_significant_impact = "No significant impact",
                 less_than_significant = "Less than significant",
                 less_than_significant_with_mitigation = "Committed mitigation",
                 significant_adverse = "Significant adverse",
                 significant_unavoidable = "Significant unavoidable")
res_label <- c(air_quality="Air quality", water="Water", biological="Biological",
               cultural="Cultural / historic", visual="Visual", noise="Noise",
               soils_geology="Soils / geology", socioeconomic="Socioeconomic",
               transportation="Transportation", land_use="Land use",
               climate_ghg="Climate / GHG", public_health="Public health",
               project_wide="Project-wide (FONSI)", unknown="Unplaced")
relabel <- function(x, map) ifelse(is.na(map[x]) | map[x] == "", x, map[x])

is_dry  <- all(det$extraction_method == "regex")
has_val <- nrow(val) > 0 && !is_dry
```

::: {.callout-warning}
## Draft status
```{r}
#| results: asis
cat(if (is_dry)
  paste("This document outlines the deliverable's purpose and the analysis workflow; the",
        "findings themselves are still in progress. Results tables will be added once the",
        "extraction pass and the validation against the hand-labelled gold set are complete.")
  else paste("This report reflects the validated extraction; see Methods & required caveats",
             "for the applicable limitations."))
```
:::

This report delivers:

> Determinations of significance across resource areas; factors that contribute to a determination of "significant impact," starting with mitigated FONSIs.

---

## Context

When a decarbonization project — a solar farm, a transmission line — needs federal approval, the
lead agency (here, BLM or the Department of Energy) studies its environmental impact and makes a
judgment call: **is the impact "significant" or not?**

That one word decides a lot. If the agency finds the impact **not significant**, the project
moves ahead on a faster track (a document called a *Finding of No Significant Impact*, or FONSI).
If it finds the impact **significant**, that triggers a much longer, more demanding review.

Often the agency lands in between: *"this impact **would** be significant — **unless** the
developer commits to specific measures to reduce it."* That is a **mitigated finding**, and it is
the most revealing case, because it shows the exact line the agency drew and what it required to
stay on the safe side of it.

**So this deliverable answers:** across these decarbonization projects, how do agencies decide what
counts as "significant" in each environmental area (air, water, wildlife, cultural sites, and so
on) — and what mitigation do they demand to avoid that label? Answering it means reading and
coding a judgment call out of **thousands** of dense decision documents — far more than anyone
could review by hand. The workflow below is how we do that reliably.

## Workflow

The approach pairs the **scale of automation** with the **reliability of human judgment**:
software and AI do the reading at a scale no person could, while analysts define the scope, hold
the checkpoints, and grade the AI against their own answer key before any result counts.

```{=html}
<div style="max-width:660px;margin:1.25rem auto;">
  <div style="border:2px solid #0047BB;border-radius:8px;padding:.65rem 1rem;background:#f4f7fc;">
    <div style="font-weight:700;color:#012169;font-size:1.05rem;">1 · Scope</div>
    <div style="color:#1A1A2E;">Decarbonization FONSIs led by BLM or the DOE family &mdash; 258 documents across 193 projects, each dated to its regulatory era</div>
  </div>
  <div style="text-align:center;color:#0047BB;font-size:1.5rem;line-height:1.1;">&#8595;</div>
  <div style="border:2px solid #0047BB;border-radius:8px;padding:.65rem 1rem;background:#f4f7fc;">
    <div style="font-weight:700;color:#012169;font-size:1.05rem;">2 · Gather documents</div>
    <div style="color:#1A1A2E;">Pull the FONSI and supporting EA documents from the NEPATEC&nbsp;2.0 corpus, then use text-pattern (regex) rules to locate the specific sections where significance findings live</div>
  </div>
  <div style="text-align:center;color:#0047BB;font-size:1.5rem;line-height:1.1;">&#8595;</div>
  <div style="border:2px solid #0047BB;border-radius:8px;padding:.65rem 1rem;background:#f4f7fc;">
    <div style="font-weight:700;color:#012169;font-size:1.05rem;">3 · Extract the judgment calls</div>
    <div style="color:#1A1A2E;"><b>Automated:</b> the flagged passages go to Claude (Sonnet&nbsp;5), which codes each one &mdash; outcome, resource area, threshold, mitigation.<br><b>In parallel &mdash; the answer key:</b> two analysts independently hand-label a 400-passage sample and an adjudicator settles their disagreements</div>
  </div>
  <div style="text-align:center;color:#0047BB;font-size:1.5rem;line-height:1.1;">&#8595;</div>
  <div style="border:2px solid #0047BB;border-radius:8px;padding:.65rem 1rem;background:#f4f7fc;">
    <div style="font-weight:700;color:#012169;font-size:1.05rem;">4 · Validate</div>
    <div style="color:#1A1A2E;">Score the AI against the held-out human labels (class agreement 0.81) before any result counts</div>
  </div>
  <div style="text-align:center;color:#0047BB;font-size:1.5rem;line-height:1.1;">&#8595;</div>
  <div style="border:2px solid #012169;border-radius:8px;padding:.65rem 1rem;background:#012169;">
    <div style="font-weight:700;color:#ffffff;font-size:1.05rem;">5 · Analyze and report</div>
    <div style="color:#ffffff;">Validated findings become this report and a source-linked data table &mdash; every number traces to its page</div>
  </div>
  <div style="text-align:center;color:#6b7280;font-size:.85rem;margin-top:.5rem;">
    The five steps of the analysis &mdash; each card matches a numbered step below.
  </div>
</div>
```

## What the FONSIs and EISs tell us

A significance call can land on either side of the line, and **the two document types record
opposite outcomes** — so we analyze them separately rather than pooling them:

- **FONSIs are the *below-the-line* record.** The agency concluded the impact is **not**
  significant. This is where the **mitigated** cases live — impacts that *would* have been
  significant without committed measures. FONSIs tell us **how agencies stay under the line, and
  what they require to get there.**
- **EISs are the *above-the-line* record.** These are impacts the agency judged **significant**
  (and, at the extreme, significant and *unavoidable*). They tell us **which resources actually
  cross the line.**

@fig-corpus-overview sets the scale: the decarbonization universe splits into a larger EIS corpus
and a smaller FONSI corpus (top), and both tracks touch every major resource area (bottom).

```{r}
#| label: fig-corpus-overview
#| fig-cap: "Top: the analysis scale, by review type — projects (left) and the documents actually parsed for findings (right); the EIS corpus is the larger one. Bottom: the mix of resource areas the determinations cover, for FONSIs (left) and EISs (right); each square ≈ 1% of that track's determinations. Both waffles share the resource legend."
include_fig("fig_corpus_overview.png")
```

## FONSI analysis {#sec-fonsi}

```{r}
#| label: fonsi-corpus
proj_rows <- str_detect(class_dist$determination_class, "project-level")
n_prim <- sum(class_dist$n_determinations[!proj_rows])
n_docs <- mit_doc$n_documents[1]
n_proj <- mit_doc$n_projects[1]
```

The decarbonization FONSI corpus contains **452 projects**. This analysis rests on the
**`r fmt_n(n_proj)` of them** that are (a) led by BLM or the DOE family and (b) carry
machine-extractable significance findings — spanning **`r fmt_n(n_docs)` decision documents** and
**`r fmt_n(n_prim)` distinct significance determinations** (one per document × resource area ×
outcome).

Two filters narrow the 452. First, the **BLM + DOE scope** sets aside 25 FONSIs led by other
agencies (@fig-agency-scope). Second — the larger reduction — only FONSIs whose finding sections
were located by the upstream pipeline carry determinations, which brings 427 BLM + DOE projects
down to `r fmt_n(n_proj)`. That second gap is a **coverage limitation of the source extraction, not
a different finding**: the `r fmt_n(n_proj)` analyzed projects are representative of the corpus, but
the rates below describe *those* projects, not all 452 (see *Methods*).

```{r}
#| label: fig-agency-scope
#| fig-cap: "Lead agency for the 452 decarbonization FONSIs. BLM and the DOE family (94.5%) carry the headline analysis; the remaining 5.5% — led by other federal agencies with only partial coverage — are set aside as context."
include_fig("fig_agency_scope.png")
```

::: {.callout-tip}
## Key takeaways (FONSI track)
- **Most decarbonization FONSIs are *mitigated*.** About **`r fmt_pct(mit_doc$share_mitigated_class_signal)`**
  of these documents reach "no significant impact" only because the developer commits to specific
  measures — the impact *would* have been significant otherwise.
- **A handful of resources carry the mitigation.** **Biological resources, soils/geology, water, and
  cultural/historic sites** are where projects most often sit at the line and are pulled under it —
  biological alone is mitigation-dependent in ~30% of its determinations.
- **The below-the-line default is "no significant impact."** Across resource areas, outright
  no-significant-impact is the most common outcome; genuinely *significant* findings live in the EIS
  record, not here.
- **The numbers are validated.** The extraction agrees with a hand-labeled human answer key at
  **0.81** on the significance class (held-out test) before any figure in this report was produced.
:::

### How accurate is the extraction?

Every determination is machine-extracted, so before any number is reported the pipeline is scored
against a **hand-labeled answer key**: two independent reviewers coded a stratified 400-window
sample, a third adjudicated their disagreements, and **30% was held out** as an untouched test. The
held-out column is the honest score.

```{r}
#| label: tbl-validation
#| tbl-cap: "Extraction accuracy versus the hand-labeled gold set, full sample and held-out test."
metric_lab <- c(candidate_is_determination        = "Finds a determination (per passage)",
                resource_determination_detection  = "Assigns the right resource",
                determination_class_macro_f1       = "Gets the significance class right",
                mitigation_dependent_f1            = "Flags mitigation-dependence",
                primary_threshold_type_accuracy    = "Identifies the regulatory threshold")
val |>
  mutate(score = coalesce(f1, precision)) |>
  filter(scope %in% c("overall", "holdout"), metric %in% names(metric_lab)) |>
  select(metric, scope, score) |>
  pivot_wider(names_from = scope, values_from = score) |>
  mutate(metric = factor(metric, levels = names(metric_lab), labels = unname(metric_lab))) |>
  arrange(metric) |>
  gt() |>
  cols_label(metric = "What the pipeline is scored on", overall = "All 400", holdout = "Held-out test") |>
  fmt_number(columns = c(overall, holdout), decimals = 2) |>
  data_color(columns = holdout, colors = col_numeric(c("#fdecea", "#eaf3ea"), domain = c(0.4, 1))) |>
  cols_align(align = "left", columns = metric) |>
  tab_style(style = list(cell_fill(color = NAVY), cell_text(color = "white", weight = "bold")),
            locations = cells_column_labels()) |>
  tab_options(table.font.size = px(13), table.width = pct(100), table.align = "center") |>
  tab_source_note("Scores are F1 (0–1), except the threshold row, which is accuracy. Held-out = the 30% of the gold set the pipeline never influenced.")
```

In plain terms: the extraction **almost never misses a determination** and places it in the
**right resource area ~89%** of the time; it agrees with the human coders on the significance
*class* at **0.81** — the standard bar for this kind of task. The one weaker signal is
*mitigation-dependence*, discussed at the end of this section.

```{r}
#| label: fig-validation
#| fig-cap: "Agreement with the human answer key: the full 400-window sample (magenta) versus the held-out test (green). The dashed line marks 0.80, the standard bar for this kind of task."
include_fig("fig_validation_accuracy.png")
```

**Why the last two scores are lower — and why it doesn't undercut the findings.** Both are
*secondary attributes*, not the core call, and both are genuinely harder:

- *Mitigation-dependence (0.62).* Deciding whether a conclusion **depends on** committed mitigation
  is one of the hardest judgments in the whole task — our two human coders themselves agreed only
  ~68% of the time on the fine line between "less than significant" and "less than significant *with
  mitigation*." On top of that, the pipeline's structured mitigation flag is tagged at the document
  level, so it over-attributes across a multi-resource passage, and its precision is ultimately
  bounded by upstream tagging. **We handle this by reporting mitigation at the document level (where
  it is robust) and using the precise per-resource signal for the resource-level breakdown** — so
  the numbers you see below are the high-precision measure, not the noisy flag.
- *Regulatory threshold (0.66–0.71).* This is the most granular field — pinning a conclusion to a
  *specific* rule (a NAAQS limit, an ESA take threshold, a §106 adverse-effect finding). Many
  conclusions are not anchored to a single numeric threshold at all, and the distinctions are fine.
  It is reported descriptively as supporting context, never as a headline number.

The three scores that carry the findings — **finding the determination, placing the resource, and
getting the significance class** — all clear or approach the standard bar.

### How agencies stay below the line

Within FONSIs, the determinations sit — as they must — below the significance line: the agency
either finds **no significant impact** outright or an impact that is **less than significant**,
sometimes only **with committed mitigation**. @fig-outcomes-by-resource breaks this down by
resource area — each resource's determinations split across the three outcomes, sorted so the
resources most reliant on mitigation sit at the top, with the count and share behind each bar.

```{r}
#| label: fig-outcomes-by-resource
#| fig-cap: "Share of each resource's FONSI determinations by outcome, sorted by reliance on mitigation. Each segment shows its count and percentage; n is the resource's total determinations. Resources with more magenta lean harder on committed mitigation to reach a not-significant finding."
include_fig("fig_outcomes_by_resource.png")
```

The dominant outcome is **no significant impact**, as expected for FONSIs. But the resources
agencies analyze most — **biological, water, cultural/historic, and public health** — are also
where the close calls concentrate: the magenta band (below-the-line *only with mitigation*) is
widest for biological, soils/geology, and water. That is what makes the mitigated cases the real
story.

### Regulatory thresholds behind the calls

When a conclusion is anchored to a *specific* regulatory yardstick, which one? @fig-threshold-profile
profiles the thresholds the FONSI conclusions cite — dominated by resource-specific numeric limits
(*other quantitative*), **wetland/floodplain** extent, **NHPA §106** adverse-effect findings for
cultural resources, **visual (VRM)** classes, and the **Endangered Species Act** (take and
jeopardy). This mirrors the resource pattern above: the same areas that drive mitigation are the
ones with the hardest regulatory lines to clear.

```{r}
#| label: fig-threshold-profile
#| fig-cap: "Regulatory thresholds cited across primary FONSI determinations. Reported descriptively: identifying the exact threshold is the pipeline's least-accurate field, and many conclusions are not anchored to a specific threshold at all."
include_fig("fig_threshold_profile.png")
```

::: {.callout-note appearance="minimal"}
Threshold identification is the extraction's weakest field (see the validation above), so treat
this profile as an indicative picture of *which rules recur*, not an exact census.
:::

### Mitigated FONSIs — the line itself

The most revealing determinations are the **mitigated** ones: impacts the agency says *would* be
significant, brought below the line only by required measures. These show exactly where the line
sits and what it costs to stay under it — and they are the *majority* of FONSIs (@fig-mitigated-share).

```{r}
#| label: fig-mitigated-share
#| fig-cap: "Share of FONSI documents that reach a not-significant finding only with committed mitigation (per-resource class signal)."
include_fig("fig_mitigated_share.png")
```

```{r}
#| label: mit-headline
mit_rate <- mit_doc$share_mitigated_class_signal[1]
top_mit  <- mit_res |> filter(n_determinations >= 20) |> arrange(desc(share_mit_class)) |>
  slice_head(n = 4) |> mutate(Resource = relabel(shared_resource_area, res_label))
top_str  <- paste0(top_mit$Resource, " (", percent(top_mit$share_mit_class, accuracy = 1), ")",
                   collapse = ", ")
```

About **`r fmt_pct(mit_rate)` of these FONSIs are mitigated** — at least one resource whose
not-significant conclusion depends explicitly on committed mitigation.

The mitigation burden is concentrated in **`r top_str`** — the resources most often needing
committed measures to keep a project below the significance line. That pattern is the deliverable's
core finding: biological, soils/geology, water, and cultural resources are where decarbonization
projects most frequently sit *at* the line and are pulled under it by mitigation.

```{r}
#| label: fig-mitigation-by-resource
#| fig-cap: "Share of a resource's FONSI conclusions that depend on committed mitigation (per-resource class signal). Point size scales with how often the resource is analyzed."
include_fig("fig_mitigation_by_resource.png")
```

Reading volume and mitigation-intensity together sharpens the picture. Resources in the **upper
right** are both analyzed often *and* mitigated often — the areas that most shape whether a
decarbonization project can take the faster FONSI path.

```{r}
#| label: fig-mitigation-landscape
#| fig-cap: "Each resource by how often it is analyzed (x) and how often its conclusion depends on mitigation (y). Biological resources dominate both; soils/geology are mitigated intensively despite lower volume. Dashed line = overall average mitigation share."
include_fig("fig_mitigation_landscape.png")
```

What do these mitigations actually look like across the board? Two verbatim examples per resource,
ordered from the most mitigation-heavy resources down — seasonal timing and species avoidance for
biology, wetland minimization for water, cultural-resource protection measures, dust control for
air, and so on:

```{r}
#| label: tbl-mitigation-examples
#| tbl-cap: "Representative mitigation measures behind less-than-significant-with-mitigation conclusions, by resource area."
if (nrow(mit_ex) > 0) {
  mit_ex |>
    gt(groupname_col = "Resource") |>
    cols_label(example = "Example mitigation drawn from the FONSI") |>
    cols_align(align = "left") |>
    cols_width(example ~ pct(100)) |>
    tab_style(style = list(cell_fill(color = NAVY), cell_text(color = "white", weight = "bold")),
              locations = cells_row_groups()) |>
    tab_style(style = cell_text(color = NAVY, weight = "bold"), locations = cells_column_labels()) |>
    tab_options(table.font.size = px(12.5), table.width = pct(100), table.align = "center",
                row_group.font.weight = "bold", data_row.padding = px(6)) |>
    tab_source_note("Verbatim from the model's per-resource rationale on determinations that depend on committed mitigation. Climate/GHG had no such conclusions in the FONSI corpus.")
}
```

Mitigated FONSIs are also **broader** analyses: a project that needs committed mitigation for any
resource tends to address more resource areas overall — a median of roughly six versus four for
non-mitigated FONSIs.

```{r}
#| label: fig-breadth
#| fig-cap: "Distinct resource areas addressed per FONSI, split by whether the document relies on committed mitigation. Boxes show the median and interquartile range; the shaded shape is the full distribution."
include_fig("fig_breadth_per_fonsi.png")
```

::: {.callout-warning appearance="minimal"}
**Read mitigation at the right grain.** The *document-level* "is this a mitigated FONSI?" rate is
robust. The *resource-level* attribution above uses the high-precision per-resource signal; a
broader structured signal exists but over-attributes across resources, and its accuracy is
ultimately bounded by upstream tagging (validation F1 ≈ 0.62). Treat the resource shares as a
well-grounded floor, not an exact census.
:::

### Does it differ by agency?

The two departments run different kinds of projects — BLM tends toward land-based sites, the DOE
family toward energy facilities and transmission — so the resources that push them to mitigate
differ too. @fig-dept-by-resource compares, for each resource, how often a conclusion depends on
mitigation for BLM versus DOE.

```{r}
#| label: fig-dept-by-resource
#| fig-cap: "Share of each resource's FONSI conclusions that depend on committed mitigation, BLM versus the DOE family. A wider gap means the resource is a bigger mitigation driver for one department than the other."
include_fig("fig_dept_by_resource.png")
```

The pattern is intuitive: **biological, cultural/historic, air quality, and visual** resources
drive mitigation harder for **BLM** (land-facing impacts on habitat, historic sites, and
viewsheds), while **noise and water** lean toward the **DOE family** (facility- and
construction-driven effects). Within DOE, the picture varies further by sub-agency
(@fig-subagency-by-resource) — though the smaller sub-agency cells are noisy and should be read for
pattern, not precision.

```{r}
#| label: fig-subagency-by-resource
#| fig-cap: "Mitigation-dependent share by sub-agency and resource area, for sub-agencies with at least 40 determinations. Darker = a larger share of that resource's conclusions depend on mitigation. Small cells are noisy."
include_fig("fig_subagency_by_resource.png")
```

## EIS analysis {#sec-eis}

The FONSI track shows how agencies stay below the significance line. The EIS track is its mirror —
the above-the-line record of where impacts actually **cross** it. It runs the identical
extraction-and-validation machinery on the EIS corpus (same controlled vocabulary, per-resource
grain, and hand-labeled held-out test) and adds three EIS-only fields the FONSI side doesn't need:
which alternative a finding attaches to, why the impact is significant, and how it reaches the
resource.

```{r}
#| label: eis-scope
eis_sig_classes <- c("significant_adverse", "significant_unavoidable", "eis_required")
eis_n_det   <- sum(eis_class$n_determinations)
eis_n_proj  <- eis_mit$n_projects[1]
eis_n_docs  <- eis_mit$n_documents[1]
eis_n_sig   <- sum(eis_class$n_determinations[eis_class$determination_class %in% eis_sig_classes])
eis_n_unav  <- eis_class$n_determinations[eis_class$determination_class == "significant_unavoidable"]
eis_class_label <- c(class_label, eis_required = "EIS required")
efun <- function(k) { v <- eis_funnel$n[eis_funnel$metric == k]; if (length(v)) v[1] else NA }
```

The decarbonization EIS corpus is the larger of the two — 753 projects. Not all carry usable
findings: significance sections were retrieved for **`r fmt_n(efun("projects_with_sections"))`**, of
which **`r fmt_n(efun("projects_analyzed"))`** yielded at least one determination (@fig-eis-funnel).
Unlike the FONSI track, the EIS analysis covers **all agencies**, not just BLM + DOE: it is purely
descriptive — which resources cross the line — and never uses a project's decision date, so projects
without a firm date are kept rather than dropped. On that basis it rests on
`r fmt_n(eis_n_proj)` projects (`r fmt_n(eis_n_docs)` documents) carrying `r fmt_n(eis_n_det)`
distinct significance determinations — of which `r fmt_n(eis_n_sig)` land above the line, including
`r fmt_n(eis_n_unav)` judged **significant and unavoidable**, impacts mitigation cannot erase.

```{r}
#| label: fig-eis-funnel
#| fig-cap: "How the 753-project EIS corpus narrows to the analyzed set (top left): 536 had significance sections retrieved, 506 yielded a determination. Those projects produced 13,240 determinations, 2,198 above the line, 698 significant and unavoidable (top right). Of the 506 analyzed projects, only 239 have a firm in-window decision date — the analysis keeps the undated ones because it never uses the date (bottom)."
include_fig("fig_eis_funnel.png")
```

::: {.callout-tip}
## Key takeaways (EIS track)
- **Visual impacts cross the line most.** Viewshed conclusions are judged significant more often than
  any other resource (~33% of visual determinations), and they most often reach *significant and
  unavoidable* — a changed landscape usually cannot be mitigated back.
- **Resources split into two shapes.** Some are managed below the line (soils/geology, water, and
  public health are routinely mitigated to "not significant" and rarely cross); others cross over
  (visual, land use, and air quality break through more often than they are mitigated below it).
- Magnitude, cumulative effects, and protected resources are the dominant reasons a call is
  significant — and roughly a third of significant findings rest on **cumulative** pathways, not the
  project's direct footprint alone.
- **Validated, and harder than FONSI.** The significance class agrees with human coders at 0.69 (vs
  0.81 for FONSI) — a genuinely finer judgment the coders themselves split on.
:::

### How accurate is the EIS extraction?

The EIS pipeline was graded exactly as the FONSI one: two reviewers coded a stratified
400-window sample, a third adjudicated, and **30% was held out** as an untouched test. The scores
are lower across the board — the EIS task is harder, and the report treats it accordingly.

```{r}
#| label: tbl-validation-eis
#| tbl-cap: "EIS extraction accuracy versus the hand-labeled gold set, full sample and held-out test. Lower than the FONSI track — the above-the-line distinctions are genuinely finer."
metric_lab <- c(candidate_is_determination        = "Finds a determination (per passage)",
                resource_determination_detection  = "Assigns the right resource",
                determination_class_macro_f1       = "Gets the significance class right",
                mitigation_dependent_f1            = "Flags mitigation-dependence",
                primary_threshold_type_accuracy    = "Identifies the regulatory threshold")
eis_val |>
  mutate(score = coalesce(f1, precision)) |>
  filter(scope %in% c("overall", "holdout"), metric %in% names(metric_lab)) |>
  select(metric, scope, score) |>
  pivot_wider(names_from = scope, values_from = score) |>
  mutate(metric = factor(metric, levels = names(metric_lab), labels = unname(metric_lab))) |>
  arrange(metric) |>
  gt() |>
  cols_label(metric = "What the pipeline is scored on", overall = "All 400", holdout = "Held-out test") |>
  fmt_number(columns = c(overall, holdout), decimals = 2) |>
  data_color(columns = holdout, colors = col_numeric(c("#fdecea", "#eaf3ea"), domain = c(0.4, 1))) |>
  cols_align(align = "left", columns = metric) |>
  tab_style(style = list(cell_fill(color = NAVY), cell_text(color = "white", weight = "bold")),
            locations = cells_column_labels()) |>
  tab_options(table.font.size = px(13), table.width = pct(100), table.align = "center") |>
  tab_source_note("Scores are F1 (0–1), except the threshold row, which is accuracy. Held-out = the 30% of the gold set the pipeline never influenced.")
```

```{r}
#| label: fig-validation-eis
#| fig-cap: "EIS agreement with the human answer key: full sample (magenta) vs held-out test (green). The shaded bottom two rows are secondary attributes that matter less to the findings (as in the FONSI figure). Every score sits below the FONSI equivalent — the above-the-line call is a finer judgment (the two human coders themselves agreed only ~58% on the class)."
include_fig("fig_validation_accuracy_eis.png")
```

The pipeline still **finds the determination reliably (0.84)** and places it in the right resource
about **two-thirds** of the time; the significance *class* lands at **0.69**. That last number is
below the FONSI bar for a real reason: separating "significant" from "significant *and
unavoidable*" is a fine distinction the human coders split on too. We therefore lead with the
robust cut — *which resources cross the line* — and treat the adverse-vs-unavoidable split as
directional.

### Which resources cross the line

Within EISs, the interesting determinations are the ones that break **above** the line. Those come in
two degrees of severity:

- **Significant adverse** — the impact is significant, but the agency does not declare it beyond
  repair; committed measures may still reduce it.
- **Significant unavoidable** — the impact is significant *and* cannot be brought below the line even
  with mitigation. This is the more severe call, and the one that most constrains a project.

@fig-eis-above-line ranks each resource by how often its EIS conclusions are judged significant,
splitting the two degrees (least severe on the left, most severe on the right).

```{r}
#| label: fig-eis-above-line
#| fig-cap: "Share of each resource's EIS determinations judged significant, sorted (top = most likely to cross the line). Each percentage (and count) is of the resource's total EIS determinations: the in-bar labels split significant-adverse (lighter, left) from significant-unavoidable (darker, right); the end label gives the combined significant share and count (e.g. visual = 316 of 959, of which 126 are unavoidable)."
include_fig("fig_eis_above_line.png")
```

```{r}
#| label: eis-cross
eis_top <- eis_res |> filter(!n_suppressed) |> arrange(desc(share_sig)) |> slice_head(n = 4) |>
  mutate(R = relabel(shared_resource_area, res_label),
         lab = paste0(R, " (", percent(share_sig, accuracy = 1), ")"))
```

`r eis_top$lab[1]`, `r eis_top$lab[2]`, `r eis_top$lab[3]`, and `r eis_top$lab[4]` are the resources
most likely to cross into significance. **Visual stands alone at the top** — the clearest case of an
impact that, once significant, tends to stay that way.

### The wall: significant *and* unavoidable

The starkest determinations are the significant and unavoidable ones — where the agency concludes
the impact is real, material, and **cannot be mitigated below the line**. These are the true limits
on a decarbonization project. @fig-eis-unavoidable counts them by resource — each bar is the number
of distinct *significant-and-unavoidable* determinations (one per document × resource) for that
resource, across the analyzed EIS projects.

```{r}
#| label: fig-eis-unavoidable
#| fig-cap: "Count of significant-unavoidable determinations by resource — the darker (right-hand) segments of the previous figure, now counted. Visual's 126 means 126 of its 316 significant determinations were judged unavoidable; biological, cultural, and air quality follow. These are the impacts that remain significant even after mitigation."
include_fig("fig_eis_unavoidable.png")
```

**Visual impacts dominate the wall.** A wind farm on a ridgeline or a transmission corridor across
open country changes a landscape in ways no mitigation fully undoes — so viewshed conclusions reach
*unavoidable* more than any other resource. Biological, cultural, and air-quality impacts follow,
where habitat loss, historic-setting change, or a regional pollutant burden can exceed what on-site
measures can offset. @tbl-eis-unavoidable-ex shows what these look like in the agencies' own words.

```{r}
#| label: tbl-eis-unavoidable-ex
#| tbl-cap: "Representative significant-and-unavoidable determinations, verbatim from the EIS, ordered from the resources that hit the wall most often."
if (nrow(eis_unav_ex) > 0) {
  eis_unav_ex |>
    gt(groupname_col = "Resource") |>
    cols_label(example = "Significant-unavoidable finding, in the agency's words") |>
    cols_align(align = "left") |>
    cols_width(example ~ pct(100)) |>
    tab_style(style = list(cell_fill(color = "#75246C"), cell_text(color = "white", weight = "bold")),
              locations = cells_row_groups()) |>
    tab_style(style = cell_text(color = "#75246C", weight = "bold"), locations = cells_column_labels()) |>
    tab_options(table.font.size = px(12.5), table.width = pct(100), table.align = "center",
                row_group.font.weight = "bold", data_row.padding = px(6)) |>
    tab_source_note("Verbatim from the model's per-determination rationale on significant-unavoidable conclusions.")
}
```

### FONSI vs EIS: two ways a resource can be a problem

Analyzing the two tracks separately pays off in one picture (@fig-fonsi-vs-eis), which plots each
resource by how often it is mitigated below the line (FONSIs, x-axis) against how often it crosses
the line (EISs, y-axis). **The magenta resources are more likely to be judged significant than to be
mitigated; the blue resources are the reverse.** (Both tracks are restricted to BLM + DOE here, so
the two are measured on the same footing — FONSI coverage is only partial outside those agencies.)

```{r}
#| label: fig-fonsi-vs-eis
#| fig-cap: "Each resource (BLM + DOE, so both tracks share a scope) plotted by how often it is mitigated below the line in FONSIs (x) versus how often it crosses into significance in EISs (y). The dashed diagonal is equal odds; the bottom legend gives the colors — magenta resources (above the line) are judged significant more often than mitigated, blue resources (below) the reverse. Point size ≈ determinations."
include_fig("fig_fonsi_vs_eis.png")
```

The split is the deliverable's sharpest structural finding:

- **More likely to be significant** (magenta) — visual, land use, air quality — break above the line
  more often than they are mitigated below it. When these become a problem, agencies more often
  accept a significant finding than engineer their way under it; visual is the extreme (within
  BLM + DOE it crosses ~38% of the time and is mitigated only ~11%).
- **More likely to be mitigated** (blue) — soils/geology, water, public health — show the opposite:
  frequently mitigated in FONSIs, rarely crossing into significance. These are the impacts the
  toolbox of committed measures reliably handles.
- Biological and cultural sit near the line — heavily analyzed, heavily mitigated, *and* still
  crossing at a meaningful rate. They shape a project on both sides of the line at once.

### What makes an impact significant

For every above-the-line determination the extraction also records why it is significant.
@fig-eis-drivers shows the dominant factors (top) and — the fuller picture — *where each factor
bites*: the resources each reason concentrates in (bottom).

```{r}
#| label: fig-eis-drivers
#| fig-cap: "Top: the dominant factor behind each significant determination. Bottom: where each factor bites — significant determinations by factor and resource. Protected-resource findings concentrate in cultural and biological resources; regulatory-threshold findings in air quality and noise; cumulative effects in biological, visual, and cultural. Darker cells = more determinations."
include_fig("fig_eis_significance_drivers.png")
```

```{r}
#| label: eis-factors
eis_fac_top <- eis_fac |> filter(!n_suppressed) |> arrange(desc(n)) |> slice_head(n = 3) |>
  pull(significance_factor)
fac_pretty <- c(magnitude = "sheer magnitude", cumulative = "cumulative effects",
                protected_resource = "a protected resource", regulatory_threshold = "a regulatory threshold")
```

The leading reasons an impact is called significant are `r fac_pretty[eis_fac_top[1]]`,
`r fac_pretty[eis_fac_top[2]]`, and `r fac_pretty[eis_fac_top[3]]`, and the heatmap shows they are
not interchangeable: **a protected resource drives significance mainly for cultural and biological
resources** (endangered species, historic properties), a **regulatory threshold** for air quality
and noise (numeric standards), and **cumulative effects** for biological, visual, and water. That
last one is notable for decarbonization siting — nearly three in ten significant findings rest on
impacts that build up across a landscape or airshed rather than on one project's direct footprint.
@tbl-eis-factor-ex gives verbatim examples of each.

```{r}
#| label: tbl-eis-factor-ex
#| tbl-cap: "What each significance factor looks like in the EIS text — verbatim examples, ordered from the most common factor down."
if (nrow(eis_fac_ex) > 0) {
  eis_fac_ex |>
    gt(groupname_col = "Factor") |>
    cols_label(example = "Significant finding, in the agency's words") |>
    cols_align(align = "left") |>
    cols_width(example ~ pct(100)) |>
    tab_style(style = list(cell_fill(color = "#75246C"), cell_text(color = "white", weight = "bold")),
              locations = cells_row_groups()) |>
    tab_style(style = cell_text(color = "#75246C", weight = "bold"), locations = cells_column_labels()) |>
    tab_options(table.font.size = px(12.5), table.width = pct(100), table.align = "center",
                row_group.font.weight = "bold", data_row.padding = px(6)) |>
    tab_source_note("Verbatim from the model's per-determination rationale, grouped by the dominant significance factor it recorded.")
}
```

::: {.callout-note appearance="minimal"}
**Findings attach mostly to the proposed action.** About half of significant determinations name a
specific alternative; where they do, the **proposed action** dominates, with named build and
no-action alternatives carrying the rest. Significance is overwhelmingly a judgment about *what the
applicant proposed*, not about the alternatives weighed against it.
:::

### Does it differ by agency?

Because the EIS track spans every lead agency (not just BLM + DOE), we can ask whether some agencies
cross the line more than others. They do — and the pattern tracks the kind of project each agency
runs (@fig-eis-by-agency).

```{r}
#| label: fig-eis-by-agency
#| fig-cap: "Share of a lead agency's EIS determinations judged significant, for agencies with at least 150 determinations. Blue = BLM and the DOE family (the two agencies with complete dataset coverage); teal = other agencies, shown for context with partial coverage. Point size ≈ number of determinations."
include_fig("fig_eis_by_agency.png")
```

```{r}
#| label: eis-agency
ea_ord <- eis_agency |> arrange(desc(share))
ea_top <- ea_ord |> slice_head(n = 2); ea_bot <- ea_ord |> slice_tail(n = 2)
```

The spread is real — from **`r ea_top$Agency[1]` (`r percent(ea_top$share[1], accuracy=1)`)** at the
top down to **`r ea_bot$Agency[2]` (`r percent(ea_bot$share[2], accuracy=1)`)**. Land- and
water-facing agencies — the **Army Corps** and **BLM** — cross into significance most often, which
fits their portfolios (dredge-and-fill permits, public-land rights-of-way with visual and habitat
exposure). The standardized nuclear and utility programs — the **NRC** and **TVA** — cross least,
consistent with siting on already-industrial land under tightly prescribed review. BLM and the DOE
family remain the only agencies with complete coverage; the others are shown for context.

::: {.callout-warning appearance="minimal"}
**EIS rates are a recall floor, not a census.** Unlike FONSIs — which ride on a curated set of
finding spans — EIS determinations are retrieved live from ~45,000 document sections by keyword and
pattern. The held-out test shows the pipeline recovers about 77% of the determinations a human finds
(some conclusions live in summary tables and records of decision the keyword pass doesn't reach). The
rates above describe the determinations we found; they are a well-grounded floor on where impacts
cross the line, not an exhaustive count.
:::

## Methods and Limitations

**Scope**

- **Decarbonization only** (`project_energy_type = 'Clean'`).
- **Resource areas.** We read *"across resource areas"* as the standard NEPA environmental
  categories the documents are organized around — air quality; water (including wetlands and
  floodplains); biological resources; cultural / historic; visual; noise; soils / geology;
  socioeconomics (including environmental justice); transportation; land use; climate / greenhouse
  gases; and public health (including hazardous materials). Agencies judge significance separately
  for each, so this set defines the rows of every resource map above.
- **Agency scope differs by track.** The **FONSI** track is limited to **BLM + DOE**: of the 452
  decarbonization FONSI projects, 427 (94.5%) are BLM + DOE and carry the headline, while 25 (5.5%,
  led by other agencies with only partial coverage or flagged for manual review) are held as context
  and never mixed into the primary rates. The **EIS** track instead covers **all agencies** — it is
  descriptive and does not rest on the completeness a rate comparison needs — with BLM + DOE called
  out only where a finding is agency-sensitive (e.g. the FONSI-vs-EIS figure, which restricts both
  tracks to BLM + DOE for a like-for-like comparison).

**Coverage — two extraction funnels**

- **FONSI coverage.** Of the 427 BLM + DOE FONSIs, only **193 projects (258 documents)** carry
  machine-extractable significance findings; the rest lacked the upstream finding-section
  extraction. Every FONSI rate describes those 193 projects — a **coverage limit of the source
  extraction, not a sampling choice**. FONSI finding sections derive from D6 artifacts.
- **EIS coverage and retrieval recall.** EIS determinations are retrieved **live from ~45,000
  document sections** by keyword and pattern (FONSIs instead ride on a curated span set). Of the 753
  EIS corpus projects, significance sections were retrieved for 536 and **506 projects (1,082
  documents, all agencies)** yielded a determination — the analyzed set (@fig-eis-funnel). Of those
  506, only 239 have a firm in-window decision date; the rest (206 undated, 57 pre-2009, 4 boundary)
  are kept because the EIS analysis is descriptive and never uses the date. The held-out test
  recovers **~77%** of the determinations a human finds — conclusions stated in summary-of-impacts
  tables and records of decision are the main miss. **Read the EIS rates as a well-grounded floor
  on where impacts cross the line, not an exhaustive census.**

**Validation**

- **The held-out test (both tracks).** Two analysts independently coded a stratified 400-window
  sample (each window read once; one label per resource conclusion); a third adjudicated every
  disagreement into a single gold answer key. A deterministic **30% of the windows was held out** —
  never used to build or tune the extraction — and the pipeline was scored separately on it. **The
  held-out scores are the ones reported.** Rare class/resource/threshold cells are reported
  descriptively, not as pass/fail macro-metrics.
- **EIS is validated lower than FONSI, for a real reason.** The EIS significance *class* agrees with
  the human key at **0.69** (vs 0.81 for FONSI). Separating *significant* from *significant and
  unavoidable* is a genuinely fine judgment — the two human coders themselves agreed only **~58%**
  on the EIS class (vs 68% for FONSI). We therefore lead with the robust cut (*which resources
  cross the line*) and treat the adverse-vs-unavoidable split as directional.
- **EIS exploratory fields are not gold-validated.** The `alternative_name`, `significance_factor`,
  and `impact_type` fields are captured for the above-the-line analysis but sit **outside** the
  hand-labeled answer key (which validates class, resource, mitigation, and threshold). Treat the
  by-factor and by-pathway breakdowns as indicative, not as validated measurements.

**EIS extraction specifics**

- **Window size.** EIS sections run longer than FONSI ones, so the EIS pass reads each section to
  **24,000 characters** (vs 16,000 for FONSI), capturing ~95% of sections in full; the ~5% that run
  longer can still lose a conclusion past the cap.
- **De-duplication.** A Draft and Final EIS repeat identical section text; such passages are counted
  **once per project**. Identical text across *different* projects is kept — it is legitimate
  per-project attribution.
- **EIS mitigation is the class signal only.** "Below the line only with committed mitigation" on
  the EIS side is the model's class judgment; there is **no** record-of-decision commitment linkage
  (the enforceable-condition artifacts are FONSI-only). ROD-commitment mitigation is out of scope
  for this pass.
- **Two dropped windows.** 2 of 21,854 EIS windows (0.009%) returned no result from the batch API
  (a transient error) and were recorded as missing rather than silently dropped — no material effect
  on any rate.

**Other**

- **D4 date/regime coverage.** Many projects lack a high-confidence decision date; regime cells are
  confidence-gated and thin post-2023. EIS decision dates are frequently proxies.
- **EPA review letters (Clean Air Act §309) are out of scope.** Under Section 309, the EPA
  independently reviews and *rates* other agencies' environmental impact statements. Those comment
  letters are not in this dataset, so the report makes no claim about EPA's own view of significance.
- **Suppressed cells.** Any cell below 5 is suppressed.

## Reproducibility

Every number and figure in this report is regenerated by committed code, run from the repository
root in the `nepa` conda environment. The two significance-extraction passes (FONSI and EIS) are
**billable LLM prerequisites** — run once, then the deterministic analysis reads their outputs.

``` bash
# 0. framework regime + corpus + cohorts (deterministic)
conda run -n nepa python phase2/code/deliverable02/00_resolve_framework_regime.py
conda run -n nepa python phase2/code/deliverable02/01_build_d2_inventory.py

# 1. FONSI extraction — billable Anthropic batch (Sonnet 5); --dry-run for cost only
conda run -n nepa python phase2/code/deliverable02/02_extract_fonsi_significance.py \
  --batch-run --model claude-sonnet-5

# 2. EIS extraction — billable Anthropic batch (Sonnet 5); ~21.9k windows (~$110–115)
conda run -n nepa python phase2/code/deliverable02/04_extract_eis_significance.py \
  --sample 0 --batch-run --model claude-sonnet-5

# 3. validate both tracks against the hand-labeled gold (held-out test)
conda run -n nepa python phase2/code/deliverable02/05_validate_significance.py             # FONSI
conda run -n nepa python phase2/code/deliverable02/05_validate_significance.py --track eis  # EIS

# 4. analysis tables + figures (writes the CSVs/PNGs this report reads)
Rscript phase2/code/deliverable02/06_analyze_significance.R

# 5. render
quarto render phase2/reports/deliverable02.qmd
```

Outputs land in `phase2/data/analysis/deliverable02/` (determination parquets, validation metrics)
and `phase2/output/deliverable02/analysis/` (CSV tables + figures); the hand-labeled answer keys
live under `…/deliverable02/gold/`. Full provenance — input/output paths, row counts, content
hashes, model, prompt/schema versions — is recorded in `significance_run_manifest.parquet`.

## Appendix: data dictionary

See `phase2/notes/deliverable02/data_dictionary.md` and the run manifest
`significance_run_manifest.parquet` (input+output paths, row counts, content hashes, model,
prompt/schema versions).
