---
title: "NEPA Decarbonization Technology Analysis: Deliverable 4"
subtitle: "Review Timelines for Categorical Exclusions, Environmental Assessments, and Environmental Impact Statements"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    toc-location: left
    fig-align: center
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(arrow)
library(dplyr)
library(tidyr)
library(readr)
library(stringr)
library(scales)
library(knitr)
library(gt)
library(here)

OUTPUT_DIR <- here("phase2", "output", "deliverable04")
FIG_DIR    <- file.path(OUTPUT_DIR, "figures")
DIAG_DIR   <- file.path(OUTPUT_DIR, "diagnostics")

# CATF brand navy (Phase 1 styling)
CATF_NAVY <- "#012169"

fmt_n   <- function(x) comma(x, accuracy = 1)
fmt_pct <- function(x, accuracy = 0.1) percent(x, accuracy = accuracy)
include_fig <- function(name) knitr::include_graphics(file.path(FIG_DIR, name))
read_diag <- function(name) read_csv(file.path(DIAG_DIR, name), show_col_types = FALSE)
kable_tbl <- function(x) knitr::kable(x, format = "pipe", align = "l")

# Navy-header gt table in the house style
navy_gt <- function(df, title = NULL) {
  g <- gt(df)
  if (!is.null(title)) g <- tab_header(g, title = title)
  g |>
    cols_align(align = "left") |>
    tab_style(style = list(cell_fill(color = CATF_NAVY),
                           cell_text(color = "white", weight = "bold", align = "left")),
              locations = cells_column_labels()) |>
    tab_options(table.font.size = px(12), table.width = pct(100))
}

# --- Coverage (complete = complete_clear + complete_with_proxy) ---
cov <- read_diag("d4_coverage_by_process.csv")
cov_tot <- cov |> group_by(process_type) |> summarise(total = sum(n), .groups = "drop")
cov_complete <- cov |>
  filter(timeline_status %in% c("complete_clear", "complete_with_proxy")) |>
  group_by(process_type) |> summarise(complete = sum(n), .groups = "drop") |>
  left_join(cov_tot, by = "process_type") |>
  mutate(pct = complete / total)
cpct <- function(pt) cov_complete$pct[cov_complete$process_type == pt]
cn   <- function(pt) cov_complete$complete[cov_complete$process_type == pt]

diag <- read_diag("d4_coverage_diagnostics.csv")
dpct <- function(pt, col) diag[[col]][diag$process_type == pt] / 100

# --- Durations (complete_clear) ---
dur <- read_diag("d4_duration_summary.csv")
dmed <- function(pt) dur$median_days[dur$process_type == pt]

# --- Pages (regulatory, all EA/EIS, all energy types) ---
pg <- read_diag("d4_pages_summary.csv")
pgv <- function(pt, per, col) pg[[col]][pg$process_type == pt & pg$fra_period == per]
```

```{=html}
<style>
figcaption, .figure-caption, .quarto-figure figcaption { text-align: left !important; }
</style>
```

## Executive Summary

::: callout-note
## Key Findings

-   **Timeline completeness varies sharply by review process.** After LLM adjudication, complete timelines (both an initiation and a decision date) reach **`r fmt_pct(cpct("CE"))` for CE**, **`r fmt_pct(cpct("EA"))` for EA**, and **`r fmt_pct(cpct("EIS"))` for EIS** (see @fig-coverage). The binding constraint differs by process: CE and EA are *initiation*-limited (no Notice of Intent requirement, often no recorded start), while EIS is *decision*-limited (Records of Decision are frequently separate documents absent from the corpus — EIS decision coverage is only `r fmt_pct(dpct("EIS","pct_decision"))`).
-   **Durations follow the expected complexity ladder.** Median review duration is **`r dmed("CE")` days for CE**, **`r dmed("EA")` days (~`r round(dmed("EA")/30.4)` months) for EA**, and **`r fmt_n(dmed("EIS"))` days (~`r round(dmed("EIS")/365.25,1)` years) for EIS** (all complete timelines; see @fig-duration-intervals).
-   **LLM adjudication added ~13,000 dates** across 11,207 routed projects at a measured cost of **\$18.20**, lifting complete coverage roughly **+7 percentage points** in every process versus the deterministic-only pipeline.
-   **Timeline outliers split into genuinely long reviews and extraction errors.** The genuinely long cases — e.g. **SunZia Southwest Transmission (14.6 yr)**, **Energia Sierra Juarez**, **Grain Belt Express**, **Cushman Hydroelectric** — are strong candidates for CATF case studies. Errors (an "initiation" anchored to a historical citation) are flagged and excluded (see @sec-outliers).
-   **Post-FRA EIS documents are shorter; EA documents are essentially unchanged.** Across **all `r fmt_n(sum(pg$n_projects))` EA/EIS projects** (all energy types, regulatory pages computed from the full page text), mean regulatory EIS length fell from **`r pgv("EIS","Pre-FRA","mean_pages")` to `r pgv("EIS","Post-FRA","mean_pages")` pages** after the Fiscal Responsibility Act, while EA length held near **`r pgv("EA","Pre-FRA","mean_pages")`–`r pgv("EA","Post-FRA","mean_pages")` pages** — though the post-FRA sample is small (see @sec-pages).
:::

This report delivers:

> Timelines for categorical exclusions, environmental assessments, and environmental impact statements — including segmentation by regulatory period (pre/post-FRA), energy category, and decision year — plus a timeline-outlier deliverable for case-study investigation and a document-length / FRA page-limit analysis refreshed on Phase 2 decision dates.

------------------------------------------------------------------------

# Methodology

This deliverable **extends the [Phase 1 timeline analysis](https://www.kaseyzapatka.com/nepa/phase1/reports/deliverable03.html#timeline-analysis-for-decarbonization-technologies)** — which examined review timelines for *decarb projects only* — to the **entire NEPATEC 2.0 database**: every categorical exclusion, environmental assessment, and environmental impact statement, across decarb, fossil, and other project types. It does so with a rebuilt extraction pipeline that adds authoritative agency dates and two machine-learning models.

## How timeline dates are extracted

This section improves on the Phase 1 timeline method. For every project we identify two milestones — the **initiation** (when the review begins: an application received, a Notice of Intent, or scoping) and the **decision** (when it concludes: a signature, a Record of Decision, or a Finding of No Significant Impact). The pipeline finds these in five high-level steps:

1.  **Extract candidate dates** from every NEPA document in NEPATEC 2.0, capturing the sentence around each date as well as document location and metadata so its meaning is clear.
2.  **Classify each date's role** — a machine-learning *classifier* labels every candidate as an `initiation`, `decision`, or `neither`.
3.  **Rank the candidates** — a second *ranking model* ranks initiation and decision candidates by their likely to be the final, decisive date for the project.
4.  **Anchor on authoritative dates** — where the BLM or DOE NEPA register records an official project-start or decision date, that date is used.
5.  **Resolve the hardest cases with an LLM** — for the projects where the operative date is still ambiguous, a large language model selects the best remaining candidate.

A project's timeline is **complete** when both an initiation and a decision date are found.

## How Phase 2 analysis improves upon Phase 1

-   **Full-database coverage** — Phase 1 covered decarbonization technology projects only (20,725); Phase 2 covers the entire NEPATEC 2.0 database (61,881 projects).
-   **API registry dates** — Phase 2 incorporates authoritative dates from *BLM and DOE NEPA registers* as an authoritative source of official project-start and decision dates. We pull in about 40,000 register dates, which is particularly helpful for CE and EA reviews.
-   **Cleaner retrieval and granularity handling** — section-aware retrieval pulls fewer but better candidate dates based on document location and the pipeline uses year-month decision dates (e.g. a Final-EIS cover month) where no better dates are available.
-   **Improved classifier *AND* a ranking model** — Phase 1 used a single date classifier (`DeBERTa` model); Phase 2 used an improve classifer (`SetFit`) and adds a dedicated **ranking model** (`LightGBM`) that decides which candidate is the operative initiation/decision, sharply improving which candidates get sent to the LLM adjudicating model.
-   **LLM adjudication at scale** — the hardest ambiguous cases across the whole corpus are resolved by a large language model, recovering dates the automated pipeline could not select on its own.

## Data Quality & Caveats


### Fixable pipeline or data quality issues

- **About 2,000 register date errors.** About 2,000 document text dates are erroneously used in place of API register dates. These can be recovered with a little bit of coding elbow grease.
- **Negative-duration stopgap.** 233 decision-before-initiation rows are reclassified to `invalid_order` at analysis time; a source-level fix in script 05 is tracked.
- **Outlier extraction errors.** — extra long "durations" where the initiation is a stray/historical date (a citation, a prior plan, a facility-construction date) rather than the NEPA start. The genuinely-long vs error split is curated in 10_outliers.R.
- **LLM API call returned no date.** About `2,000` project dates were returned as null -- hoping to dig into these to see if we can recover them with some coding elbow grease. 


### Known data or timeline quality issues

- **Heavy reliance on proxy dates.** ~55% of CE and ~68% of EIS complete timelines rest on a *proxy* date that stands in for a milestone the documents never recorded directly. The proxy differs by process (below). *Separately*, where an extracted date is only has `year-month` granularity (e.g. a Final-EIS cover month like "June 2015"), the day is imputed to the **mid-month 15th** to minimize bias; the few year-only dates are excluded from the duration figures.
  - **CE — imputed initiations.** Most CEs record no distinct start, so the initiation is imputed: the pipeline takes the **earliest qualifying signature/start date** on the form (often the initiator's signature, which precedes the approving official's). ~13,100 CE initiations are proxies.
  - **EIS — Final-EIS-publication decisions.** Only ~18% of EISs have a Record of Decision in the corpus, so for another ~18% the pipeline uses the **Final-EIS publication date** (often reported as month + year → imputed to the 15th) as a stand-in decision. ~58% of EISs still end up with no decision date at all.
- **EAs are register-heavy.** EAs have no Notice-of-Intent requirement and often skip scoping, so a documented start date is frequently missing — **54% of EA initiations come from the BLM/DOE register** rather than the EA document. Because those register "project start" dates are often **late administrative entries made near the decision** (some only 1–2 days before the FONSI), register-anchored EAs show artificially **short durations** (~60-day median vs ~12 months for document-based starts).


::: callout-note
## Next steps

-   **Lift EIS decision coverage** — re-pull Final-EIS cover/publication dates to see if we can add more year-month dates for EIS.
-   **Fix the ordering check at the source** — resolve the 233 decision-before-initiation rows in the selection step (currently handled by a downstream stopgap filter).
-   **Extend the analysis** — BLM field-office learning curves, geothermal timelines by field office, and whether post-FRA adoption of categorical exclusions correlates with faster reviews (see Requested Extensions).

------------------------------------------------------------------------

:::

## Date provenance

Every extracted date traces to one of two sources: the **BLM/DOE NEPA registers** (official agency dates pulled via the registers' metadata API) or **document text** (parsed from the NEPA documents themselves, including LLM-adjudicated picks and Final-EIS publication dates). @fig-source shows the split by process and endpoint — a transparency check on where the timeline rests. Registers anchor a meaningful share of CE dates; EA and EIS rely overwhelmingly on document text.

```{r}
#| label: fig-source
#| fig-cap: "Source of each project's initiation and decision date by review process — Register API (BLM/DOE NEPA registers) vs Doc Text (parsed from documents) vs No Date (not extracted)."
#| out-width: "100%"
include_fig("fig_d4_register_source_projects.png")
```

------------------------------------------------------------------------

# Timeline Analysis

## Timeline completeness

@fig-complete-share shows the share of projects with complete timelines by review process; @fig-coverage breaks down *which* component is missing. The patterns differ by process:

-   **CE** (`r fmt_pct(cpct("CE"))` complete): decision coverage is high (`r fmt_pct(dpct("CE","pct_decision"))`) but initiation coverage is low (`r fmt_pct(dpct("CE","pct_initiation"))`) — most CEs are single-signature determinations with no recorded start.
-   **EA** (`r fmt_pct(cpct("EA"))` complete): the strongest of the three, but still initiation-limited — EAs have no Notice-of-Intent requirement and often skip scoping.
-   **EIS** (`r fmt_pct(cpct("EIS"))` complete): initiations are usually documented (an NOI is required), but **decision** coverage is only `r fmt_pct(dpct("EIS","pct_decision"))` because Records of Decision are frequently separate documents not in the corpus.

**Comparison with Phase 1.** Phase 1 measured completeness for *decarb projects only* and reported [CE ~30% · EA ~62% · EIS ~48%](https://www.kaseyzapatka.com/nepa/phase1/reports/deliverable03.html#fig-complete-share). On the **full database**, Phase 2 reaches CE `r fmt_pct(cpct("CE"))` · EA `r fmt_pct(cpct("EA"))` · EIS `r fmt_pct(cpct("EIS"))`. CE is *higher* — the much larger universe is dominated by routine, register-dated determinations — but **EA and EIS land below their Phase 1 decarb-only rates**, for two reasons this analysis surfaced:

-   **The aggregate is diluted by non-decarb projects.** Phase 1's denominator was decarb energy; Phase 2's is everything. Decarbonization EIS alone completes at **42.4%** (much closer to Phase 1's 48%) — it is the heterogeneous "Other" bucket that pulls the EIS aggregate down to `r fmt_pct(cpct("EIS"))`.
-   **The remaining gaps are structural, not fixable by better parsing.** EIS completeness is capped by **missing Records of Decision** (a separate document, often outside the corpus — decision coverage only `r fmt_pct(dpct("EIS","pct_decision"))`); CE and EA completeness is capped by **missing initiations**, because neither requires a Notice of Intent and a start date is frequently never recorded. A minority of EIS "projects" are also comment letters or document fragments with no milestone date to extract. Lifting EIS coverage toward Phase 1 is the focus of the [Next steps](#methodology) above.

```{r}
#| label: fig-complete-share
#| fig-cap: "Share of projects with a complete timeline (both initiation and decision dates) by NEPA review process. Dot marks the completion rate."
#| out-width: "90%"
include_fig("fig_d4_complete_timeline_share_boxplot.png")
```

```{r}
#| label: fig-coverage
#| fig-cap: "Timeline coverage breakdown by review process — both dates, decision only, initiation only, or neither."
#| out-width: "100%"
include_fig("fig_d4_coverage_by_process.png")
```

Coverage also varies by energy category (@fig-cov-energy). Notably, **decarbonization EIS projects complete at 42.4% — well above the EIS aggregate of `r fmt_pct(cpct("EIS"))`** — because decarb EISs are better documented than the heterogeneous "Other" bucket that drags the aggregate down.

```{r}
#| label: fig-cov-energy
#| fig-cap: "Timeline coverage by process and energy category (both dates / decision only / initiation only / none)."
#| out-width: "100%"
include_fig("fig_d4_coverage_by_process_and_energy.png")
```

## Duration analysis

Durations use **all complete timelines** (both dates present, including those resolved with a proxy date such as a Final-EIS publication date in place of a missing Record of Decision). Month-only dates are imputed to the **mid-month 15th**; the small number of year-only dates (almost entirely a subset of CE initiation dates) are excluded, since a day cannot be responsibly imputed from a year alone.

On that basis, durations track the expected complexity ladder from CE to EA to EIS — a median of about **`r dmed("CE")` days for CE**, **`r dmed("EA")` days (~`r round(dmed("EA")/30.4)` months) for EA**, and **`r fmt_n(dmed("EIS"))` days (~`r round(dmed("EIS")/365.25,1)` years) for EIS** (n = `r fmt_n(dur$n[dur$process_type=="CE"])` / `r fmt_n(dur$n[dur$process_type=="EA"])` / `r fmt_n(dur$n[dur$process_type=="EIS"])`). @fig-duration-intervals summarises the spread; @fig-duration-hist shows the full distributions; @fig-spans plots individual project spans.

```{r}
#| label: fig-duration-intervals
#| fig-cap: "Timeline duration by review process. Thin bar = 10th–90th percentile; thick bar = interquartile range (25th–75th); point = median. All complete timelines."
#| out-width: "100%"
include_fig("fig_d4_duration_summary_intervals.png")
```

Splitting duration by energy category (@fig-dur-energy) exposes a striking artifact: **fossil EAs run far shorter than decarbonization or "other" EAs** — a median of ~40 days, vs ~310 days for Decarb and ~209 days for Other. This is not because fossil EAs are genuinely faster: **84% of them are oil-and-gas well-abandonment EAs whose start date comes from the BLM register**, where the recorded "project start" is a *late administrative entry made within days of the decision*. Because fossil is the **largest** EA group (754 of the ~1,730 complete EAs), those artificially short spans **pull the overall EA median down** — the truer EA review length is better reflected by the document-anchored (decarb/other) EAs.

```{r}
#| label: fig-dur-energy
#| fig-cap: "Timeline duration by process and energy category. Fossil EAs are short largely because most are register-anchored well-abandonment EAs (late administrative start dates)."
#| out-width: "100%"
include_fig("fig_d4_duration_summary_intervals_by_energy.png")
```

The full duration distributions (@fig-duration-hist) show the same complexity ladder as a *shape*, not just a median: **CE durations pile up near zero** (most are days-to-weeks determinations), **EA spreads out to a year or two**, and **EIS is broad and heavy-tailed** — a meaningful mass of reviews runs 3–10+ years. That long EIS tail is real (large, contested reviews spanning multiple administrations) and is examined directly in the outlier case studies below.

```{r}
#| label: fig-duration-hist
#| fig-cap: "Review duration distribution by process; all complete timelines (month-only dates imputed to mid-month)."
#| out-width: "100%"
include_fig("fig_d4_duration_histogram.png")
```

```{r}
#| label: fig-spans
#| fig-cap: "Individual project timelines from initiation to decision, sorted by duration within process (complete timelines, sampled)."
#| out-width: "100%"
include_fig("fig_d4_project_timeline_spans.png")
```

## Projects by decision year

@fig-by-year shows project volume by decision year (all projects), faceted by review process. CE volumes ramp from the late-2000s (ARRA-era funding) with a second rise into the 2020s (BIL/IRA); EIS decisions lag CE and EA by several years, reflecting longer review cycles. @fig-by-year-doe restricts to the **Department of Energy** — the single largest agency in the dataset — where the ARRA/BIL/IRA grant-and-loan cycles are even more pronounced.

```{r}
#| label: fig-by-year
#| fig-cap: "Reviews by decision year (all projects), faceted by NEPA review process."
#| out-width: "100%"
include_fig("fig_d4_projects_by_decision_year.png")
```

```{r}
#| label: fig-by-year-doe
#| fig-cap: "DOE projects by decision year, faceted by review process (Department of Energy projects only)."
#| out-width: "100%"
include_fig("fig_d4_projects_by_decision_year_doe.png")
```

## Timeline Outliers: Case-Study Candidates {#sec-outliers}

Per the deliverable request, timeline outliers are surfaced for CATF staff to investigate via case studies (including whether NEPA itself was a cause of delay). Outliers are init→decision spans **longer than 5,000 days (~13.7 years)**, produced reproducibly by `code/deliverable04/10_outliers.R`.

**The critical distinction is real vs. extraction error.** A multi-year span can be a genuinely protracted review *or* an artifact where the "initiation" was anchored to a historical citation (a facility's construction date, a reserve's founding, a prior plan, a regulatory-compliance milestone) rather than the NEPA start. The two are separable only by reading the evidence text, which the outlier table preserves.

::: callout-note
## How to read these (process matters)

-   **EIS outliers are mostly real.** EIS median duration is already ~3 years; a 14–17-year EIS sits in the legitimate tail (transmission, restoration, resource-management plans genuinely run that long).
-   **CE outliers are mostly errors** — CEs are fast determinations (median `r dmed("CE")` days), so a multi-year CE almost always reflects a mis-extracted initiation. CE outliers are therefore excluded from the client list.
-   **Initiation dates before ~1990 are a strong error signal.** Of the four EIS outliers with 1980s-or-earlier initiations, three are confirmed errors (the "initiation" is a park founding, a prior lock's construction-completion date, or a RCRA compliance milestone).
:::

The decarb EA/EIS candidates below are the highest-value case studies. **SunZia** (ROW application Sep 2008 → Record of Decision Apr 2023) is the textbook long-permitting saga; **Energia Sierra Juarez**, **Grain Belt Express**, and **Cushman Hydroelectric** are similar. *Confirmed errors are excluded* — e.g. the Palisades Nuclear Restart "19-year" span glues a 2005 license-renewal application to a 2024 restart Record of Decision (two different NEPA actions).

```{r}
#| label: tbl-outliers
#| tbl-cap: "Genuinely long EA/EIS reviews — case-study candidates. Full list with evidence text in output/deliverable04/diagnostics/d4_duration_outliers_client.csv. Project IDs are full UUIDs."
client <- read_diag("d4_duration_outliers_client.csv")
# Curated exclusions: confirmed cross-action extraction errors found on evidence review.
exclude_ids <- c("da6bbc7a3ca3aeca7bc1c06eaafa8d44")  # Palisades restart (license-renewal init)
client |>
  filter(!project_id %in% exclude_ids) |>
  arrange(desc(duration_years)) |>
  transmute(`Project` = project_title,
            `Process` = process_type,
            `Energy` = project_energy_type,
            `Lead agency` = str_remove_all(lead_agency_harmonized, '\\[|\\]|"'),
            `Years` = round(duration_years, 1),
            `Initiation` = as.character(initiation_date),
            `Decision` = as.character(decision_date),
            `project_id` = project_id) |>
  head(18) |>
  navy_gt()
```

------------------------------------------------------------------------

# Document Length & FRA Page Limits {#sec-pages}

This section **builds directly on [Phase 1 Deliverable 5](https://www.kaseyzapatka.com/nepa/phase1/reports/deliverable05.html)** ("Document Length Over Time and Fiscal Responsibility Act Impact"), which analyzed document length for *decarb* EA/EIS projects. Here we reproduce that analysis on the **entire Phase 2 EA/EIS corpus, all energy types**, refreshed on Phase 2 LLM-adjudicated **decision dates** (inclusion requires only a decision date — the time axis and FRA classification depend only on it). Each subsection below links to its Phase 1 counterpart for comparison.

::: callout-note
## Method & scope

-   **Regulatory pages** = body word count ÷ 500, excluding embedded appendices and low-content pages, per **40 C.F.R. § 1508.1(bb)** — the measure that matches the FRA limits (EA ≤ 75 pages; EIS ≤ 150, or 300 if extraordinarily complex).[^rawpages]
-   These counts are computed **from the actual page text** by `fra/01_extract_pages.py` (a DuckDB pass over `phase2/data/processed/{ea,eis}/pages.parquet` — 6.1M EIS pages — detecting the appendix boundary and low-content pages), producing regulatory pages for **5,032 EA/EIS projects** (2,765 EA, 2,267 EIS) across **all energy types** — not just decarbonization.
-   The pre/post-FRA analysis below uses the **`r fmt_n(sum(pg$n_projects))` projects** that also have a decision date (EA `r fmt_n(sum(pg$n_projects[pg$process_type=="EA"]))` / EIS `r fmt_n(sum(pg$n_projects[pg$process_type=="EIS"]))`).
-   **FRA date = enactment (June 3, 2023)**, matching Phase 1 D5 and the timeline-duration period split above.
:::

[^rawpages]: We use *regulatory* pages rather than raw PDF page counts because raw counts substantially overstate the length the statute limits — they include embedded appendices and sparse pages (covers, maps, dividers). Mean raw pages run roughly 2× regulatory pages for EAs and higher for EISs, so a raw count cannot be compared to the 75/150-page limits.

## Document length over time

@fig-pages-time tracks regulatory page length by decision date across all EA/EIS. Compare the decarb-only Phase 1 version: [Phase 1 D5 — Document Length Over Time](https://www.kaseyzapatka.com/nepa/phase1/reports/deliverable05.html#fig-pages-over-time).

```{r}
#| label: fig-pages-time
#| fig-cap: "Regulatory page counts for individual EA/EIS projects (all energy types) by decision date, coloured by FRA period; navy line = 3-month rolling mean. Red dashed line = FRA enactment (June 3, 2023)."
#| out-width: "100%"
include_fig("fig_d4_pages_over_time.png")
```

## Pre vs post-FRA

Across all EA/EIS, **EIS regulatory length fell after FRA** (mean `r pgv("EIS","Pre-FRA","mean_pages")` → `r pgv("EIS","Post-FRA","mean_pages")` pages; median `r pgv("EIS","Pre-FRA","median_pages")` → `r pgv("EIS","Post-FRA","median_pages")`), while **EA length was essentially flat** (mean `r pgv("EA","Pre-FRA","mean_pages")` → `r pgv("EA","Post-FRA","mean_pages")`). The post-FRA window is short, so these are provisional (n = `r pgv("EA","Post-FRA","n_projects")` EA, `r pgv("EIS","Post-FRA","n_projects")` EIS). Compare the decarb-only Phase 1 result: [Phase 1 D5 — Pre vs Post FRA](https://www.kaseyzapatka.com/nepa/phase1/reports/deliverable05.html#fig-pre-post) and its [distribution view](https://www.kaseyzapatka.com/nepa/phase1/reports/deliverable05.html#fig-distribution).

```{r}
#| label: tbl-pages
#| tbl-cap: "Regulatory page counts by process and FRA period — all EA/EIS, all energy types. Regulatory pages = body word count ÷ 500."
pg |>
  transmute(Process = process_type, `FRA period` = fra_period, `n` = n_projects,
            Mean = mean_pages, Median = median_pages, SD = sd_pages,
            p25 = p25_pages, p75 = p75_pages) |>
  navy_gt()
```

```{r}
#| label: fig-pages-bar
#| fig-cap: "Mean regulatory pages, pre- vs post-FRA. Bar = mean; diamond = median; n shown."
#| out-width: "100%"
include_fig("fig_d4_pages_pre_post_fra.png")
```

Breaking the same comparison out **by energy category** (@fig-pages-energy) shows the post-FRA EIS decline is **broad** — it appears in decarbonization *and* fossil EIS, not just one segment — while EA length is essentially flat across all categories. That consistency makes the EIS drop more credible as a genuine post-FRA effect rather than a composition artifact.

```{r}
#| label: tbl-pages-energy
#| tbl-cap: "Regulatory page counts by process, energy category, and FRA period (all EA/EIS). Regulatory pages."
read_diag("d4_pages_summary_by_energy.csv") |>
  transmute(Process = process_type, Energy = energy, `FRA period` = fra_period,
            `n` = n, `Mean pages` = mean_pages, `Median pages` = median_pages) |>
  navy_gt()
```

```{r}
#| label: fig-pages-energy
#| fig-cap: "Mean regulatory pages by energy category, pre- vs post-FRA. Labels = n projects."
#| out-width: "100%"
include_fig("fig_d4_pages_pre_post_fra_by_energy.png")
```

The full distributions (@fig-pages-dist) confirm the shift is not an outlier effect: the **EIS distribution moves down** post-FRA (lower median, thinner upper tail) while the **EA distribution barely moves**. Splitting that distribution by energy (@fig-pages-dist-energy) shows the EIS contraction again holds across Decarb and Fossil.

```{r}
#| label: fig-pages-dist
#| fig-cap: "Distribution of regulatory page counts, pre- vs post-FRA (violin + box; diamond = median; y capped at p99)."
#| out-width: "100%"
include_fig("fig_d4_pages_distribution.png")
```

```{r}
#| label: fig-pages-dist-energy
#| fig-cap: "Distribution of regulatory page counts by energy category, pre- vs post-FRA (violin + box; diamond = median; y capped at p99)."
#| out-width: "100%"
include_fig("fig_d4_pages_distribution_by_energy.png")
```

## FRA page-limit compliance

```{r}
#| label: fig-pages-compliance
#| fig-cap: "FRA page-limit compliance among post-FRA projects (regulatory pages). EA limit 75; EIS 150 (300 if extraordinarily complex). The bracket marks the EIS share within the 300-page extraordinary-complexity threshold."
#| out-width: "90%"
include_fig("fig_d4_pages_compliance.png")
```

------------------------------------------------------------------------

# Requested Extensions (Planned / Not Yet Built)

The following items from the deliverable request require additional analysis beyond the current timeline build. They are scoped here for the next phase:

```{r}
tibble::tibble(
  `Requested analysis` = c(
    "BLM field-office learning curve",
    "Geothermal timelines by BLM field office",
    "CE adoption post-FRA vs. review speed"
  ),
  `What it needs` = c(
    "Join decision dates to BLM field-office identifiers; test whether per-office median duration falls as cumulative application count rises (experience effect).",
    "Geothermal project flags (Deliverable 1/3 tech tags) × field office × duration; limited by geothermal sample size per office.",
    "CE-adoption events (which agencies adopted other agencies' CEs post-FRA, BLM/DOE/USFS) × subsequent review duration; needs the CE-adoption register."
  )
) |>
  navy_gt()
```

------------------------------------------------------------------------

# Reproduction

Run from the repository root in the `nepa` conda environment:

``` bash
# Timeline pipeline (candidates -> classify -> select -> LLM adjudicate)
conda run -n nepa python phase2/code/deliverable04/run_pipeline.py
ANTHROPIC_API_KEY=$(security find-generic-password -s nepa-anthropic -w) \
  conda run -n nepa python phase2/code/deliverable04/06_adjudicate_llm.py \
  --mode candidate_adjudication --process CE EA EIS --model claude-haiku-4-5-20251001 --workers 24

# Analysis + figures
Rscript phase2/code/deliverable04/08_analyze.R       # coverage, durations, FRA, energy
Rscript phase2/code/deliverable04/10_outliers.R      # timeline outliers (case-study candidates)

# FRA document-length analysis (regulatory pages from the full page text, all EA/EIS)
conda run -n nepa python phase2/code/deliverable04/fra/01_extract_pages.py --run
Rscript phase2/code/deliverable04/fra/02_pages_fra.R # document length / FRA page limits
```

Outputs are written to `phase2/output/deliverable04/` (`figures/`, `diagnostics/`).

------------------------------------------------------------------------

*Draft generated `r Sys.Date()` | NEPA Decarbonization Technology Analysis — Phase 2, Deliverable 4*
