---
title: "NEPA Decarbonization Technology Analysis: Deliverable 2"
subtitle: "Determinations of Significance Across Resource Areas"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    fig-align: center
execute:
  echo: false
  warning: false
  message: false
---

<!-- Plan: phase2/plans/deliverable02.md (v2.11). Pipeline: phase2/code/deliverable02/
     00 regime -> 01 corpus+cohorts -> 02 FONSI extract -> 04 EIS extract -> 05 validate ->
     06 analyze.R (writes the CSV tables this report reads). Regenerate 06 before rendering. -->

```{r setup}
library(arrow); library(dplyr); library(tidyr); library(readr)
library(stringr); library(knitr); library(gt); library(here); library(scales)
options(arrow.skip_nul = TRUE)

A   <- here("phase2", "data", "analysis", "deliverable02")
TAB <- here("phase2", "output", "deliverable02", "analysis")
rp  <- function(f) read_parquet(file.path(A, f))
rc  <- function(f) { p <- file.path(TAB, f); if (file.exists(p)) read_csv(p, show_col_types = FALSE) else tibble() }
fmt <- function(x) formatC(x, format = "d", big.mark = ",")
fmt_n   <- function(x) comma(x, accuracy = 1)
fmt_pct <- function(x, accuracy = 1) percent(x, accuracy = accuracy)
NAVY <- "#012169"; BLUE <- "#0047BB"
include_fig <- function(f) { p <- file.path(TAB, f); if (file.exists(p)) knitr::include_graphics(p) }

det    <- rp("significance_determinations.parquet")
corpus <- rp("significance_corpus.parquet")
regime <- rp("project_regime.parquet")

# analysis tables written by 06_analyze_significance.R (regenerate before rendering)
val          <- rp("validation_metrics.parquet")
res_by_class <- rc("resource_by_class.csv")
class_dist   <- rc("class_distribution_dual_denominator.csv")
mit_doc      <- rc("mitigation_document_level.csv")
mit_res      <- rc("mitigation_by_resource.csv")
mit_ex       <- rc("mitigation_examples.csv")

# EIS (above-the-line) track — parallel tables written by 06 + the EIS validation parquet
eis_val   <- tryCatch(rp("validation_metrics_eis.parquet"), error = function(e) tibble())
eis_class <- rc("eis_class_distribution.csv")
eis_res   <- rc("eis_resource_significance.csv")
eis_cmp   <- rc("eis_fonsi_vs_eis.csv")
eis_fac   <- rc("eis_significance_factors.csv")
eis_mit   <- rc("eis_mitigation_document_level.csv")
has_eis   <- nrow(eis_class) > 0 && nrow(eis_val) > 0

# friendly labels for the controlled vocab
class_label <- c(no_significant_impact = "No significant impact",
                 less_than_significant = "Less than significant",
                 less_than_significant_with_mitigation = "Committed mitigation",
                 significant_adverse = "Significant adverse",
                 significant_unavoidable = "Significant & unavoidable")
res_label <- c(air_quality="Air quality", water="Water", biological="Biological",
               cultural="Cultural / historic", visual="Visual", noise="Noise",
               soils_geology="Soils / geology", socioeconomic="Socioeconomic",
               transportation="Transportation", land_use="Land use",
               climate_ghg="Climate / GHG", public_health="Public health",
               project_wide="Project-wide (FONSI)", unknown="Unplaced")
relabel <- function(x, map) ifelse(is.na(map[x]) | map[x] == "", x, map[x])

is_dry  <- all(det$extraction_method == "regex")
has_val <- nrow(val) > 0 && !is_dry
```

::: {.callout-warning}
## Draft status
```{r}
#| results: asis
cat(if (is_dry)
  paste("This document outlines the deliverable's purpose and the analysis workflow; the",
        "findings themselves are still in progress. Results tables will be added once the",
        "extraction pass and the validation against the hand-labelled gold set are complete.")
  else paste("This report reflects the validated extraction; see Methods & required caveats",
             "for the applicable limitations."))
```
:::

This report delivers:

> Determinations of significance across resource areas; factors that contribute to a determination of "significant impact," starting with mitigated FONSIs.

---

## Context

When a decarbonization project — a solar farm, a transmission line — needs federal approval, the
lead agency (here, BLM or the Department of Energy) studies its environmental impact and makes a
judgment call: **is the impact "significant" or not?**

That one word decides a lot. If the agency finds the impact **not significant**, the project
moves ahead on a faster track (a document called a *Finding of No Significant Impact*, or FONSI).
If it finds the impact **significant**, that triggers a much longer, more demanding review.

Often the agency lands in between: *"this impact **would** be significant — **unless** the
developer commits to specific measures to reduce it."* That is a **mitigated finding**, and it is
the most revealing case, because it shows the exact line the agency drew and what it required to
stay on the safe side of it.

**So this deliverable answers:** across these decarbonization projects, how do agencies decide what
counts as "significant" in each environmental area (air, water, wildlife, cultural sites, and so
on) — and what mitigation do they demand to avoid that label? Answering it means reading and
coding a judgment call out of **thousands** of dense decision documents — far more than anyone
could review by hand. The workflow below is how we do that reliably.

## Workflow

The approach pairs the **scale of automation** with the **reliability of human judgment**:
software and AI do the reading at a scale no person could, while analysts define the scope, hold
the checkpoints, and grade the AI against their own answer key before any result counts.

```{=html}
<div style="max-width:660px;margin:1.25rem auto;">
  <div style="border:2px solid #0047BB;border-radius:8px;padding:.65rem 1rem;background:#f4f7fc;">
    <div style="font-weight:700;color:#012169;font-size:1.05rem;">1 · Scope</div>
    <div style="color:#1A1A2E;">Decarbonization FONSIs led by BLM or the DOE family &mdash; 258 documents across 193 projects, each dated to its regulatory era</div>
  </div>
  <div style="text-align:center;color:#0047BB;font-size:1.5rem;line-height:1.1;">&#8595;</div>
  <div style="border:2px solid #0047BB;border-radius:8px;padding:.65rem 1rem;background:#f4f7fc;">
    <div style="font-weight:700;color:#012169;font-size:1.05rem;">2 · Gather documents</div>
    <div style="color:#1A1A2E;">Pull the FONSI and supporting EA documents from the NEPATEC&nbsp;2.0 corpus, then use text-pattern (regex) rules to locate the specific sections where significance findings live</div>
  </div>
  <div style="text-align:center;color:#0047BB;font-size:1.5rem;line-height:1.1;">&#8595;</div>
  <div style="border:2px solid #0047BB;border-radius:8px;padding:.65rem 1rem;background:#f4f7fc;">
    <div style="font-weight:700;color:#012169;font-size:1.05rem;">3 · Extract the judgment calls</div>
    <div style="color:#1A1A2E;"><b>Automated:</b> the flagged passages go to Claude (Sonnet&nbsp;5), which codes each one &mdash; outcome, resource area, threshold, mitigation.<br><b>In parallel &mdash; the answer key:</b> two analysts independently hand-label a 400-passage sample and an adjudicator settles their disagreements</div>
  </div>
  <div style="text-align:center;color:#0047BB;font-size:1.5rem;line-height:1.1;">&#8595;</div>
  <div style="border:2px solid #0047BB;border-radius:8px;padding:.65rem 1rem;background:#f4f7fc;">
    <div style="font-weight:700;color:#012169;font-size:1.05rem;">4 · Validate</div>
    <div style="color:#1A1A2E;">Score the AI against the held-out human labels (class agreement 0.81) before any result counts</div>
  </div>
  <div style="text-align:center;color:#0047BB;font-size:1.5rem;line-height:1.1;">&#8595;</div>
  <div style="border:2px solid #012169;border-radius:8px;padding:.65rem 1rem;background:#012169;">
    <div style="font-weight:700;color:#ffffff;font-size:1.05rem;">5 · Analyze and report</div>
    <div style="color:#ffffff;">Validated findings become this report and a source-linked data table &mdash; every number traces to its page</div>
  </div>
  <div style="text-align:center;color:#6b7280;font-size:.85rem;margin-top:.5rem;">
    The five steps of the analysis &mdash; each card matches a numbered step below.
  </div>
</div>
```

## What the two document types tell us

A significance call can land on either side of the line, and **the two document types record
opposite outcomes** — so we analyze them separately rather than pooling them:

- **FONSIs are the *below-the-line* record.** The agency concluded the impact is **not**
  significant. This is where the **mitigated** cases live — impacts that *would* have been
  significant without committed measures. FONSIs tell us **how agencies stay under the line, and
  what they require to get there.**
- **EISs are the *above-the-line* record.** These are impacts the agency judged **significant**
  (and, at the extreme, significant and *unavoidable*). They tell us **which resources actually
  cross the line.**

@fig-corpus-overview sets the scale: the decarbonization universe splits into a larger EIS corpus
and a smaller FONSI corpus, and within the FONSIs the analysis touches every major resource area.

```{r}
#| label: fig-corpus-overview
#| fig-cap: "Left: decarbonization projects by review type (the EIS corpus is the larger one). Right: the mix of resource areas covered by the FONSI determinations, one square ≈ 1% of determinations."
include_fig("fig_corpus_overview.png")
```

## FONSI analysis {#sec-fonsi}

```{r}
#| label: fonsi-corpus
proj_rows <- str_detect(class_dist$determination_class, "project-level")
n_prim <- sum(class_dist$n_determinations[!proj_rows])
n_docs <- mit_doc$n_documents[1]
n_proj <- mit_doc$n_projects[1]
```

The decarbonization FONSI corpus contains **452 projects**. This analysis rests on the
**`r fmt_n(n_proj)` of them** that are (a) led by BLM or the DOE family and (b) carry
machine-extractable significance findings — spanning **`r fmt_n(n_docs)` decision documents** and
**`r fmt_n(n_prim)` distinct significance determinations** (one per document × resource area ×
outcome).

Two filters narrow the 452. First, the **BLM + DOE scope** sets aside 25 FONSIs led by other
agencies (@fig-agency-scope). Second — the larger reduction — only FONSIs whose finding sections
were located by the upstream pipeline carry determinations, which brings 427 BLM + DOE projects
down to `r fmt_n(n_proj)`. That second gap is a **coverage limitation of the source extraction, not
a different finding**: the `r fmt_n(n_proj)` analyzed projects are representative of the corpus, but
the rates below describe *those* projects, not all 452 (see *Methods*).

```{r}
#| label: fig-agency-scope
#| fig-cap: "Lead agency for the 452 decarbonization FONSIs. BLM and the DOE family (94.5%) carry the headline analysis; the remaining 5.5% — led by other federal agencies with only partial coverage — are set aside as context."
include_fig("fig_agency_scope.png")
```

::: {.callout-tip}
## Key takeaways (FONSI track)
- **Most decarbonization FONSIs are *mitigated*.** About **`r fmt_pct(mit_doc$share_mitigated_class_signal)`**
  of these documents reach "no significant impact" only because the developer commits to specific
  measures — the impact *would* have been significant otherwise.
- **A handful of resources carry the mitigation.** **Biological resources, soils/geology, water, and
  cultural/historic sites** are where projects most often sit at the line and are pulled under it —
  biological alone is mitigation-dependent in ~30% of its determinations.
- **The below-the-line default is "no significant impact."** Across resource areas, outright
  no-significant-impact is the most common outcome; genuinely *significant* findings live in the EIS
  record, not here.
- **The numbers are validated.** The extraction agrees with a hand-labeled human answer key at
  **0.81** on the significance class (held-out test) before any figure in this report was produced.
:::

### How accurate is the extraction?

Every determination is machine-extracted, so before any number is reported the pipeline is scored
against a **hand-labeled answer key**: two independent reviewers coded a stratified 400-window
sample, a third adjudicated their disagreements, and **30% was held out** as an untouched test. The
held-out column is the honest score.

```{r}
#| label: tbl-validation
#| tbl-cap: "Extraction accuracy versus the hand-labeled gold set, full sample and held-out test."
metric_lab <- c(candidate_is_determination        = "Finds a determination (per passage)",
                resource_determination_detection  = "Assigns the right resource",
                determination_class_macro_f1       = "Gets the significance class right",
                mitigation_dependent_f1            = "Flags mitigation-dependence",
                primary_threshold_type_accuracy    = "Identifies the regulatory threshold")
val |>
  mutate(score = coalesce(f1, precision)) |>
  filter(scope %in% c("overall", "holdout"), metric %in% names(metric_lab)) |>
  select(metric, scope, score) |>
  pivot_wider(names_from = scope, values_from = score) |>
  mutate(metric = factor(metric, levels = names(metric_lab), labels = unname(metric_lab))) |>
  arrange(metric) |>
  gt() |>
  cols_label(metric = "What the pipeline is scored on", overall = "All 400", holdout = "Held-out test") |>
  fmt_number(columns = c(overall, holdout), decimals = 2) |>
  data_color(columns = holdout, colors = col_numeric(c("#fdecea", "#eaf3ea"), domain = c(0.4, 1))) |>
  cols_align(align = "left", columns = metric) |>
  tab_style(style = list(cell_fill(color = NAVY), cell_text(color = "white", weight = "bold")),
            locations = cells_column_labels()) |>
  tab_options(table.font.size = px(13), table.width = pct(100), table.align = "center") |>
  tab_source_note("Scores are F1 (0–1), except the threshold row, which is accuracy. Held-out = the 30% of the gold set the pipeline never influenced.")
```

In plain terms: the extraction **almost never misses a determination** and places it in the
**right resource area ~89%** of the time; it agrees with the human coders on the significance
*class* at **0.81** — the standard bar for this kind of task. The one weaker signal is
*mitigation-dependence*, discussed at the end of this section.

```{r}
#| label: fig-validation
#| fig-cap: "Agreement with the human answer key: the full 400-window sample (magenta) versus the held-out test (green). The dashed line marks 0.80, the standard bar for this kind of task."
include_fig("fig_validation_accuracy.png")
```

**Why the last two scores are lower — and why it doesn't undercut the findings.** Both are
*secondary attributes*, not the core call, and both are genuinely harder:

- *Mitigation-dependence (0.62).* Deciding whether a conclusion **depends on** committed mitigation
  is one of the hardest judgments in the whole task — our two human coders themselves agreed only
  ~68% of the time on the fine line between "less than significant" and "less than significant *with
  mitigation*." On top of that, the pipeline's structured mitigation flag is tagged at the document
  level, so it over-attributes across a multi-resource passage, and its precision is ultimately
  bounded by upstream tagging. **We handle this by reporting mitigation at the document level (where
  it is robust) and using the precise per-resource signal for the resource-level breakdown** — so
  the numbers you see below are the high-precision measure, not the noisy flag.
- *Regulatory threshold (0.66–0.71).* This is the most granular field — pinning a conclusion to a
  *specific* rule (a NAAQS limit, an ESA take threshold, a §106 adverse-effect finding). Many
  conclusions are not anchored to a single numeric threshold at all, and the distinctions are fine.
  It is reported descriptively as supporting context, never as a headline number.

The three scores that carry the findings — **finding the determination, placing the resource, and
getting the significance class** — all clear or approach the standard bar.

### How agencies stay below the line

Within FONSIs, the determinations sit — as they must — below the significance line: the agency
either finds **no significant impact** outright or an impact that is **less than significant**,
sometimes only **with committed mitigation**. @fig-outcomes-by-resource breaks this down by
resource area — each resource's determinations split across the three outcomes, sorted so the
resources most reliant on mitigation sit at the top, with the count and share behind each bar.

```{r}
#| label: fig-outcomes-by-resource
#| fig-cap: "Share of each resource's FONSI determinations by outcome, sorted by reliance on mitigation. Each segment shows its count and percentage; n is the resource's total determinations. Resources with more magenta lean harder on committed mitigation to reach a not-significant finding."
include_fig("fig_outcomes_by_resource.png")
```

The dominant outcome is **no significant impact**, as expected for FONSIs. But the resources
agencies analyze most — **biological, water, cultural/historic, and public health** — are also
where the close calls concentrate: the magenta band (below-the-line *only with mitigation*) is
widest for biological, soils/geology, and water. That is what makes the mitigated cases the real
story.

### Regulatory thresholds behind the calls

When a conclusion is anchored to a *specific* regulatory yardstick, which one? @fig-threshold-profile
profiles the thresholds the FONSI conclusions cite — dominated by resource-specific numeric limits
(*other quantitative*), **wetland/floodplain** extent, **NHPA §106** adverse-effect findings for
cultural resources, **visual (VRM)** classes, and the **Endangered Species Act** (take and
jeopardy). This mirrors the resource pattern above: the same areas that drive mitigation are the
ones with the hardest regulatory lines to clear.

```{r}
#| label: fig-threshold-profile
#| fig-cap: "Regulatory thresholds cited across primary FONSI determinations. Reported descriptively: identifying the exact threshold is the pipeline's least-accurate field, and many conclusions are not anchored to a specific threshold at all."
include_fig("fig_threshold_profile.png")
```

::: {.callout-note appearance="minimal"}
Threshold identification is the extraction's weakest field (see the validation above), so treat
this profile as an indicative picture of *which rules recur*, not an exact census.
:::

### Mitigated FONSIs — the line itself

The most revealing determinations are the **mitigated** ones: impacts the agency says *would* be
significant, brought below the line only by required measures. These show exactly where the line
sits and what it costs to stay under it — and they are the *majority* of FONSIs (@fig-mitigated-share).

```{r}
#| label: fig-mitigated-share
#| fig-cap: "Share of FONSI documents that reach a not-significant finding only with committed mitigation (per-resource class signal)."
include_fig("fig_mitigated_share.png")
```

```{r}
#| label: mit-headline
mit_rate <- mit_doc$share_mitigated_class_signal[1]
top_mit  <- mit_res |> filter(n_determinations >= 20) |> arrange(desc(share_mit_class)) |>
  slice_head(n = 4) |> mutate(Resource = relabel(shared_resource_area, res_label))
top_str  <- paste0(top_mit$Resource, " (", percent(top_mit$share_mit_class, accuracy = 1), ")",
                   collapse = ", ")
```

About **`r fmt_pct(mit_rate)` of these FONSIs are mitigated** — at least one resource whose
not-significant conclusion depends explicitly on committed mitigation.

The mitigation burden is concentrated in **`r top_str`** — the resources most often needing
committed measures to keep a project below the significance line. That pattern is the deliverable's
core finding: biological, soils/geology, water, and cultural resources are where decarbonization
projects most frequently sit *at* the line and are pulled under it by mitigation.

```{r}
#| label: fig-mitigation-by-resource
#| fig-cap: "Share of a resource's FONSI conclusions that depend on committed mitigation (per-resource class signal). Point size scales with how often the resource is analyzed."
include_fig("fig_mitigation_by_resource.png")
```

Reading volume and mitigation-intensity together sharpens the picture. Resources in the **upper
right** are both analyzed often *and* mitigated often — the areas that most shape whether a
decarbonization project can take the faster FONSI path.

```{r}
#| label: fig-mitigation-landscape
#| fig-cap: "Each resource by how often it is analyzed (x) and how often its conclusion depends on mitigation (y). Biological resources dominate both; soils/geology are mitigated intensively despite lower volume. Dashed line = overall average mitigation share."
include_fig("fig_mitigation_landscape.png")
```

What do these mitigations actually look like across the board? Two verbatim examples per resource,
ordered from the most mitigation-heavy resources down — seasonal timing and species avoidance for
biology, wetland minimization for water, cultural-resource protection measures, dust control for
air, and so on:

```{r}
#| label: tbl-mitigation-examples
#| tbl-cap: "Representative mitigation measures behind less-than-significant-with-mitigation conclusions, by resource area."
if (nrow(mit_ex) > 0) {
  mit_ex |>
    gt(groupname_col = "Resource") |>
    cols_label(example = "Example mitigation drawn from the FONSI") |>
    cols_align(align = "left") |>
    cols_width(example ~ pct(100)) |>
    tab_style(style = list(cell_fill(color = NAVY), cell_text(color = "white", weight = "bold")),
              locations = cells_row_groups()) |>
    tab_style(style = cell_text(color = NAVY, weight = "bold"), locations = cells_column_labels()) |>
    tab_options(table.font.size = px(12.5), table.width = pct(100), table.align = "center",
                row_group.font.weight = "bold", data_row.padding = px(6)) |>
    tab_source_note("Verbatim from the model's per-resource rationale on determinations that depend on committed mitigation. Climate/GHG had no such conclusions in the FONSI corpus.")
}
```

Mitigated FONSIs are also **broader** analyses: a project that needs committed mitigation for any
resource tends to address more resource areas overall — a median of roughly six versus four for
non-mitigated FONSIs.

```{r}
#| label: fig-breadth
#| fig-cap: "Distinct resource areas addressed per FONSI, split by whether the document relies on committed mitigation. Boxes show the median and interquartile range; the shaded shape is the full distribution."
include_fig("fig_breadth_per_fonsi.png")
```

::: {.callout-warning appearance="minimal"}
**Read mitigation at the right grain.** The *document-level* "is this a mitigated FONSI?" rate is
robust. The *resource-level* attribution above uses the high-precision per-resource signal; a
broader structured signal exists but over-attributes across resources, and its accuracy is
ultimately bounded by upstream tagging (validation F1 ≈ 0.62). Treat the resource shares as a
well-grounded floor, not an exact census.
:::

### Does it differ by agency?

The two departments run different kinds of projects — BLM tends toward land-based sites, the DOE
family toward energy facilities and transmission — so the resources that push them to mitigate
differ too. @fig-dept-by-resource compares, for each resource, how often a conclusion depends on
mitigation for BLM versus DOE.

```{r}
#| label: fig-dept-by-resource
#| fig-cap: "Share of each resource's FONSI conclusions that depend on committed mitigation, BLM versus the DOE family. A wider gap means the resource is a bigger mitigation driver for one department than the other."
include_fig("fig_dept_by_resource.png")
```

The pattern is intuitive: **biological, cultural/historic, air quality, and visual** resources
drive mitigation harder for **BLM** (land-facing impacts on habitat, historic sites, and
viewsheds), while **noise and water** lean toward the **DOE family** (facility- and
construction-driven effects). Within DOE, the picture varies further by sub-agency
(@fig-subagency-by-resource) — though the smaller sub-agency cells are noisy and should be read for
pattern, not precision.

```{r}
#| label: fig-subagency-by-resource
#| fig-cap: "Mitigation-dependent share by sub-agency and resource area, for sub-agencies with at least 40 determinations. Darker = a larger share of that resource's conclusions depend on mitigation. Small cells are noisy."
include_fig("fig_subagency_by_resource.png")
```

## EIS analysis {#sec-eis}

The FONSI track shows how agencies *stay below* the significance line. The EIS track is its mirror:
the *above-the-line* record of where impacts actually **cross** it. It runs the **identical
extraction-and-validation machinery** on the EIS corpus — same controlled vocabulary, same
per-resource grain, same hand-labeled answer key and held-out test — and adds three EIS-only fields
the FONSI side doesn't need: **which alternative** the finding attaches to, **why** the impact is
significant, and **how** it reaches the resource.

```{r}
#| label: eis-scope
eis_sig_classes <- c("significant_adverse", "significant_unavoidable", "eis_required")
eis_n_det   <- sum(eis_class$n_determinations)
eis_n_proj  <- eis_mit$n_projects[1]
eis_n_docs  <- eis_mit$n_documents[1]
eis_n_sig   <- sum(eis_class$n_determinations[eis_class$determination_class %in% eis_sig_classes])
eis_n_unav  <- eis_class$n_determinations[eis_class$determination_class == "significant_unavoidable"]
eis_class_label <- c(class_label, eis_required = "EIS required")
```

The decarbonization EIS corpus is the **larger** of the two (753 projects; @fig-corpus-overview).
Within the BLM + DOE primary scope, the analysis rests on **`r fmt_n(eis_n_proj)` projects**
(`r fmt_n(eis_n_docs)` documents) carrying **`r fmt_n(eis_n_det)` distinct significance
determinations** — of which **`r fmt_n(eis_n_sig)` land above the line** (significant), including
**`r fmt_n(eis_n_unav)` judged *significant and unavoidable*** — impacts mitigation cannot erase.

::: {.callout-tip}
## Key takeaways (EIS track)
- **Visual impacts cross the line most.** Of every resource, **viewshed** conclusions are the most
  likely to be judged significant (**~38%** of visual determinations) — and they most often reach
  *significant and unavoidable*, because a changed landscape usually cannot be mitigated back.
- **Two different problem shapes.** Some resources are *managed below the line* — **soils/geology,
  water, public health** are routinely mitigated to "not significant" in FONSIs and rarely cross.
  Others *cross over* — **visual, land use, air quality** more often break through into significance
  than they get mitigated below it.
- **Magnitude, cumulative effects, and protected resources drive it.** The dominant reasons an
  impact is called significant are **sheer magnitude**, **cumulative effects**, and the presence of
  a **protected resource** — and nearly a third of significant findings rest on **cumulative**
  impact pathways, not the project's direct footprint alone.
- **Validated, and honestly harder than FONSI.** The same held-out grading shows the EIS task is
  more difficult — the significance *class* agrees with human coders at **0.69** (vs 0.81 for
  FONSI), reflecting a genuinely finer judgment that the human coders themselves split on.
:::

### How accurate is the EIS extraction?

The EIS pipeline was graded exactly as the FONSI one: two reviewers coded a stratified
400-window sample, a third adjudicated, and **30% was held out** as an untouched test. The scores
are lower across the board — the EIS task is harder, and the report treats it accordingly.

```{r}
#| label: tbl-validation-eis
#| tbl-cap: "EIS extraction accuracy versus the hand-labeled gold set, full sample and held-out test. Lower than the FONSI track — the above-the-line distinctions are genuinely finer."
metric_lab <- c(candidate_is_determination        = "Finds a determination (per passage)",
                resource_determination_detection  = "Assigns the right resource",
                determination_class_macro_f1       = "Gets the significance class right",
                mitigation_dependent_f1            = "Flags mitigation-dependence",
                primary_threshold_type_accuracy    = "Identifies the regulatory threshold")
eis_val |>
  mutate(score = coalesce(f1, precision)) |>
  filter(scope %in% c("overall", "holdout"), metric %in% names(metric_lab)) |>
  select(metric, scope, score) |>
  pivot_wider(names_from = scope, values_from = score) |>
  mutate(metric = factor(metric, levels = names(metric_lab), labels = unname(metric_lab))) |>
  arrange(metric) |>
  gt() |>
  cols_label(metric = "What the pipeline is scored on", overall = "All 400", holdout = "Held-out test") |>
  fmt_number(columns = c(overall, holdout), decimals = 2) |>
  data_color(columns = holdout, colors = col_numeric(c("#fdecea", "#eaf3ea"), domain = c(0.4, 1))) |>
  cols_align(align = "left", columns = metric) |>
  tab_style(style = list(cell_fill(color = NAVY), cell_text(color = "white", weight = "bold")),
            locations = cells_column_labels()) |>
  tab_options(table.font.size = px(13), table.width = pct(100), table.align = "center") |>
  tab_source_note("Scores are F1 (0–1), except the threshold row, which is accuracy. Held-out = the 30% of the gold set the pipeline never influenced.")
```

```{r}
#| label: fig-validation-eis
#| fig-cap: "EIS agreement with the human answer key: full sample (magenta) vs held-out test (green). Every score sits below the FONSI equivalent — the above-the-line call is a finer judgment (the two human coders themselves agreed only ~58% on the class)."
include_fig("fig_validation_accuracy_eis.png")
```

The pipeline still **finds the determination reliably (0.84)** and places it in the right resource
about **two-thirds** of the time; the significance *class* lands at **0.69**. That last number is
below the FONSI bar for a real reason: separating "significant" from "significant *and
unavoidable*" is a fine distinction the human coders split on too. We therefore lead with the
robust cut — *which resources cross the line* — and treat the adverse-vs-unavoidable split as
directional.

### Which resources cross the line

Within EISs, the interesting determinations are the ones that break **above** the line.
@fig-eis-above-line ranks each resource by how often its EIS conclusions are judged significant,
splitting the merely *significant* from the *significant and unavoidable*.

```{r}
#| label: fig-eis-above-line
#| fig-cap: "Share of each resource's EIS determinations judged significant, sorted (top = most likely to cross the line). The darker segment is significant AND unavoidable — impacts mitigation cannot bring back below the line. Label shows the share and the count of significant determinations."
include_fig("fig_eis_above_line.png")
```

```{r}
#| label: eis-cross
eis_top <- eis_res |> filter(!n_suppressed) |> arrange(desc(share_sig)) |> slice_head(n = 4) |>
  mutate(R = relabel(shared_resource_area, res_label),
         lab = paste0(R, " (", percent(share_sig, accuracy = 1), ")"))
```

**`r eis_top$lab[1]`, `r eis_top$lab[2]`, `r eis_top$lab[3]`, and `r eis_top$lab[4]`** are the
resources most likely to cross into significance. Visual stands alone at the top — and it is the
clearest case of an impact that, once significant, tends to stay that way.

### The wall: significant *and* unavoidable

The starkest determinations are the **significant and unavoidable** ones — where the agency
concludes the impact is real, material, and **cannot be mitigated below the line**. These are the
true limits on a decarbonization project. @fig-eis-unavoidable counts them by resource.

```{r}
#| label: fig-eis-unavoidable
#| fig-cap: "Determinations judged significant AND unavoidable, by resource. Visual, biological, and air-quality impacts most often reach the point where mitigation cannot fix them."
include_fig("fig_eis_unavoidable.png")
```

**Visual impacts dominate the wall.** A wind farm on a ridgeline or a transmission corridor across
open country changes a landscape in ways no mitigation fully undoes — so viewshed conclusions reach
*unavoidable* more than any other resource. Biological and air-quality impacts follow, where habitat
loss or a regional pollutant burden can exceed what on-site measures can offset.

### FONSI vs EIS: two ways a resource can be a problem

Putting the two tracks side by side is the payoff of analyzing them separately. A resource can be a
problem in **two distinct ways**: it can be *mitigated below the line* (its FONSI conclusions lean
on committed measures) or it can *cross the line* (its EIS conclusions are judged significant).
These are not the same resources — @fig-fonsi-vs-eis contrasts them.

```{r}
#| label: fig-fonsi-vs-eis
#| fig-cap: "For each resource: how often it crosses into significance in EISs (magenta) versus how often it is mitigated below the line in FONSIs (blue). Resources far to the right on magenta cross over (often unmitigable); far to the right on blue are routinely managed below the line."
include_fig("fig_fonsi_vs_eis.png")
```

The contrast is the deliverable's sharpest structural finding:

- **Cross-over resources** — **visual, land use, air quality** — break above the line *more often
  than they are mitigated below it*. When these become a problem, agencies more often accept a
  significant finding than engineer their way under it. Visual is the extreme case (crosses ~38% of
  the time, mitigated only ~11%).
- **Managed-below resources** — **soils/geology, water, public health** — show the opposite:
  frequently mitigated in FONSIs, rarely crossing into significance. These are the impacts the
  toolbox of committed measures reliably handles.
- **Biological and cultural** sit in the middle — heavily analyzed, heavily mitigated, *and* still
  crossing the line at a meaningful rate. They are the resources that shape a project on both sides
  of the line at once.

### What makes an impact significant

For every above-the-line determination the extraction also records **why** it is significant and
**how** the impact reaches the resource. @fig-eis-drivers summarizes both.

```{r}
#| label: fig-eis-drivers
#| fig-cap: "Top: the dominant factor behind each significant determination. Bottom: the impact pathway. Cumulative effects are a leading driver on both — nearly a third of significant findings turn on impacts that accumulate rather than the project's direct footprint alone."
include_fig("fig_eis_significance_drivers.png")
```

```{r}
#| label: eis-factors
eis_fac_top <- eis_fac |> filter(!n_suppressed) |> arrange(desc(n)) |> slice_head(n = 3) |>
  pull(significance_factor)
fac_pretty <- c(magnitude = "sheer magnitude", cumulative = "cumulative effects",
                protected_resource = "a protected resource", regulatory_threshold = "a regulatory threshold")
```

The leading reasons an impact is called significant are **`r fac_pretty[eis_fac_top[1]]`**,
**`r fac_pretty[eis_fac_top[2]]`**, and **`r fac_pretty[eis_fac_top[3]]`**. The prominence of
**cumulative effects** — both as a stated factor and as an impact *pathway* — is notable for
decarbonization siting: the significant call frequently rests not on one project's direct footprint
but on its contribution to effects that build up across a landscape or airshed.

::: {.callout-note appearance="minimal"}
**Findings attach mostly to the proposed action.** About half of significant determinations name a
specific alternative; where they do, the **proposed action** dominates, with named build and
no-action alternatives carrying the rest. Significance is overwhelmingly a judgment about *what the
applicant proposed*, not about the alternatives weighed against it.
:::

::: {.callout-warning appearance="minimal"}
**EIS coverage is a recall floor, not a census.** Unlike FONSIs — which ride on a curated set of
finding spans — EIS determinations are retrieved live from ~45,000 document sections by keyword and
pattern. The held-out test shows the pipeline recovers about **77%** of the determinations a human
finds (some conclusions live in summary tables and records of decision the keyword pass doesn't
reach). The rates above describe the determinations we *found*; they are a well-grounded floor on
where impacts cross the line, not an exhaustive count.
:::

## Scope decisions to confirm

A few items shape the deliverable's scope and breadth. Each has a recommended default below;
the goal is to confirm or adjust these before the full analysis runs.

1. **FONSIs *and* EISs — both now delivered.** Significance calls appear in both document types:
   **FONSIs** capture the *not-significant* side (including the mitigated cases sitting right at the
   line), while **EISs** capture the *significant* side. Both tracks are now built and validated,
   and both appear above. The EIS side required extracting findings live from a
   **~6.1-million-page** document set — a substantially larger AI and analyst effort, run as a
   Message-Batch pass against Claude Sonnet 5 and graded against its own hand-labeled answer key.
   The one residual caveat is retrieval recall (below): EIS findings are a well-grounded floor, not
   an exhaustive census.

2. **Agencies: BLM + DOE, or broader?** These are the only two agencies with complete coverage in
   the dataset and account for the large majority of decarbonization FONSIs; other agencies (NRC,
   Forest Service, BOEM) appear mainly on the EIS side, with partial coverage.
   **Recommendation:** BLM + DOE as the primary comparison; other agencies reported as context
   only, never mixed into the headline rates.

3. **What counts as a "resource area"?** We read *"across resource areas"* as the standard NEPA
   environmental categories the documents are organized around — air quality; water (including
   wetlands and floodplains); biological resources; cultural / historic; visual; noise;
   soils / geology; socioeconomics (including environmental justice); transportation; land use;
   climate / greenhouse gases; and public health (including hazardous materials). Agencies judge
   significance separately for each, so this set defines the rows of the headline map.
   **Confirm:** is this the intended list, or does CATF have a specific set or a priority subset
   in mind?

## Methods & required caveats

- **Agency scope — what the BLM + DOE limit costs.** Of the **452 decarbonization FONSI projects** in
  the corpus, **427 (94.5%) are BLM + DOE** and carry the headline; **25 (5.5%) are dropped** from
  the headline rates — 23 led by other federal agencies (which have only partial coverage in the
  dataset) and 2 flagged for manual scope review. They are held as a separate context universe,
  never mixed into the primary rates. So the BLM + DOE limit sets aside about **1 in 20 FONSIs**.
  The **EIS side** (753 projects — the *above-the-line* story) is analyzed separately in its own
  section above, under the same BLM + DOE scope.
- **Extraction coverage (the bigger funnel).** Of the 427 BLM + DOE FONSIs, only **193 projects
  (258 documents)** carry machine-extractable significance findings; the rest lacked the upstream
  finding-section extraction that feeds this analysis. Every rate in this report describes those
  193 projects. This is a **coverage limit of the source extraction, not a sampling choice** — the
  analyzed set is representative of the corpus, but it is not all 452 FONSIs.
- **D4 date/regime coverage.** Many projects lack a high-confidence decision date; regime cells
  are confidence-gated and thin post-2023. EIS decision dates are frequently proxies.
- **EPA review letters (Clean Air Act §309) are out of scope.** Under Section 309, the EPA
  independently reviews and *rates* other agencies' environmental impact statements (e.g.
  "environmental concerns" or "objections"). Those comment letters are not in this dataset, so
  the report makes no claim about EPA's own view of a project's significance.
- **Decarbonization scope only** (`project_energy_type = 'Clean'`).
- **FONSI coverage** for the mitigated/straight tiers derives from D6 artifacts.
- **EIS retrieval recall.** EIS determinations are retrieved live from ~45,000 document sections by
  keyword and pattern (FONSIs ride on a curated span set instead). The held-out test recovers ~77%
  of the determinations a human finds; conclusions in summary tables and records of decision are the
  main miss. Read the EIS rates as a well-grounded floor on where impacts cross the line, not an
  exhaustive census.
- **Validation & the held-out test.** Two analysts independently coded a stratified 400-window
  sample (each window read once; one label per resource conclusion); a third analyst adjudicated
  every disagreement into a single gold answer key. A deterministic **30% of the windows was held
  out** — set aside and never used to build or tune the extraction — and the pipeline was scored
  separately on it. **The held-out scores are the ones reported**, so they can't be inflated by the
  pipeline having "seen" those windows. Rare class/resource/threshold cells are reported
  descriptively, not as pass/fail macro-metrics.
- **Suppressed cells.** Any cell below 5 is suppressed.

## Appendix: data dictionary

See `phase2/notes/deliverable02/data_dictionary.md` and the run manifest
`significance_run_manifest.parquet` (input+output paths, row counts, content hashes, model,
prompt/schema versions).
