---
title: "NEPA Decarbonization Technology Analysis: Deliverable 2"
subtitle: "Determinations of Significance Across Resource Areas"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    fig-align: center
execute:
  echo: false
  warning: false
  message: false
---

<!-- Plan: phase2/plans/deliverable02.md (v2.11). Pipeline: phase2/code/deliverable02/
     00 regime -> 01 corpus+cohorts -> 02 FONSI extract -> 04 EIS extract -> 05 validate ->
     06 analyze.R (writes the CSV tables this report reads). Regenerate 06 before rendering. -->

```{r setup}
library(arrow); library(dplyr); library(tidyr); library(readr)
library(stringr); library(knitr); library(gt); library(here); library(scales)
options(arrow.skip_nul = TRUE)

A   <- here("phase2", "data", "analysis", "deliverable02")
TAB <- here("phase2", "output", "deliverable02", "analysis")
rp  <- function(f) read_parquet(file.path(A, f))
rc  <- function(f) { p <- file.path(TAB, f); if (file.exists(p)) read_csv(p, show_col_types = FALSE) else tibble() }
fmt <- function(x) formatC(x, format = "d", big.mark = ",")
fmt_n   <- function(x) comma(x, accuracy = 1)
fmt_pct <- function(x, accuracy = 1) percent(x, accuracy = accuracy)
NAVY <- "#012169"; BLUE <- "#0047BB"
include_fig <- function(f) { p <- file.path(TAB, f); if (file.exists(p)) knitr::include_graphics(p) }

det    <- rp("significance_determinations.parquet")
corpus <- rp("significance_corpus.parquet")
regime <- rp("project_regime.parquet")

# analysis tables written by 06_create_figures.R (regenerate before rendering)
val          <- rp("validation_metrics.parquet")
res_by_class <- rc("resource_by_class.csv")
class_dist   <- rc("class_distribution_dual_denominator.csv")
mit_doc      <- rc("mitigation_document_level.csv")
mit_res      <- rc("mitigation_by_resource.csv")
mit_ex       <- rc("mitigation_examples.csv")
fonsi_tech   <- rc("fonsi_technology.csv")
fonsi_enf    <- rc("fonsi_enforceability.csv")

# EIS (above-the-line) track — parallel tables written by 06 + the EIS validation parquet
eis_val   <- tryCatch(rp("validation_metrics_eis.parquet"), error = function(e) tibble())
eis_class <- rc("eis_class_distribution.csv")
eis_res   <- rc("eis_resource_significance.csv")
eis_cmp   <- rc("eis_fonsi_vs_eis.csv")
eis_fac   <- rc("eis_significance_factors.csv")
eis_mit   <- rc("eis_mitigation_document_level.csv")
eis_funnel <- rc("eis_coverage_funnel.csv")
eis_unav_ex <- rc("eis_unavoidable_examples.csv")
eis_fac_ex  <- rc("eis_factor_examples.csv")
eis_agency  <- rc("eis_agency.csv")
eis_tech    <- rc("eis_technology.csv")
eis_mitigable <- rc("eis_mitigable.csv")
eis_mit_ex  <- rc("eis_mitigable_examples.csv")
has_eis   <- nrow(eis_class) > 0 && nrow(eis_val) > 0

# friendly labels for the controlled vocab
class_label <- c(no_significant_impact = "No significant impact",
                 less_than_significant = "Less than significant",
                 less_than_significant_with_mitigation = "Committed mitigation",
                 significant_adverse = "Significant adverse",
                 significant_unavoidable = "Significant unavoidable")
res_label <- c(air_quality="Air quality", water="Water", biological="Biological",
               cultural="Cultural / historic", visual="Visual", noise="Noise",
               soils_geology="Soils / geology", socioeconomic="Socioeconomic",
               transportation="Transportation", land_use="Land use",
               climate_ghg="Climate / GHG", public_health="Public health",
               project_wide="Project-wide (FONSI)", unknown="Unplaced")
relabel <- function(x, map) ifelse(is.na(map[x]) | map[x] == "", x, map[x])

is_dry  <- all(det$extraction_method == "regex")
has_val <- nrow(val) > 0 && !is_dry
```

This report delivers:

> Determinations of significance across resource areas; factors that contribute to a determination of "significant impact," starting with mitigated FONSIs.

---

## Context

When a decarbonization project — a solar farm, a transmission line — needs federal approval, the
lead agency (here, BLM or the Department of Energy) studies its environmental impact and makes a
judgment call: is the impact "significant" or not?

That one word decides a lot. If the agency finds the impact **not significant**, the project
moves ahead on a faster track (a document called a *Finding of No Significant Impact*, or FONSI).
If it finds the impact **significant**, that triggers a much longer, more demanding review.

Often the agency lands in between: *"this impact would be significant — unless the
developer commits to specific measures to reduce it."* That is a **mitigated finding**, and it is
the most revealing case, because it shows the exact line the agency drew and what it required to
stay on the safe side of it.

So this deliverable answers: across these decarbonization projects, how do agencies decide what
counts as "significant" in each environmental area (air, water, wildlife, cultural sites, and so
on) — and what mitigation do they demand to avoid that label? Answering it means reading and
coding a judgment call out of thousands of dense decision documents — far more than anyone
could review by hand. The workflow below is how we do that reliably.

## Workflow

The approach pairs the **scale of automation** with the **reliability of human judgment**:
software and AI do the reading at a scale no person could, while analysts define the scope, hold
the checkpoints, and grade the AI against their own answer key before any result counts.

```{=html}
<div style="max-width:660px;margin:1.25rem auto;">
  <div style="border:2px solid #0047BB;border-radius:8px;padding:.65rem 1rem;background:#f4f7fc;">
    <div style="font-weight:700;color:#012169;font-size:1.05rem;">1 · Scope</div>
    <div style="color:#1A1A2E;">Two document tracks: decarbonization <b>FONSIs</b> led by BLM or the DOE family (258 documents across 193 projects, each dated to its regulatory era) and final <b>EISs</b> from every lead agency (1,082 documents across 506 projects)</div>
  </div>
  <div style="text-align:center;color:#0047BB;font-size:1.5rem;line-height:1.1;">&#8595;</div>
  <div style="border:2px solid #0047BB;border-radius:8px;padding:.65rem 1rem;background:#f4f7fc;">
    <div style="font-weight:700;color:#012169;font-size:1.05rem;">2 · Gather documents</div>
    <div style="color:#1A1A2E;">Pull the FONSI and supporting EA documents &mdash; and, on the EIS track, the final EIS impact chapters &mdash; from the NEPATEC&nbsp;2.0 corpus, then use text-pattern (regex) rules to locate the specific sections where significance findings live</div>
  </div>
  <div style="text-align:center;color:#0047BB;font-size:1.5rem;line-height:1.1;">&#8595;</div>
  <div style="border:2px solid #0047BB;border-radius:8px;padding:.65rem 1rem;background:#f4f7fc;">
    <div style="font-weight:700;color:#012169;font-size:1.05rem;">3 · Extract the judgment calls</div>
    <div style="color:#1A1A2E;"><b>Automated:</b> the flagged passages go to Claude (Sonnet&nbsp;5), which codes each one &mdash; outcome, resource area, threshold, mitigation.<br><b>In parallel &mdash; the answer key:</b> two AI coders independently label a 400-passage sample per track (FONSI and EIS) and a human adjudicator settles every disagreement</div>
  </div>
  <div style="text-align:center;color:#0047BB;font-size:1.5rem;line-height:1.1;">&#8595;</div>
  <div style="border:2px solid #0047BB;border-radius:8px;padding:.65rem 1rem;background:#f4f7fc;">
    <div style="font-weight:700;color:#012169;font-size:1.05rem;">4 · Validate</div>
    <div style="color:#1A1A2E;">Score the AI against the held-out human labels on each track (class agreement 0.81 FONSI, 0.69 EIS) before any result counts</div>
  </div>
  <div style="text-align:center;color:#0047BB;font-size:1.5rem;line-height:1.1;">&#8595;</div>
  <div style="border:2px solid #012169;border-radius:8px;padding:.65rem 1rem;background:#012169;">
    <div style="font-weight:700;color:#ffffff;font-size:1.05rem;">5 · Analyze and report</div>
    <div style="color:#ffffff;">Validated findings become this report and a source-linked data table &mdash; every number traces to its page</div>
  </div>
  <div style="text-align:center;color:#6b7280;font-size:.85rem;margin-top:.5rem;">
    The five steps of the analysis &mdash; each card matches a numbered step below.
  </div>
</div>
```

## What the FONSIs and EISs tell us

A significance call can land on either side of the line, and **the two document types record
opposite outcomes** — so we analyze them separately rather than pooling them:

- **FONSIs are the *below-the-line* record.** The agency concluded the impact is not
  significant. This is where the mitigated cases live — impacts that *would* have been
  significant without committed measures. FONSIs tell us how agencies stay under the line, and
  what they require to get there.
- **EISs are the *above-the-line* record.** These are impacts the agency judged significant
  (and, at the extreme, significant and *unavoidable*). They tell us which resources actually
  cross the line.

The diagram below is the vocabulary the rest of this report uses — every determination lands in
one of these outcomes, and "the line" is the significance threshold that separates them:

```{=html}
<div style="max-width:760px;margin:1.25rem auto;display:grid;grid-template-columns:1fr 44px 62px;grid-column-gap:8px;grid-row-gap:.6rem;align-items:stretch;">
  <div style="grid-column:1;grid-row:1;border:2px solid #75246C;border-radius:8px;padding:.65rem 1rem;background:#f3eaf2;">
    <div style="font-weight:700;color:#75246C;font-size:1.0rem;">Significant and unavoidable</div>
    <div style="color:#1A1A2E;">The extreme case: the impact is significant and <b>no available measure</b> can
    bring it below the threshold.</div>
  </div>
  <div style="grid-column:1;grid-row:2;border:2px solid #C22A90;border-radius:8px;padding:.65rem 1rem;background:#f9e9f4;">
    <div style="font-weight:700;color:#C22A90;font-size:1.0rem;">Significant adverse</div>
    <div style="color:#1A1A2E;">The impact crosses the significance threshold, so the full review applies.</div>
  </div>
  <div style="grid-column:1;grid-row:3;display:flex;align-items:center;gap:.75rem;margin:.3rem 0;">
    <div style="flex:1;border-top:3px solid #012169;"></div>
    <div style="font-weight:700;color:#012169;font-size:.9rem;letter-spacing:.05em;">THE SIGNIFICANCE LINE</div>
    <div style="flex:1;border-top:3px solid #012169;"></div>
  </div>
  <div style="grid-column:1;grid-row:4;border:2px solid #C22A90;border-radius:8px;padding:.65rem 1rem;background:#f4f7fc;">
    <div style="font-weight:700;color:#C22A90;font-size:1.0rem;">Less than significant <u>with committed mitigation</u></div>
    <div style="color:#1A1A2E;">The impact <b>would</b> cross the line, but specific required measures pull it
    under &mdash; the <b>mitigated FONSI</b>. These sit right at the line and are this report's central subject.</div>
  </div>
  <div style="grid-column:1;grid-row:5;border:2px solid #0047BB;border-radius:8px;padding:.65rem 1rem;background:#f4f7fc;">
    <div style="font-weight:700;color:#0047BB;font-size:1.0rem;">Less than significant</div>
    <div style="color:#1A1A2E;">A real but minor impact that stays under the line on its own, with no committed
    measures needed.</div>
  </div>
  <div style="grid-column:1;grid-row:6;border:2px solid #0047BB;border-radius:8px;padding:.65rem 1rem;background:#f4f7fc;">
    <div style="font-weight:700;color:#0047BB;font-size:1.0rem;">No significant impact</div>
    <div style="color:#1A1A2E;">The agency finds no meaningful impact on the resource at all &mdash; the
    below-the-line default.</div>
  </div>
  <div style="grid-column:2;grid-row:1/5;display:flex;align-items:center;">
    <div style="width:9px;align-self:stretch;border:2.5px solid #75246C;border-left:none;border-radius:0 8px 8px 0;"></div>
    <div style="font-weight:700;color:#75246C;font-size:.9rem;padding-left:5px;letter-spacing:.06em;">EIS</div>
  </div>
  <div style="grid-column:3;grid-row:4/7;display:flex;align-items:center;">
    <div style="width:9px;align-self:stretch;border:2.5px solid #0047BB;border-left:none;border-radius:0 8px 8px 0;"></div>
    <div style="font-weight:700;color:#0047BB;font-size:.9rem;padding-left:5px;letter-spacing:.06em;">FONSI</div>
  </div>
  <div style="grid-column:1/-1;grid-row:7;text-align:center;color:#6b7280;font-size:.85rem;">
    The brackets show where each outcome is recorded: above-the-line findings in an <b>EIS</b>, below-the-line
    in a <b>FONSI</b> &mdash; and a less-than-significant-with-committed-mitigation call appears in both
    (the mitigated FONSI below the line; its EIS analog is a significant impact judged reducible).
    Boxes closer to the line are closer calls.
  </div>
</div>
```

@fig-corpus-overview sets the scale: the decarbonization universe splits into a larger EIS corpus
and a smaller FONSI corpus (top), and both tracks touch every major resource area (bottom).

```{r}
#| label: fig-corpus-overview
#| fig-cap: "Top: the analysis scale, by review type — projects (left) and the documents actually parsed for findings (right); the EIS corpus is the larger one. Bottom: the mix of resource areas the determinations cover, for FONSIs (left) and EISs (right); each square ≈ 1% of that track's determinations. Both waffles share the resource legend."
include_fig("fig_corpus_overview.png")
```

## FONSI analysis {#sec-fonsi}

```{r}
#| label: fonsi-corpus
proj_rows <- str_detect(class_dist$determination_class, "project-level")
n_prim <- sum(class_dist$n_determinations[!proj_rows])
n_docs <- mit_doc$n_documents[1]
n_proj <- mit_doc$n_projects[1]
```

The decarbonization FONSI corpus contains 452 projects. This analysis rests on the
`r fmt_n(n_proj)` of them that are (a) led by BLM or the DOE family and (b) carry
machine-extractable significance findings — spanning `r fmt_n(n_docs)` decision documents and
`r fmt_n(n_prim)` distinct significance determinations (one per document × resource area ×
outcome).

```{r}
#| label: fig-agency-scope
#| fig-cap: "Lead agency for the 452 decarbonization FONSIs. BLM and the DOE family (94.5%) carry the headline analysis; the remaining 5.5% — led by other federal agencies with only partial coverage — are set aside as context."
include_fig("fig_agency_scope.png")
```

Three filters take the 452 down to the `r fmt_n(n_proj)` analyzed here, and @fig-fonsi-funnel
traces each step:

1. **Agency scope (452 → 427).** The 25 FONSIs led by agencies other than BLM and the DOE family
   are set aside as context (@fig-agency-scope) — only BLM + DOE have complete coverage in the
   dataset.
2. **A dated, 2009-or-later decision (427 → 261) — the largest cut.** NEPA's ground rules changed
   repeatedly over this period, most recently with the 2023 Fiscal Responsibility Act, and
   comparing agency behavior across those regimes is one of this deliverable's core questions
   — so every rate is read against the regulatory era the decision was made under. A FONSI with no
   extractable decision date (105 projects) cannot be assigned an era, pre-2009 decisions (55)
   predate the consistently documented modern corpus, and 6 sit on a regime boundary.
3. **Extraction coverage (261 → 193).** The upstream pipeline locates the finding sections in 224
   of the 261 — the other 37 documents were parsed but phrase their finding in wording the
   extraction does not recognize — and in 31 more, the flagged text turns out to contain no
   codable determination.

The first two are scope choices; the third is a **coverage limitation, not a different finding**.
The `r fmt_n(n_proj)` analyzed projects are representative of the corpus, but the rates below
describe *those* projects, not all 452 (see *Methods*).

```{r}
#| label: fig-fonsi-funnel
#| fig-cap: "How the 452-project FONSI corpus narrows to the 193 analyzed: agency scope, then the dated 2009–present window, then extraction coverage. Darker bars are closer to the analyzed set; every rate in this report describes the darkest bar."
include_fig("fig_fonsi_funnel.png")
```

::: {.callout-tip}
## Key takeaways (FONSI track)
- **Most analyzed decarbonization FONSIs are *mitigated*.** About `r fmt_pct(mit_doc$share_mitigated_class_signal)`
  of the 258 analyzed FONSI decision documents — the BLM + DOE, dated, machine-extractable set
  traced in @fig-fonsi-funnel, not all 452 in the corpus — reach "no significant impact" only
  because the developer commits to specific measures; the impact *would* have been significant
  otherwise.
- **A handful of resources carry the mitigation.** Biological resources, soils/geology, water, and
  cultural/historic sites are where projects most often sit at the line and are pulled under it —
  biological alone is mitigation-dependent in ~30% of its determinations.
- **The below-the-line default is "no significant impact."** Across resource areas, outright
  no-significant-impact is the most common outcome; genuinely *significant* findings live in the EIS
  record, not here.
- **The numbers are validated.** The extraction agrees with an AI-human reviewed answer key at
  0.81 on the significance class (held-out test) before any figure in this report was produced.
:::

### How accurate is the extraction?

Every determination is machine-extracted, so before any number is reported the pipeline is scored
against an **AI-human reviewed answer key**^[The answer key was built by two independent AI coding
agents (Claude and Codex), each labeling every passage in the sample on its own; a human analyst
then audited the two sets — values the coders agreed on were accepted, and every disagreement
(over half the rows, on at least one field) was adjudicated by hand. "AI-human reviewed" is
shorthand for that process: AI-coded, human-audited.]. The sample is 400 windows — a "window"
is one flagged passage the extraction reads — drawn **stratified**, meaning it deliberately covers
every outcome type and the agency mix rather than being a simple random draw, plus negative
passages to catch false positives. 30% of the windows were held out as an untouched test; the
held-out column is the honest score.

```{r}
#| label: tbl-validation
#| tbl-cap: "Extraction accuracy versus the AI-human reviewed gold set, full sample and held-out test."
metric_lab <- c(candidate_is_determination        = "Finds a determination (per passage)",
                resource_determination_detection  = "Assigns the right resource",
                determination_class_macro_f1       = "Gets the significance class right",
                mitigation_dependent_f1            = "Flags mitigation-dependence",
                primary_threshold_type_accuracy    = "Identifies the regulatory threshold")
val |>
  mutate(score = coalesce(f1, precision)) |>
  filter(scope %in% c("overall", "holdout"), metric %in% names(metric_lab)) |>
  select(metric, scope, score) |>
  pivot_wider(names_from = scope, values_from = score) |>
  mutate(metric = factor(metric, levels = names(metric_lab), labels = unname(metric_lab))) |>
  arrange(metric) |>
  gt() |>
  cols_label(metric = "What the pipeline is scored on", overall = "All 400", holdout = "Held-out test") |>
  fmt_number(columns = c(overall, holdout), decimals = 2) |>
  data_color(columns = holdout, colors = col_numeric(c("#fdecea", "#eaf3ea"), domain = c(0.4, 1))) |>
  cols_align(align = "left", columns = metric) |>
  tab_style(style = list(cell_fill(color = NAVY), cell_text(color = "white", weight = "bold")),
            locations = cells_column_labels()) |>
  tab_options(table.font.size = px(13), table.width = pct(100), table.align = "center") |>
  tab_source_note("Scores are F1 (0–1), except the threshold row, which is accuracy. Held-out = the 30% of the gold set the pipeline never influenced.")
```

In plain terms: the extraction **almost never misses a determination** and places it in the
right resource area ~89% of the time; it agrees with the answer key on the significance
*class* at 0.81 — the standard bar for this kind of task. The one weaker signal is
*mitigation-dependence*, discussed at the end of this section.

```{r}
#| label: fig-validation
#| fig-cap: "Agreement with the AI-human reviewed answer key on the stratified 400-window sample (one window = one flagged passage; sampled to cover every outcome type and the agency mix): full sample in magenta versus the untouched 30% held-out test in blue. The dashed line marks 0.80, the standard bar for this kind of task."
include_fig("fig_validation_accuracy.png")
```

Why the last two scores are lower — and why it doesn't undercut the findings. Both are
*secondary attributes*, not the core call, and both are genuinely harder:

- *Mitigation-dependence (0.62).* Deciding whether a conclusion depends on committed mitigation
  is one of the hardest judgments in the whole task — the two independent AI coders themselves agreed only
  ~68% of the time on the fine line between "less than significant" and "less than significant *with
  mitigation*." On top of that, the pipeline's structured mitigation flag is tagged at the document
  level, so it over-attributes across a multi-resource passage, and its precision is ultimately
  bounded by upstream tagging. We handle this by reporting mitigation at the document level (where
  it is robust) and using the precise per-resource signal for the resource-level breakdown — so
  the numbers you see below are the high-precision measure, not the noisy flag.
- *Regulatory threshold (0.66–0.71).* This is the most granular field — pinning a conclusion to a
  *specific* rule (a NAAQS limit, an ESA take threshold, a §106 adverse-effect finding). Many
  conclusions are not anchored to a single numeric threshold at all, and the distinctions are fine.
  It is reported descriptively as supporting context, never as a headline number.

The three scores that carry the findings — finding the determination, placing the resource, and
getting the significance class — all clear or approach the standard bar.

### How agencies stay below the line

Within FONSIs, the determinations sit — as they must — below the significance line: the agency
either finds no significant impact outright or an impact that is less than significant,
sometimes only with committed mitigation. @fig-outcomes-by-resource breaks this down by
resource area — each resource's determinations split across the three outcomes, sorted so the
resources most reliant on mitigation sit at the top, with the count and share behind each bar.

```{r}
#| label: fig-outcomes-by-resource
#| fig-cap: "Share of each resource's FONSI determinations by outcome, sorted by reliance on mitigation. Each segment shows its count and percentage; n is the resource's total determinations. Resources with more magenta lean harder on committed mitigation to reach a not-significant finding."
include_fig("fig_outcomes_by_resource.png")
```

The dominant outcome is no significant impact, as expected for FONSIs. But the resources
agencies analyze most — biological, water, cultural/historic, and public health — are also
where the close calls concentrate: the magenta band (*committed mitigation*) is
widest for biological, soils/geology, and water. That is what makes the mitigated cases the real
story.

### Regulatory thresholds behind the calls

When a conclusion is anchored to a *specific* regulatory yardstick, which one? @fig-threshold-profile
profiles the thresholds the FONSI conclusions cite — dominated by quantitative numeric limits
(resource-specific caps and standards), wetland/floodplain extent, NHPA §106 adverse-effect findings for
cultural resources, visual (VRM) classes, and the Endangered Species Act (take and
jeopardy). This mirrors the resource pattern above: the same areas that drive mitigation are the
ones with the hardest regulatory lines to clear.

```{r}
#| label: fig-threshold-profile
#| fig-cap: "Regulatory thresholds cited across primary FONSI determinations. Reported descriptively: identifying the exact threshold is the pipeline's least-accurate field, and many conclusions are not anchored to a specific threshold at all."
include_fig("fig_threshold_profile.png")
```

::: {.callout-note appearance="minimal"}
Threshold identification is the extraction's weakest field (see the validation above), so treat
this profile as an indicative picture of *which rules recur*, not an exact census.
:::

### Mitigated FONSIs — the line itself

The most revealing determinations are the mitigated ones: impacts the agency says *would* be
significant, brought below the line only by required measures. These show exactly where the line
sits and what it costs to stay under it — and they are the *majority* of FONSIs (@fig-mitigated-share).

```{r}
#| label: fig-mitigated-share
#| fig-cap: "Share of FONSI documents that reach a not-significant finding only with committed mitigation (per-resource class signal)."
include_fig("fig_mitigated_share.png")
```

```{r}
#| label: mit-headline
mit_rate <- mit_doc$share_mitigated_class_signal[1]
top_mit  <- mit_res |> filter(n_determinations >= 20) |> arrange(desc(share_mit_class)) |>
  slice_head(n = 4) |> mutate(Resource = relabel(shared_resource_area, res_label))
top_str  <- paste0(top_mit$Resource, " (", percent(top_mit$share_mit_class, accuracy = 1), ")",
                   collapse = ", ")
```

About `r fmt_pct(mit_rate)` of the analyzed FONSI documents are mitigated — at least one
resource whose not-significant conclusion depends explicitly on committed mitigation. The
denominator is **documents, not projects**: the analyzed set is 258 decision documents across 193
projects (some projects file more than one FONSI), so the figure's counts (149 mitigated + 109
not) sum to 258.

The mitigation burden is concentrated in **`r top_str`** — the resources most often needing
committed measures to keep a project below the significance line. That pattern is the deliverable's
core finding: biological, soils/geology, water, and cultural resources are where decarbonization
projects most frequently sit *at* the line and are pulled under it by mitigation.

```{r}
#| label: fig-mitigation-by-resource
#| fig-cap: "Share of a resource's FONSI conclusions that depend on committed mitigation (per-resource class signal). Point size scales with how often the resource is analyzed."
include_fig("fig_mitigation_by_resource.png")
```

Reading volume and mitigation-intensity together sharpens the picture. Resources in the upper
right are both analyzed often *and* mitigated often — the areas that most shape whether a
decarbonization project can take the faster FONSI path.

```{r}
#| label: fig-mitigation-landscape
#| fig-cap: "Each resource by how often it is analyzed (x) and how often its conclusion depends on mitigation (y). Biological resources dominate both; soils/geology are mitigated intensively despite lower volume. Dashed line = overall average mitigation share."
include_fig("fig_mitigation_landscape.png")
```

What do these mitigations actually look like across the board? Two verbatim examples per resource,
ordered from the most mitigation-heavy resources down — seasonal timing and species avoidance for
biology, wetland minimization for water, cultural-resource protection measures, dust control for
air, and so on:

```{r}
#| label: tbl-mitigation-examples
#| tbl-cap: "Representative mitigation measures behind less-than-significant-with-mitigation conclusions, by resource area."
if (nrow(mit_ex) > 0) {
  mit_ex |>
    gt(groupname_col = "Resource") |>
    cols_label(example = "Example mitigation drawn from the FONSI") |>
    cols_align(align = "left") |>
    cols_width(example ~ pct(100)) |>
    tab_style(style = list(cell_fill(color = NAVY), cell_text(color = "white", weight = "bold")),
              locations = cells_row_groups()) |>
    tab_style(style = cell_text(color = NAVY, weight = "bold"), locations = cells_column_labels()) |>
    tab_options(table.font.size = px(12.5), table.width = pct(100), table.align = "center",
                row_group.font.weight = "bold", data_row.padding = px(6)) |>
    tab_source_note("Verbatim from the model's per-resource rationale on determinations that depend on committed mitigation. Climate/GHG had no such conclusions in the FONSI corpus.")
}
```

Mitigated FONSIs are also broader analyses: a project that needs committed mitigation for any
resource tends to address more resource areas overall — a median of roughly six versus four for
non-mitigated FONSIs.

```{r}
#| label: fig-breadth
#| fig-cap: "Distinct resource areas addressed per FONSI, split by whether the document relies on committed mitigation. Boxes show the median and interquartile range; the shaded shape is the full distribution."
include_fig("fig_breadth_per_fonsi.png")
```

::: {.callout-warning appearance="minimal"}
**Read mitigation at the right grain.** The *document-level* "is this a mitigated FONSI?" rate is
robust. The *resource-level* attribution above uses the high-precision per-resource signal; a
broader structured signal exists but over-attributes across resources, and its accuracy is
ultimately bounded by upstream tagging (validation F1 ≈ 0.62). Treat the resource shares as a
well-grounded floor, not an exact census.
:::

### Does it differ by agency?

The two departments run different kinds of projects — BLM tends toward land-based sites, the DOE
family toward energy facilities and transmission — so the resources that push them to mitigate
differ too. @fig-dept-by-resource compares, for each resource, how often a conclusion depends on
mitigation for BLM versus DOE.

```{r}
#| label: fig-dept-by-resource
#| fig-cap: "Share of each resource's FONSI conclusions that depend on committed mitigation, BLM versus the DOE family. A wider gap means the resource is a bigger mitigation driver for one department than the other."
include_fig("fig_dept_by_resource.png")
```

The pattern is intuitive: biological, cultural/historic, air quality, and visual resources
drive mitigation harder for **BLM** (land-facing impacts on habitat, historic sites, and
viewsheds), while noise and water lean toward the **DOE family** (facility- and
construction-driven effects). Within DOE, the picture varies further by sub-agency
(@fig-subagency-by-resource) — though the smaller sub-agency cells are noisy and should be read for
pattern, not precision.

```{r}
#| label: fig-subagency-by-resource
#| fig-cap: "Mitigation-dependent share by sub-agency and resource area, for sub-agencies with at least 40 determinations. Darker = a larger share of that resource's conclusions depend on mitigation. Small cells are noisy."
include_fig("fig_subagency_by_resource.png")
```

### Significance by clean-energy technology

The kind of project matters as much as the agency. Classifying each project by its primary
technology (from the dataset's project-type field) lets us ask which technologies most often need
mitigation to stay below the line (@fig-fonsi-technology).

```{r}
#| label: fig-fonsi-technology
#| fig-cap: "Share of a technology's FONSI conclusions that depend on committed mitigation. Technology is derived from the dataset's project-type classification; about 1 in 5 projects resolve to an 'Other / mixed' bucket and are excluded, so read this for the named technologies."
include_fig("fig_fonsi_technology.png")
```

**Transmission and wind lean hardest on mitigation** — linear corridors and turbine sites touch
habitat, viewsheds, and cultural resources that repeatedly need committed measures to reach a
not-significant finding. Solar and storage sit lower, closer to the corpus average.

### Is the committed mitigation enforceable?

"Mitigated" only means something if the measures are actually required. The extraction flags whether
each mitigated conclusion is tied to an **enforceable permit condition** or is committed only in the
document's text. @fig-fonsi-enforceability shows the split by resource.

```{r}
#| label: fig-fonsi-enforceability
#| fig-cap: "Share of a resource's mitigated FONSI conclusions tied to an enforceable permit condition (versus committed only in the document text). Reported descriptively — an unmatched conclusion may still be enforced through channels the extraction doesn't capture."
include_fig("fig_fonsi_enforceability.png")
```

The pattern tracks regulatory leverage. Biological, air-quality, and water conclusions — backed
by the Endangered Species Act and the Clean Air and Water Acts — most often carry an enforceable
permit condition. Visual, cultural, and soils measures are far less often tied to one; they
lean on commitments in the document text. So for the softer resources, the mitigation behind a
not-significant finding is the least demonstrably binding — a caveat worth carrying into how much
weight the "mitigated FONSI" label should bear for those areas.

## EIS analysis {#sec-eis}

The FONSI track shows how agencies stay below the significance line. The EIS track is its mirror —
the above-the-line record of where impacts actually cross it. It runs the identical
extraction-and-validation machinery on the EIS corpus (same controlled vocabulary, per-resource
grain, and AI-human reviewed held-out test) and adds three EIS-only fields the FONSI side doesn't need:
which alternative a finding attaches to, why the impact is significant, and how it reaches the
resource.

```{r}
#| label: eis-scope
eis_sig_classes <- c("significant_adverse", "significant_unavoidable", "eis_required")
eis_n_det   <- sum(eis_class$n_determinations)
eis_n_proj  <- eis_mit$n_projects[1]
eis_n_docs  <- eis_mit$n_documents[1]
eis_n_sig   <- sum(eis_class$n_determinations[eis_class$determination_class %in% eis_sig_classes])
eis_n_unav  <- eis_class$n_determinations[eis_class$determination_class == "significant_unavoidable"]
eis_class_label <- c(class_label, eis_required = "EIS required")
efun <- function(k) { v <- eis_funnel$n[eis_funnel$metric == k]; if (length(v)) v[1] else NA }
```

The decarbonization EIS corpus is the larger of the two — 753 projects. Not all carry usable
findings: significance sections were retrieved for **`r fmt_n(efun("projects_with_sections"))`**, of
which **`r fmt_n(efun("projects_analyzed"))`** yielded at least one determination (@fig-eis-funnel).
Unlike the FONSI track, the EIS analysis covers **all agencies**, not just BLM + DOE: it is purely
descriptive — which resources cross the line — and never uses a project's decision date, so projects
without a firm date are kept rather than dropped. On that basis it rests on
`r fmt_n(eis_n_proj)` projects (`r fmt_n(eis_n_docs)` documents) carrying `r fmt_n(eis_n_det)`
distinct significance determinations — of which `r fmt_n(eis_n_sig)` land above the line, including
`r fmt_n(eis_n_unav)` judged significant and unavoidable, impacts mitigation cannot erase.

```{r}
#| label: fig-eis-funnel
#| fig-cap: "How the 753-project EIS corpus narrows to the analyzed set (top left): 536 had significance sections retrieved, 506 yielded a determination. Those projects produced 13,240 determinations, of which 2,198 are significant (cross the line) and 698 significant-unavoidable (top right). Of the 506 analyzed projects, only 239 have a firm in-window decision date — the analysis keeps the undated ones because it never uses the date (bottom)."
include_fig("fig_eis_funnel.png")
```

::: {.callout-tip}
## Key takeaways (EIS track)
- **Visual impacts cross the line most.** Viewshed conclusions are judged significant more often than
  any other resource (~33% of visual determinations), and they most often reach *significant and
  unavoidable* — a changed landscape usually cannot be mitigated back.
- **Resources split into two shapes.** Some are managed below the line (soils/geology, water, and
  public health are routinely mitigated to "not significant" and rarely cross); others cross over
  (visual, land use, and air quality break through more often than they are mitigated below it).
- Magnitude, cumulative effects, and protected resources are the dominant reasons a call is
  significant — and roughly a third of significant findings rest on cumulative pathways, not the
  project's direct footprint alone.
:::

### How accurate is the EIS extraction?

The EIS pipeline was graded exactly as the FONSI one: two reviewers coded a stratified
400-window sample, a third adjudicated, and 30% was held out as an untouched test. The scores
are lower across the board — the EIS task is harder, and the report treats it accordingly.

```{r}
#| label: tbl-validation-eis
#| tbl-cap: "EIS extraction accuracy versus the AI-human reviewed gold set, full sample and held-out test. Lower than the FONSI track — the above-the-line distinctions are genuinely finer."
metric_lab <- c(candidate_is_determination        = "Finds a determination (per passage)",
                resource_determination_detection  = "Assigns the right resource",
                determination_class_macro_f1       = "Gets the significance class right",
                mitigation_dependent_f1            = "Flags mitigation-dependence",
                primary_threshold_type_accuracy    = "Identifies the regulatory threshold")
eis_val |>
  mutate(score = coalesce(f1, precision)) |>
  filter(scope %in% c("overall", "holdout"), metric %in% names(metric_lab)) |>
  select(metric, scope, score) |>
  pivot_wider(names_from = scope, values_from = score) |>
  mutate(metric = factor(metric, levels = names(metric_lab), labels = unname(metric_lab))) |>
  arrange(metric) |>
  gt() |>
  cols_label(metric = "What the pipeline is scored on", overall = "All 400", holdout = "Held-out test") |>
  fmt_number(columns = c(overall, holdout), decimals = 2) |>
  data_color(columns = holdout, colors = col_numeric(c("#fdecea", "#eaf3ea"), domain = c(0.4, 1))) |>
  cols_align(align = "left", columns = metric) |>
  tab_style(style = list(cell_fill(color = NAVY), cell_text(color = "white", weight = "bold")),
            locations = cells_column_labels()) |>
  tab_options(table.font.size = px(13), table.width = pct(100), table.align = "center") |>
  tab_source_note("Scores are F1 (0–1), except the threshold row, which is accuracy. Held-out = the 30% of the gold set the pipeline never influenced.")
```

```{r}
#| label: fig-validation-eis
#| fig-cap: "EIS agreement with the AI-human reviewed answer key: full sample (magenta) vs held-out test (blue). The shaded bottom two rows are secondary attributes that matter less to the findings (as in the FONSI figure). Every score sits below the FONSI equivalent — the above-the-line call is a finer judgment (the two independent AI coders themselves agreed only ~58% on the class)."
include_fig("fig_validation_accuracy_eis.png")
```

The pipeline still **finds the determination reliably** (0.84) and places it in the right resource
about two-thirds of the time; the significance *class* lands at 0.69. That last number is
below the FONSI bar for a real reason: separating "significant" from "significant *and
unavoidable*" is a fine distinction the AI coders split on too. We therefore lead with the
robust cut — *which resources cross the line* — and treat the adverse-vs-unavoidable split as
directional.

### Which resources cross the line

Within EISs, the interesting determinations are the ones that break **above** the line. Those come in
two degrees of severity:

- **Significant adverse** — the impact is significant, but the agency does not declare it beyond
  repair; committed measures may still reduce it.
- **Significant unavoidable** — the impact is significant *and* cannot be brought below the line even
  with mitigation. This is the more severe call, and the one that most constrains a project.

@fig-eis-above-line ranks each resource by how often its EIS conclusions are judged significant,
splitting the two degrees (least severe on the left, most severe on the right).

```{r}
#| label: fig-eis-above-line
#| fig-cap: "Share of each resource's EIS determinations judged significant, sorted (top = most likely to cross the line). Each percentage (and count) is of the resource's total EIS determinations: the in-bar labels split significant-adverse (lighter, left) from significant-unavoidable (darker, right); the end label gives the combined significant share and count (e.g. visual = 316 of 959, of which 126 are unavoidable)."
include_fig("fig_eis_above_line.png")
```

```{r}
#| label: eis-cross
eis_top <- eis_res |> filter(!n_suppressed) |> arrange(desc(share_sig)) |> slice_head(n = 4) |>
  mutate(R = relabel(shared_resource_area, res_label),
         lab = paste0(R, " (", percent(share_sig, accuracy = 1), ")"))
```

`r eis_top$lab[1]`, `r eis_top$lab[2]`, `r eis_top$lab[3]`, and `r eis_top$lab[4]` are the resources
most likely to cross into significance. **Visual stands alone at the top** — the clearest case of an
impact that, once significant, tends to stay that way.

### The wall: significant *and* unavoidable

The starkest determinations are the significant and unavoidable ones — where the agency concludes
the impact is real, material, and cannot be mitigated below the line. These are the true limits
on a decarbonization project. @fig-eis-unavoidable counts them by resource — each bar is the number
of distinct *significant-and-unavoidable* determinations (one per document × resource) for that
resource, across the analyzed EIS projects.

```{r}
#| label: fig-eis-unavoidable
#| fig-cap: "Count of significant-unavoidable determinations by resource — the darker (right-hand) segments of the previous figure, now counted. Visual's 126 means 126 of its 316 significant determinations were judged unavoidable; biological, cultural, and air quality follow. These are the impacts that remain significant even after mitigation."
include_fig("fig_eis_unavoidable.png")
```

**Visual impacts dominate the wall.** A wind farm on a ridgeline or a transmission corridor across
open country changes a landscape in ways no mitigation fully undoes — so viewshed conclusions reach
*unavoidable* more than any other resource. Biological, cultural, and air-quality impacts follow,
where habitat loss, historic-setting change, or a regional pollutant burden can exceed what on-site
measures can offset. @tbl-eis-unavoidable-ex shows what these look like in the agencies' own words.

```{r}
#| label: tbl-eis-unavoidable-ex
#| tbl-cap: "Representative significant-and-unavoidable determinations, verbatim from the EIS, ordered from the resources that hit the wall most often."
if (nrow(eis_unav_ex) > 0) {
  eis_unav_ex |>
    gt(groupname_col = "Resource") |>
    cols_label(example = "Significant-unavoidable finding, in the agency's words") |>
    cols_align(align = "left") |>
    cols_width(example ~ pct(100)) |>
    tab_style(style = list(cell_fill(color = "#75246C"), cell_text(color = "white", weight = "bold")),
              locations = cells_row_groups()) |>
    tab_style(style = cell_text(color = "#75246C", weight = "bold"), locations = cells_column_labels()) |>
    tab_options(table.font.size = px(12.5), table.width = pct(100), table.align = "center",
                row_group.font.weight = "bold", data_row.padding = px(6)) |>
    tab_source_note("Verbatim from the model's per-determination rationale on significant-unavoidable conclusions.")
}
```

### How broadly does an EIS cross the line?

The resource views above pool all EISs together. Stepping back to the document level asks a
different question: when an EIS *does* cross the line, does it do so on one resource or many?
@fig-eis-breadth answers it.

```{r}
#| label: fig-eis-breadth
#| fig-cap: "Distinct resource areas judged significant per EIS document, among the 646 documents with at least one significant finding. Most cross narrowly — 234 on a single resource — but a long tail crosses on three or more; a handful span the whole environmental slate."
include_fig("fig_eis_breadth.png")
```

Crossing the line is usually a focused event: over a third of EISs that cross do so on a single
resource, and the median is two. But a meaningful minority — 284 documents — cross on three or
more resources, and a few span nearly the entire environmental slate. (This counts only resource-area
determinations; documents whose significant conclusion is stated project-wide, or not captured by the
retrieval, sit outside it — see *Methods*.)

### FONSI vs EIS: two ways a resource can be a problem

Analyzing the two tracks separately pays off in one picture (@fig-fonsi-vs-eis), which plots each
resource by how often it is mitigated below the line (FONSIs, x-axis) against how often it crosses
the line (EISs, y-axis). **The magenta resources are more likely to be judged significant than to be
mitigated; the blue resources are the reverse.** (Both tracks are restricted to BLM + DOE here, so
the two are measured on the same footing — FONSI coverage is only partial outside those agencies.)

```{r}
#| label: fig-fonsi-vs-eis
#| fig-cap: "Each resource (BLM + DOE, so both tracks share a scope) plotted by how often it is mitigated below the line in FONSIs (x) versus how often it crosses into significance in EISs (y). The dashed diagonal is equal odds; the bottom legend gives the colors — magenta resources (above the line) are judged significant more often than mitigated, blue resources (below) the reverse. Point size ≈ determinations."
include_fig("fig_fonsi_vs_eis.png")
```

The split is the deliverable's sharpest structural finding:

- **More likely to be significant** (magenta) — visual, land use, air quality — break above the line
  more often than they are mitigated below it. When these become a problem, agencies more often
  accept a significant finding than engineer their way under it; visual is the extreme (within
  BLM + DOE it crosses ~38% of the time and is mitigated only ~11%).
- **More likely to be mitigated** (blue) — soils/geology, water, public health — show the opposite:
  frequently mitigated in FONSIs, rarely crossing into significance. These are the impacts the
  toolbox of committed measures reliably handles.
- **Cultural / historic sits right on the line** — it crosses into significance about as often as it
  is mitigated below it (~23% vs ~21%), the one resource that genuinely straddles both sides.

### What makes an impact significant

For every above-the-line determination the extraction also records why it is significant.
@fig-eis-factors gives the dominant factors overall; @fig-eis-heatmap is the fuller picture — *where
each factor bites*, i.e. the resources each reason concentrates in.

```{r}
#| label: fig-eis-factors
#| fig-cap: "The dominant factor behind each significant determination. Sheer magnitude, a protected resource, and cumulative effects lead."
include_fig("fig_eis_factors.png")
```

```{r}
#| label: fig-eis-heatmap
#| fig-cap: "Where each factor bites. Each row is shaded by the share within that factor (not across the whole grid), so the darkest cells show where a reason concentrates; the cell number is the raw count. Sheer magnitude concentrates in visual, water, and land use; a protected resource in cultural and biological; a regulatory threshold in air quality and noise; cumulative effects in biological and visual."
include_fig("fig_eis_factor_heatmap.png")
```

```{r}
#| label: eis-factors
eis_fac_top <- eis_fac |> filter(!n_suppressed) |> arrange(desc(n)) |> slice_head(n = 3) |>
  pull(significance_factor)
fac_pretty <- c(magnitude = "sheer magnitude", cumulative = "cumulative effects",
                protected_resource = "a protected resource", regulatory_threshold = "a regulatory threshold")
```

The leading reasons an impact is called significant are `r fac_pretty[eis_fac_top[1]]`,
`r fac_pretty[eis_fac_top[2]]`, and `r fac_pretty[eis_fac_top[3]]`, and the heatmap shows they are
not interchangeable: a protected resource drives significance mainly for cultural and biological
resources (endangered species, historic properties), a regulatory threshold for air quality
and noise (numeric standards), and cumulative effects for biological, visual, and water. That
last one is notable for decarbonization siting — about one in four significant findings rests on
impacts that build up across a landscape or airshed rather than on one project's direct footprint.
@tbl-eis-factor-ex gives verbatim examples of each.

```{r}
#| label: tbl-eis-factor-ex
#| tbl-cap: "What each significance factor looks like in the EIS text — verbatim examples, ordered from the most common factor down."
if (nrow(eis_fac_ex) > 0) {
  eis_fac_ex |>
    gt(groupname_col = "Factor") |>
    cols_label(example = "Significant finding, in the agency's words") |>
    cols_align(align = "left") |>
    cols_width(example ~ pct(100)) |>
    tab_style(style = list(cell_fill(color = "#75246C"), cell_text(color = "white", weight = "bold")),
              locations = cells_row_groups()) |>
    tab_style(style = cell_text(color = "#75246C", weight = "bold"), locations = cells_column_labels()) |>
    tab_options(table.font.size = px(12.5), table.width = pct(100), table.align = "center",
                row_group.font.weight = "bold", data_row.padding = px(6)) |>
    tab_source_note("Verbatim from the model's per-determination rationale, grouped by the dominant significance factor it recorded.")
}
```

::: {.callout-note appearance="minimal"}
**Findings attach mostly to the proposed action.** About half of significant determinations name a
specific alternative; where they do, the proposed action dominates, with named build and
no-action alternatives carrying the rest. Significance is overwhelmingly a judgment about *what the
applicant proposed*, not about the alternatives weighed against it.
:::

### Does it differ by agency?

Because the EIS track spans every lead agency (not just BLM + DOE), we can ask whether some agencies
cross the line more than others. They do — and the pattern tracks the kind of project each agency
runs (@fig-eis-by-agency).

```{r}
#| label: fig-eis-by-agency
#| fig-cap: "Share of a lead agency's EIS determinations judged significant, for agencies with at least 150 determinations. Dark blue = BLM and the DOE family (the two agencies with complete dataset coverage); light blue = other agencies, shown for context with partial coverage. Point size ≈ number of determinations."
include_fig("fig_eis_by_agency.png")
```

```{r}
#| label: eis-agency
ea_ord <- eis_agency |> arrange(desc(share))
ea_top <- ea_ord |> slice_head(n = 2); ea_bot <- ea_ord |> slice_tail(n = 2)
```

The spread is real — from `r ea_top$Agency[1]` (`r percent(ea_top$share[1], accuracy=1)`) at the
top down to `r ea_bot$Agency[2]` (`r percent(ea_bot$share[2], accuracy=1)`). Land- and
water-facing agencies — the Army Corps and BLM — cross into significance most often, which
fits their portfolios (dredge-and-fill permits, public-land rights-of-way with visual and habitat
exposure). The utility-service programs — the Rural Utilities Service and TVA — cross least,
consistent with routine utility infrastructure on already-developed land and corridors. BLM and the DOE
family remain the only agencies with complete coverage; the others are shown for context.

### Significance by clean-energy technology

The most decision-relevant cut for a decarbonization portfolio is the technology. Classifying each
project by its primary technology (from the dataset's project-type field), @fig-eis-technology ranks
how often each crosses the line, and @fig-eis-technology-resource shows *where* — each technology's
distinct "significance signature."

```{r}
#| label: fig-eis-technology
#| fig-cap: "Share of a technology's EIS determinations judged significant. Technology is derived from the dataset's project-type classification; about 1 in 5 projects resolve to an 'Other / mixed' bucket and are excluded, so read this for the named technologies."
include_fig("fig_eis_technology.png")
```

```{r}
#| label: eis-tech
et <- eis_tech |> arrange(desc(share))
et_top <- et |> slice_head(n = 1)
nuc <- et |> filter(tech == "Nuclear")
```

**Hydropower crosses the line most** (`r percent(et_top$share, accuracy=1)`) — dams carry
aquatic and water impacts that are hard to avoid — followed by transmission and solar (~19% and ~18%).
**Nuclear crosses least among the large technologies** (`r percent(nuc$share, accuracy=1)`), consistent
with siting on already-industrial land under highly prescribed review. But the *rate* is only half the
story; the resources each technology hits are distinct:

```{r}
#| label: fig-eis-technology-resource
#| fig-cap: "Where each technology crosses the line, row-normalized so each technology's own pattern stands out (cell = count of significant determinations). Wind's signature is biological (bird and bat collision) and noise; solar's is visual, cultural, and air quality; transmission spreads across biological, visual, and cultural; nuclear and hydro concentrate on biological and water."
include_fig("fig_eis_technology_resource.png")
```

The signatures are intuitive and actionable: **wind** is defined by biological impacts (avian and
bat collision) plus noise; **solar** by visual, cultural, and air-quality (desert dust);
**transmission** spreads its crossings across biological, visual, and cultural along its corridor;
and **hydro and nuclear** concentrate on biological and water. For siting decisions, this says the
environmental fight is largely predictable from the technology: propose utility-scale solar and expect
viewshed and cultural-resource conflicts; propose wind and expect the fight to be about wildlife.

### Significant, but reducible

Not every above-the-line finding is a hard stop. One of the significance factors the extraction
records is **mitigable** — the impact crosses the line, but the agency notes that measures can still
lessen it. This is the EIS analog of a mitigated FONSI, and @fig-eis-mitigable shows where it occurs.

```{r}
#| label: fig-eis-mitigable
#| fig-cap: "Significant EIS determinations the agency flags as still mitigable, by resource — the above-the-line counterpart to a mitigated FONSI."
include_fig("fig_eis_mitigable.png")
```

```{r}
#| label: tbl-eis-mitigable-ex
#| tbl-cap: "Representative significant-but-mitigable determinations, verbatim from the EIS."
if (nrow(eis_mit_ex) > 0) {
  eis_mit_ex |> slice_head(n = 8) |>
    gt(groupname_col = "Resource") |>
    cols_label(example = "Significant-but-mitigable finding, in the agency's words") |>
    cols_align(align = "left") |> cols_width(example ~ pct(100)) |>
    tab_style(style = list(cell_fill(color = "#75246C"), cell_text(color = "white", weight = "bold")),
              locations = cells_row_groups()) |>
    tab_style(style = cell_text(color = "#75246C", weight = "bold"), locations = cells_column_labels()) |>
    tab_options(table.font.size = px(12.5), table.width = pct(100), table.align = "center",
                row_group.font.weight = "bold", data_row.padding = px(6)) |>
    tab_source_note("Verbatim from the model's rationale on determinations tagged with the 'mitigable' significance factor.")
}
```

Tellingly, visual leads this list too: even the resource most likely to hit the wall sometimes
admits partial mitigation (screening, siting adjustments), just not enough to drop below the line.
These cases are the closest the EIS record comes to the mitigated-FONSI story — significant enough to
require a full statement, but not beyond the reach of committed measures.

## Reproducibility

Every number and figure in this report is regenerated by committed code, run from the repository
root in the `nepa` conda environment. The two significance-extraction passes (FONSI and EIS) are
**billable LLM prerequisites** — run once, then the deterministic analysis reads their outputs.

``` bash
# 0. framework regime + corpus + cohorts (deterministic)
conda run -n nepa python phase2/code/deliverable02/00_resolve_framework_regime.py
conda run -n nepa python phase2/code/deliverable02/01_build_d2_inventory.py

# 1. FONSI extraction — billable Anthropic batch (Sonnet 5); --dry-run for cost only
conda run -n nepa python phase2/code/deliverable02/02_extract_fonsi_significance.py \
  --batch-run --model claude-sonnet-5

# 2. EIS extraction — billable Anthropic batch (Sonnet 5); ~21.9k windows (~$110–115)
conda run -n nepa python phase2/code/deliverable02/04_extract_eis_significance.py \
  --sample 0 --batch-run --model claude-sonnet-5

# 3. validate both tracks against the AI-human reviewed gold (held-out test)
conda run -n nepa python phase2/code/deliverable02/05_validate_significance.py             # FONSI
conda run -n nepa python phase2/code/deliverable02/05_validate_significance.py --track eis  # EIS

# 4. analysis tables + figures (writes the CSVs/PNGs this report reads)
Rscript phase2/code/deliverable02/06_create_figures.R

# 5. render
quarto render phase2/reports/deliverable02.qmd
```

Outputs land in `phase2/data/analysis/deliverable02/` (determination parquets, validation metrics)
and `phase2/output/deliverable02/analysis/` (CSV tables + figures); the answer keys (AI-coded,
human-adjudicated) live under `…/deliverable02/gold/`. Schemas for every emitted parquet are
documented in `phase2/notes/deliverable02/data_dictionary.md`, and full provenance — input/output
paths, row counts, content hashes, model, prompt/schema versions — is recorded in
`significance_run_manifest.parquet`.
