---
title: "NEPA Decarbonization Technology Analysis: Deliverable 4"
subtitle: "Review Timelines for Categorical Exclusions, Environmental Assessments, and Environmental Impact Statements"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    toc-location: left
    fig-align: center
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(arrow)
library(dplyr)
library(tidyr)
library(readr)
library(stringr)
library(scales)
library(knitr)
library(gt)
library(here)

OUTPUT_DIR <- here("phase2", "output", "deliverable04")
FIG_DIR    <- file.path(OUTPUT_DIR, "figures")
DIAG_DIR   <- file.path(OUTPUT_DIR, "diagnostics")

# CATF brand navy (Phase 1 styling)
CATF_NAVY <- "#012169"

fmt_n   <- function(x) comma(x, accuracy = 1)
fmt_pct <- function(x, accuracy = 0.1) percent(x, accuracy = accuracy)
include_fig <- function(name) knitr::include_graphics(file.path(FIG_DIR, name))
read_diag <- function(name) read_csv(file.path(DIAG_DIR, name), show_col_types = FALSE)
kable_tbl <- function(x) knitr::kable(x, format = "pipe", align = "l")

# Navy-header gt table in the house style
navy_gt <- function(df, title = NULL) {
  g <- gt(df)
  if (!is.null(title)) g <- tab_header(g, title = title)
  g |>
    cols_align(align = "left") |>
    tab_style(style = list(cell_fill(color = CATF_NAVY),
                           cell_text(color = "white", weight = "bold", align = "left")),
              locations = cells_column_labels()) |>
    tab_options(table.font.size = px(12), table.width = pct(100))
}

# --- Coverage (complete = BOTH dates present, matching @fig-complete-share / @fig-coverage) ---
# Read the exact per-process complete-timeline shares written by 08_analyze.R so the narrative
# "% complete" numbers match Figure 2 (fig-complete-share) by construction. "Complete" = both an
# initiation and a decision date are present (date coverage), consistent with the figures and the
# Executive Summary's stated definition. A small share of projects have both dates but in invalid
# order (decision before initiation); those are excluded from the duration analysis as errors.
cov_complete <- read_diag("d4_complete_share.csv")
cpct <- function(pt) cov_complete$share_complete[cov_complete$process_type == pt]
cn   <- function(pt) cov_complete$n_complete[cov_complete$process_type == pt]

diag <- read_diag("d4_coverage_diagnostics.csv")
dpct <- function(pt, col) diag[[col]][diag$process_type == pt] / 100

# --- Durations (complete_clear) ---
dur <- read_diag("d4_duration_summary.csv")
dmed <- function(pt) dur$median_days[dur$process_type == pt]

# --- Pages (regulatory, all EA/EIS, all energy types) ---
pg <- read_diag("d4_pages_summary.csv")
pgv <- function(pt, per, col) pg[[col]][pg$process_type == pt & pg$fra_period == per]
```

```{=html}
<style>
figcaption, .figure-caption, .quarto-figure figcaption { text-align: left !important; }
</style>
```

## Executive Summary

::: callout-note
## Key Findings

-   **Review duration climbs steeply from days to months to years.** Median review duration is **`r dmed("CE")` days for CE**, **`r dmed("EA")` days (~`r round(dmed("EA")/30.4)` months) for EA**, and **`r fmt_n(dmed("EIS"))` days (~`r round(dmed("EIS")/365.25,1)` years) for EIS** — the expected complexity ladder, now quantified across the full database (complete timelines; @fig-duration-intervals).
-   **Timeline completeness varies sharply by process, and the gaps are structural.** Complete timelines (both an initiation and a decision date) reach **`r fmt_pct(cpct("CE"), 1)` for CE**, **`r fmt_pct(cpct("EA"), 1)` for EA**, and **`r fmt_pct(cpct("EIS"), 1)` for EIS** (@fig-coverage). CE and EA are *initiation*-limited (no Notice-of-Intent requirement, often no recorded start); EIS is *decision*-limited — Records of Decision are usually separate documents outside the corpus, so EIS decision coverage is just **`r fmt_pct(dpct("EIS","pct_decision"))`**. LLM adjudication of the ambiguous cases lifted complete coverage roughly **+7 points** per process.
-   **Post-FRA, EIS documents got shorter while EAs held steady.** Across **all `r fmt_n(sum(pg$n_projects))` EA/EIS projects** (all energy types, regulatory pages from the full text), mean EIS length fell from **`r pgv("EIS","Pre-FRA","mean_pages")` to `r pgv("EIS","Post-FRA","mean_pages")` pages** after the Fiscal Responsibility Act, while EA length stayed near **`r pgv("EA","Pre-FRA","mean_pages")`–`r pgv("EA","Post-FRA","mean_pages")` pages** — though the post-FRA window is still small (@sec-pages).
-   **The longest reviews are prime case-study candidates.** Genuinely protracted reviews — **SunZia Southwest Transmission (14.6 yr)**, **Energia Sierra Juarez**, **Grain Belt Express**, **Cushman Hydroelectric** — are surfaced for CATF investigation and separated from extraction errors (an "initiation" anchored to a historical citation), which are excluded (@sec-outliers).
-   **BLM field offices do not measurably speed up with experience.** The apparent office "learning curve" disappears once calendar time is controlled: the speed-up is a corpus-wide secular decline in CE durations, not individual offices getting faster as their caseload grows (@sec-field-office).
:::

This report delivers:

> Timelines for categorical exclusions, environmental assessments, and environmental impact statements — including segmentation by regulatory period (pre/post-FRA), energy category, and decision year — plus a timeline-outlier deliverable for case-study investigation and a document-length / FRA page-limit analysis refreshed on Phase 2 decision dates.

------------------------------------------------------------------------

# Methodology

This deliverable **extends the [Phase 1 timeline analysis](https://www.kaseyzapatka.com/nepa/phase1/reports/deliverable03.html#timeline-analysis-for-decarbonization-technologies)** — which examined review timelines for *decarb projects only* — to the **entire NEPATEC 2.0 database**, which includes decarb, fossil, and other project types. It does so with a rebuilt extraction pipeline that adds authoritative agency dates, section aware retrieval, and two machine-learning models.

## How timeline dates are extracted

This section improves on the Phase 1 timeline method. For every project we identify two milestones — the **initiation** (when the review begins: an application received, a Notice of Intent, or scoping) and the **decision** (when it concludes: a signature, a Record of Decision, or a Finding of No Significant Impact). The pipeline finds these in five high-level steps:

1.  **Extract candidate dates** from every NEPA document in NEPATEC 2.0, capturing the sentence around each date as well as document location and metadata so its meaning is clear.
2.  **Classify each date's role** — a machine-learning *classifier* labels every candidate as an `initiation`, `decision`, or `neither`.
3.  **Rank the candidates** — a second *ranking model* ranks initiation and decision candidates by their likely to be the final, decisive date for the project.
4.  **Anchor on authoritative dates** — where the BLM or DOE NEPA register records an official project-start or decision date, that date is used.
5.  **Resolve the hardest cases with an LLM** — for the projects where the operative date is still ambiguous, a large language model selects the best remaining candidate.

A project's timeline is **complete** when both an initiation and a decision date are found.

## How Phase 2 analysis improves upon Phase 1

-   **Full-database coverage** — Phase 1 covered decarbonization technology projects only (20,725); Phase 2 covers the entire NEPATEC 2.0 database (61,881 projects).
-   **API registry dates** — Phase 2 incorporates authoritative dates from *BLM and DOE NEPA registers* as an authoritative source of official project-start and decision dates. We pull in about 40,000 register dates, which is particularly helpful for CE and EA reviews.
-   **Cleaner retrieval and granularity handling** — section-aware retrieval pulls fewer but better candidate dates based on document location and the pipeline uses year-month decision dates (e.g. a Final-EIS cover month) where no better dates are available.
-   **Improved classifier *AND* a ranking model** — Phase 1 used a single date classifier (`DeBERTa` model); Phase 2 used an improve classifer (`SetFit`) and adds a dedicated **ranking model** (`LightGBM`) that decides which candidate is the operative initiation/decision, sharply improving which candidates get sent to the LLM adjudicating model.

## Data Quality & Caveats


### Data-quality issues resolved in this pass

- **Register vs document CE decisions.** Where a document-text CE decision disagreed with the authoritative BLM/DOE register determination by more than ~2 years (typically a stray historical date printed on a CX form), the pipeline now prefers the register — correcting ~650 CE decisions.
- **Negative-duration ordering.** Decision-before-initiation rows are now reclassified to `invalid_order` **at source** (scripts 05 / 05c), with an assertion guarding the analysis step against regressions — replacing the earlier downstream stopgap filter.

### Known issues, understood and deferred

None of the items below change the headline medians (which are robust to these tails), and each is flagged in the data so it can be filtered or audited:

- **Long-duration outliers (historical-citation initiations).** ~106 CE reviews (0.4%) show 13–45-year "durations" because the extractor selected a decades-old citation (an original grant, a covering PEIS, a prior plan) as the initiation rather than the recent application/renewal date. These are flagged `suspect_error` in the outlier deliverable (`10_outliers.R`) and excluded from case studies; a ranking fix (up-weighting the existing `historical_gap_candidate` flag) is deferred because it has no effect on the median.
- **EIS decision coverage.** ~58% of EIS still lack a decision date; most have a Final-EIS or ROD document in the corpus whose date was not extracted at retrieval. A targeted FEIS/ROD re-read is the highest-leverage follow-up and would lift EIS complete coverage from ~32% toward ~46%. Deferred to a separate pass.
- **Smaller register/document disagreements.** Sub-2-year CE decision disagreements are left to the ranker (low impact, deferred).
- **Classifier calibration.** Candidate-classifier confidence is currently calibrated in-sample; a dedicated out-of-sample calibration split is deferred (it affects internal confidence scores, not the selected dates).
- **Unrecoverable-PDF projects.** ~959 projects produced no extractable text at retrieval (scanned or corrupt PDFs) and carry no dates; not recoverable without OCR reprocessing.
- **Post-FRA CE durations.** Post-FRA CE reviews run modestly longer than pre-FRA (median ~37 vs ~19 days), shrinking further once proxy-flagged rows are excluded; the residual is spread across date sources rather than any single extraction bug.
- **End-to-end accuracy not yet formally validated.** Coverage and internal model metrics are reported; formal precision/recall against a hand-labeled, held-out gold set is a follow-up.


### Known data or timeline quality issues

- **Heavy reliance on proxy dates.** ~55% of CE and ~68% of EIS complete timelines rest on a *proxy* date that stands in for a milestone the documents never recorded directly. The proxy differs by process (below).
  - **CE — imputed initiations.** Most CEs record no distinct start, so the initiation is imputed: the pipeline takes the **earliest qualifying signature/start date** on the form (often the initiator's signature, which precedes the approving official's). ~13,450 CE initiations are proxies.
  - **EIS — Final-EIS-publication decisions.** Only ~18% of EISs have a Record of Decision in the corpus, so for another ~18% the pipeline uses the **Final-EIS publication date** (often reported as month + year → imputed to the 15th) as a stand-in decision. ~58% of EISs still end up with no decision date at all.
- **EAs are register-heavy.** EAs have no Notice-of-Intent requirement, so a documented start date is frequently missing — **54% of EA initiations come from the BLM/DOE register** rather than the EA document. Because those register "project start" dates are often **late administrative entries made near the decision** (some only 1–2 days before the FONSI), register-anchored EAs show artificially **short durations** (~60-day median vs ~12 months for document-based starts).
- **EA coverage is not governmentwide.** In NEPATEC 2.0, EA (and CE) intake is concentrated in DOE, BLM, and a small USDA/Forest Service contribution, whereas EPA supplies a broad, governmentwide EIS crawl. Every EA figure here therefore describes **EAs *in this corpus*** — heavily BLM/DOE — not the full federal EA population, and the corpus contains fewer EAs (3,083) than EISs (4,130), the inverse of the true national ratio.


::: callout-note
## Next steps

-   **Lift EIS decision coverage** — re-pull Final-EIS cover/publication dates to see if we can add more year-month dates for EIS.
-   **Extend the analysis** — the [BLM field-office experience](#sec-field-office) learning-curve test is now built (no experience effect once calendar time is controlled); geothermal timelines by field office and whether post-FRA adoption of categorical exclusions correlates with faster reviews (see Requested Extensions).

------------------------------------------------------------------------

:::

## Date provenance

Every extracted date traces to one of two sources: the **BLM/DOE NEPA registers** (official agency dates pulled via the registers' metadata API) or **document text** (parsed from the NEPA documents themselves, including LLM-adjudicated picks and Final-EIS publication dates). @fig-source shows the split by process and endpoint — a transparency check on where the timeline rests. Registers anchor a meaningful share of CE dates; EA and EIS rely overwhelmingly on document text.

```{r}
#| label: fig-source
#| fig-cap: "Source of each project's initiation and decision date by review process — Register API (BLM/DOE NEPA registers) vs Doc Text (parsed from documents) vs No Date (not extracted)."
#| out-width: "100%"
include_fig("fig_d4_register_source_projects.png")
```

------------------------------------------------------------------------

# Timeline Analysis

## Timeline completeness

@fig-complete-share shows the share of projects with complete timelines by review process; @fig-coverage breaks down *which* component is missing. The patterns differ by process:

-   **CE** (`r fmt_pct(cpct("CE"), 1)` complete): decision coverage is high (`r fmt_pct(dpct("CE","pct_decision"))`) but initiation coverage is low (`r fmt_pct(dpct("CE","pct_initiation"))`) — most CEs are single-signature determinations with no recorded start.
-   **EA** (`r fmt_pct(cpct("EA"), 1)` complete): the strongest of the three, but still initiation-limited — EAs have no Notice-of-Intent requirement and often skip scoping.
-   **EIS** (`r fmt_pct(cpct("EIS"), 1)` complete): initiations are usually documented (an NOI is required), but **decision** coverage is only `r fmt_pct(dpct("EIS","pct_decision"))` because Records of Decision are frequently separate documents not in the corpus.

**Comparison with Phase 1.** Phase 1 measured completeness for *decarb projects only* and reported [CE ~30% · EA ~62% · EIS ~48%](https://www.kaseyzapatka.com/nepa/phase1/reports/deliverable03.html#fig-complete-share). On the **full database**, Phase 2 reaches CE `r fmt_pct(cpct("CE"), 1)` · EA `r fmt_pct(cpct("EA"), 1)` · EIS `r fmt_pct(cpct("EIS"), 1)`. CE is *higher* — the much larger universe is dominated by routine, register-dated determinations — but **EA and EIS land below their Phase 1 decarb-only rates**, for two reasons this analysis surfaced:

-   **The aggregate is diluted by non-decarb projects.** Phase 1's denominator was decarb energy; Phase 2's is everything. Decarbonization EIS alone completes at **42.4%** (much closer to Phase 1's 48%) — it is the heterogeneous "Other" bucket that pulls the EIS aggregate down to `r fmt_pct(cpct("EIS"), 1)`.
-   **The remaining gaps are structural, not fixable by better parsing.** EIS completeness is capped by **missing Records of Decision** (a separate document, often outside the corpus — decision coverage only `r fmt_pct(dpct("EIS","pct_decision"))`); CE and EA completeness is capped by **missing initiations**, because neither requires a Notice of Intent and a start date is frequently never recorded. A minority of EIS "projects" are also comment letters or document fragments with no milestone date to extract. Lifting EIS coverage toward Phase 1 is the focus of the [Next steps](#methodology) above.

```{r}
#| label: fig-complete-share
#| fig-cap: "Share of projects with a complete timeline (both initiation and decision dates) by NEPA review process. Dot marks the completion rate."
#| out-width: "90%"
include_fig("fig_d4_complete_timeline_share_boxplot.png")
```

```{r}
#| label: fig-coverage
#| fig-cap: "Timeline coverage breakdown by review process — both dates, decision only, initiation only, or neither."
#| out-width: "100%"
include_fig("fig_d4_coverage_by_process.png")
```

Coverage also varies by energy category (@fig-cov-energy). Notably, **decarbonization EIS projects complete at 42.4% — well above the EIS aggregate of `r fmt_pct(cpct("EIS"), 1)`** — because decarb EISs are better documented than the heterogeneous "Other" bucket that drags the aggregate down.

```{r}
#| label: fig-cov-energy
#| fig-cap: "Timeline coverage by process and energy category (both dates / decision only / initiation only / none)."
#| out-width: "100%"
include_fig("fig_d4_coverage_by_process_and_energy.png")
```

## Duration analysis

Durations use **all complete timelines** (both dates present, including those resolved with a proxy date such as a Final-EIS publication date in place of a missing Record of Decision). Month-only dates are imputed to the **mid-month 15th**.

On that basis, durations track the expected complexity ladder from CE to EA to EIS — a median of about **`r dmed("CE")` days for CE**, **`r dmed("EA")` days (~`r round(dmed("EA")/30.4)` months) for EA**, and **`r fmt_n(dmed("EIS"))` days (~`r round(dmed("EIS")/365.25,1)` years) for EIS** (n = `r fmt_n(dur$n[dur$process_type=="CE"])` / `r fmt_n(dur$n[dur$process_type=="EA"])` / `r fmt_n(dur$n[dur$process_type=="EIS"])`). @fig-duration-intervals summarises the spread; @fig-duration-hist shows the full distributions; @fig-spans plots individual project spans.

```{r}
#| label: fig-duration-intervals
#| fig-cap: "Timeline duration by review process. Thin bar = 10th–90th percentile; thick bar = interquartile range (25th–75th); point = median. All complete timelines."
#| out-width: "100%"
include_fig("fig_d4_duration_summary_intervals.png")
```

Splitting duration by energy category (@fig-dur-energy) exposes a striking artifact: **fossil EAs run far shorter than decarbonization or "other" EAs** — a median of ~40 days, vs ~310 days for Decarb and ~209 days for Other. This is not because fossil EAs are genuinely faster: **84% of them take their start date from the BLM register** rather than the document. This is a broad convention across routine BLM oil-and-gas EAs — drilling permits, lease sales, and well-abandonments alike (well-abandonment is 53% of complete fossil EAs, not the whole story) — where the recorded "project start" is a *late administrative entry made within days of the decision*. Because fossil is the **largest** EA group (754 of the ~1,730 complete EAs), those artificially short spans **pull the overall EA median down**. Restricting fossil EAs to the document-anchored starts lifts their median to **~5 months** — much closer to the Other and Decarb EAs and the more defensible review-length estimate — but that view survives for only a fraction of them.[^ea-doc-anchored]

```{r}
#| label: fig-dur-energy
#| fig-cap: "Timeline duration by process and energy category. Fossil EAs are short largely because most take register-anchored start dates (late administrative entries near the decision); the EA panel adds a second, magenta **Fossil (doc-anchored)** row — the ~5-month median for the 119 fossil EAs with a document start date — as the more defensible review-length view."
#| out-width: "100%"
include_fig("fig_d4_duration_summary_intervals_by_energy.png")
```

[^ea-doc-anchored]: The EA panel's magenta **Fossil (doc-anchored)** row recomputes the fossil-EA duration using only *document-stated* initiation dates (`initiation_source_type ≠ register`), setting aside the BLM register's administrative "project start" entries that compress the register-anchored median to ~40 days. Its ~5-month median is the more defensible estimate of true review length, but it survives for only **119 of the 754** complete fossil EAs — the register is the sole start-date signal for the other 635. Both rows are shown so the coverage-versus-plausibility trade-off is explicit; CE and EIS are unaffected by this issue and keep their three standard energy rows.

The full duration distributions (@fig-duration-hist) show the same complexity ladder as a *shape*, not just a median: **CE durations pile up near zero** (most are days-to-weeks determinations), **EA spreads out to a year or two**, and **EIS is broad and heavy-tailed** — a meaningful mass of reviews runs 3–10+ years. That long EIS tail is real (large, contested reviews spanning multiple administrations) and is examined directly in the outlier case studies below.

```{r}
#| label: fig-duration-hist
#| fig-cap: "Review duration distribution by process; all complete timelines (month-only dates imputed to mid-month)."
#| out-width: "100%"
include_fig("fig_d4_duration_histogram.png")
```


@fig-spans plots each complete review as a horizontal bar from its initiation to its decision date, sorted by length within each process so the spread in durations is visible one project at a time rather than only in aggregate. CE spans collapse to short marks, EA bars range from a few months to a couple of years, and the EIS panel is dominated by multi-year bars — the same CE-to-EA-to-EIS complexity ladder seen in the histograms, now resolved to individual reviews.

```{r}
#| label: fig-spans
#| fig-cap: "Individual project timelines from initiation to decision, sorted by duration within process (complete timelines, sampled)."
#| out-width: "100%"
include_fig("fig_d4_project_timeline_spans.png")
```

## Projects by decision year

@fig-by-year shows project volume by decision year (all projects), faceted by review process. CE volumes ramp from the late-2000s (ARRA-era funding) with a second rise into the 2020s (BIL/IRA); EIS decisions lag CE and EA by several years, reflecting longer review cycles. @fig-by-year-doe restricts to the **Department of Energy** — the single largest agency in the dataset — where the ARRA/BIL/IRA grant-and-loan cycles are even more pronounced.

```{r}
#| label: fig-by-year
#| fig-cap: "Reviews by decision year (all projects), faceted by NEPA review process."
#| out-width: "100%"
include_fig("fig_d4_projects_by_decision_year.png")
```

```{r}
#| label: fig-by-year-doe
#| fig-cap: "DOE projects by decision year, faceted by review process (Department of Energy projects only)."
#| out-width: "100%"
include_fig("fig_d4_projects_by_decision_year_doe.png")
```

## Timeline Outliers: Case-Study Candidates {#sec-outliers}

Per the deliverable request, timeline outliers are surfaced for CATF staff to investigate via case studies (including whether NEPA itself was a cause of delay). Outliers are init→decision spans **longer than 5,000 days (~13.7 years)**, produced reproducibly by `code/deliverable04/10_outliers.R`.

**The critical distinction is real vs. extraction error.** A multi-year span can be a genuinely protracted review *or* an artifact where the "initiation" was anchored to a historical citation (a facility's construction date, a reserve's founding, a prior plan, a regulatory-compliance milestone) rather than the NEPA start. The two are separable only by reading the evidence text, which the outlier table preserves.

::: callout-note
## How to read these (process matters)

-   **EIS outliers are mostly real.** EIS median duration is already ~3 years; a 14–17-year EIS sits in the legitimate tail (transmission, restoration, resource-management plans genuinely run that long).
-   **CE outliers are mostly errors** — CEs are fast determinations (median `r dmed("CE")` days), so a multi-year CE almost always reflects a mis-extracted initiation. CE outliers are therefore excluded from the client list.
-   **Initiation dates before ~1990 are a strong error signal.** Of the four EIS outliers with 1980s-or-earlier initiations, three are confirmed errors (the "initiation" is a park founding, a prior lock's construction-completion date, or a RCRA compliance milestone).
:::

The decarb EA/EIS candidates below are the highest-value case studies. **SunZia** (ROW application Sep 2008 → Record of Decision Apr 2023) is the textbook long-permitting saga; **Energia Sierra Juarez**, **Grain Belt Express**, and **Cushman Hydroelectric** are similar. *Confirmed errors are excluded* — e.g. the Palisades Nuclear Restart "19-year" span glues a 2005 license-renewal application to a 2024 restart Record of Decision (two different NEPA actions).

```{r}
#| label: tbl-outliers
#| tbl-cap: "Genuinely long EA/EIS reviews — case-study candidates. Full list with evidence text in output/deliverable04/diagnostics/d4_duration_outliers_client.csv. Project IDs are full UUIDs."
client <- read_diag("d4_duration_outliers_client.csv")
# Curated exclusions: confirmed cross-action extraction errors found on evidence review.
exclude_ids <- c("da6bbc7a3ca3aeca7bc1c06eaafa8d44")  # Palisades restart (license-renewal init)
client |>
  filter(!project_id %in% exclude_ids) |>
  arrange(desc(duration_years)) |>
  transmute(`Project` = project_title,
            `Process` = process_type,
            `Energy` = project_energy_type,
            `Lead agency` = str_remove_all(lead_agency_harmonized, '\\[|\\]|"'),
            `Years` = round(duration_years, 1),
            `Initiation` = as.character(initiation_date),
            `Decision` = as.character(decision_date),
            `project_id` = project_id) |>
  head(18) |>
  navy_gt()
```

------------------------------------------------------------------------

# Document Length & FRA Page Limits {#sec-pages}

This section **builds directly on [Phase 1 Deliverable 5](https://www.kaseyzapatka.com/nepa/phase1/reports/deliverable05.html)** ("Document Length Over Time and Fiscal Responsibility Act Impact"), which analyzed document length for *decarb* EA/EIS projects. Here we reproduce that analysis on the **entire Phase 2 EA/EIS corpus, all energy types**, refreshed on Phase 2 LLM-adjudicated **decision dates** (inclusion requires only a decision date — the time axis and FRA classification depend only on it). Each subsection below links to its Phase 1 counterpart for comparison.

::: callout-note
## Method & scope

-   **Regulatory pages** = body word count ÷ 500, excluding embedded appendices and low-content pages, per **40 C.F.R. § 1508.1(bb)** — the measure that matches the FRA limits (EA ≤ 75 pages; EIS ≤ 150, or 300 if extraordinarily complex).[^rawpages]
-   These counts are computed **from the actual page text** by `fra/01_extract_pages.py` (a DuckDB pass over `phase2/data/processed/{ea,eis}/pages.parquet` — 6.1M EIS pages — detecting the appendix boundary and low-content pages), producing regulatory pages for **5,032 EA/EIS projects** (2,765 EA, 2,267 EIS) across **all energy types** — not just decarbonization.
-   The pre/post-FRA analysis below uses the **`r fmt_n(sum(pg$n_projects))` projects** that also have a decision date (EA `r fmt_n(sum(pg$n_projects[pg$process_type=="EA"]))` / EIS `r fmt_n(sum(pg$n_projects[pg$process_type=="EIS"]))`).
-   **FRA date = enactment (June 3, 2023)**, matching Phase 1 D5 and the timeline-duration period split above.
:::

[^rawpages]: We use *regulatory* pages rather than raw PDF page counts because raw counts substantially overstate the length the statute limits — they include embedded appendices and sparse pages (covers, maps, dividers). Mean raw pages run roughly 2× regulatory pages for EAs and higher for EISs, so a raw count cannot be compared to the 75/150-page limits.

## Document length over time

@fig-pages-time tracks regulatory page length by decision date across all EA/EIS. Two patterns stand out. **EA length sits low and essentially flat** across the decade — the rolling mean hugs the low tens of pages, with only occasional long outliers — consistent with EAs being short by design. **EIS length is far higher and far more dispersed**: individual EISs run from tens of pages to well over a thousand, and the rolling mean hovers in the low-to-mid hundreds with no strong secular trend before the FRA. The post-FRA window (right of the red line) is denser — more recent decisions — and is where the EIS rolling mean steps down.

Compared with the **decarb-only [Phase 1 version](https://www.kaseyzapatka.com/nepa/phase1/reports/deliverable05.html#fig-pages-over-time)**, two differences follow from the wider Phase 2 scope. First, the **all-energy EA cloud sits well below Phase 1's**: Phase 1's decarb EAs averaged ~62 regulatory pages, whereas the full corpus averages ~33, because the large pool of short fossil and well-abandonment EAs (median ~7 pages) pulls the all-energy mean down. Second, the **EIS picture is the same in shape** — high, heavy-tailed, and dominated by project-to-project variation rather than a clear time trend — confirming Phase 1's caution that average trends obscure substantial spread, now borne out on a roughly 5× larger sample.

```{r}
#| label: fig-pages-time
#| fig-cap: "Regulatory page counts for individual EA/EIS projects (all energy types) by decision date, coloured by FRA period; navy line = 3-month rolling mean. Red dashed line = FRA enactment (June 3, 2023)."
#| out-width: "100%"
include_fig("fig_d4_pages_over_time.png")
```

## Pre vs post-FRA

Across all EA/EIS, **EIS regulatory length fell after FRA** (mean `r pgv("EIS","Pre-FRA","mean_pages")` → `r pgv("EIS","Post-FRA","mean_pages")` pages; median `r pgv("EIS","Pre-FRA","median_pages")` → `r pgv("EIS","Post-FRA","median_pages")`), while **EA length was essentially flat** (mean `r pgv("EA","Pre-FRA","mean_pages")` → `r pgv("EA","Post-FRA","mean_pages")`). The post-FRA window is short, so these are provisional (n = `r pgv("EA","Post-FRA","n_projects")` EA, `r pgv("EIS","Post-FRA","n_projects")` EIS). The **direction matches the decarb-only Phase 1 result closely** ([Phase 1 D5 — Pre vs Post FRA](https://www.kaseyzapatka.com/nepa/phase1/reports/deliverable05.html#fig-pre-post) and its [distribution view](https://www.kaseyzapatka.com/nepa/phase1/reports/deliverable05.html#fig-distribution)): Phase 1's EISs fell from a mean of **368 → 270** regulatory pages, and the all-energy corpus here falls **317 → 269** — landing at essentially the **same ~270-page post-FRA EIS mean** despite a very different sample. EA stayed flat in both (Phase 1 **62 → 57**; here **33 → 39**), reflecting that EAs were already well under the 75-page limit and had little to compress.

```{r}
#| label: fig-pages-bar
#| fig-cap: "Mean regulatory pages, pre- vs post-FRA. Bar = mean; diamond = median; n shown."
#| out-width: "100%"
include_fig("fig_d4_pages_pre_post_fra.png")
```

Breaking the same comparison out **by energy category** (@fig-pages-energy) shows the post-FRA EIS decline is **broad** — it appears in decarbonization *and* fossil EIS, not just one segment — while EA length is essentially flat across all categories. That consistency makes the EIS drop more credible as a genuine post-FRA effect rather than a composition artifact.

```{r}
#| label: fig-pages-energy
#| fig-cap: "Mean regulatory pages by energy category, pre- vs post-FRA. Labels = n projects."
#| out-width: "100%"
include_fig("fig_d4_pages_pre_post_fra_by_energy.png")
```

The full distributions (@fig-pages-dist) confirm the shift is not an outlier effect: the **EIS distribution moves down** post-FRA (lower median, thinner upper tail) while the **EA distribution barely moves**. Splitting that distribution by energy category — shown separately for decarbonization (@fig-pages-dist-decarb), fossil (@fig-pages-dist-fossil), and other (@fig-pages-dist-other) projects so each panel is legible — shows the EIS contraction holds in both decarbonization and fossil EIS, while EA length stays flat across all three.

```{r}
#| label: fig-pages-dist
#| fig-cap: "Distribution of regulatory page counts, pre- vs post-FRA (violin + box; diamond = median; y capped at p99)."
#| out-width: "100%"
include_fig("fig_d4_pages_distribution.png")
```

```{r}
#| label: fig-pages-dist-decarb
#| fig-cap: "Decarbonization EA/EIS — distribution of regulatory page counts, pre- vs post-FRA (violin + box; diamond = median; y capped at p99)."
#| out-width: "100%"
include_fig("fig_d4_pages_distribution_decarb.png")
```

```{r}
#| label: fig-pages-dist-fossil
#| fig-cap: "Fossil EA/EIS — distribution of regulatory page counts, pre- vs post-FRA (violin + box; diamond = median; y capped at p99)."
#| out-width: "100%"
include_fig("fig_d4_pages_distribution_fossil.png")
```

```{r}
#| label: fig-pages-dist-other
#| fig-cap: "Other EA/EIS — distribution of regulatory page counts, pre- vs post-FRA (violin + box; diamond = median; y capped at p99)."
#| out-width: "100%"
include_fig("fig_d4_pages_distribution_other.png")
```

## Descriptive statistics

@tbl-pages reports the full descriptive statistics — mean, median, standard deviation, and interquartile range — by process and FRA period across all energy types (the all-energy counterpart to [Phase 1 D5's descriptive-statistics table](https://www.kaseyzapatka.com/nepa/phase1/reports/deliverable05.html#tbl-summary)).

```{r}
#| label: tbl-pages
#| tbl-cap: "Descriptive statistics for regulatory page counts by process type and FRA period — all EA/EIS, all energy types. Regulatory pages = body word count ÷ 500, excluding embedded appendix pages and low-content pages, per 40 C.F.R. § 1508.1(bb)."
pg |>
  select(process_type, fra_period, n_projects, mean_pages, median_pages,
         sd_pages, p25_pages, p75_pages) |>
  mutate(
    process_type = factor(process_type, levels = c("EA", "EIS")),
    fra_period   = factor(fra_period, levels = c("Pre-FRA", "Post-FRA"))
  ) |>
  arrange(process_type, fra_period) |>
  gt(groupname_col = "process_type") |>
  cols_label(
    fra_period   = "FRA Period",
    n_projects   = "N",
    mean_pages   = "Mean",
    median_pages = "Median",
    sd_pages     = "SD",
    p25_pages    = "P25",
    p75_pages    = "P75"
  ) |>
  cols_align(align = "left",  columns = fra_period) |>
  cols_align(align = "right", columns = where(is.numeric)) |>
  fmt_number(columns = where(is.numeric), decimals = 0, use_seps = TRUE) |>
  tab_spanner(
    label = "Regulatory Pages (body word count ÷ 500)",
    columns = c(mean_pages, median_pages, sd_pages, p25_pages, p75_pages)
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_style(
    style = list(
      cell_fill(color = CATF_NAVY),
      cell_text(color = "white", weight = "bold", size = px(13))
    ),
    locations = cells_row_groups()
  ) |>
  tab_style(
    style = cell_fill(color = "#f0f7ff"),
    locations = cells_body(rows = fra_period == "Post-FRA")
  ) |>
  tab_footnote(
    footnote = "Post-FRA projects are those with a decision date on or after June 3, 2023. Regulatory pages exclude embedded appendix pages and low-content pages (maps, figures, blanks); documents whose filename already omits appendices use their page count directly, and documents without extractable text are excluded."
  ) |>
  tab_style(
    style = cell_text(align = "right"),
    locations = cells_footnotes()
  ) |>
  tab_options(
    table.font.size = px(13),
    data_row.padding = px(6),
    row_group.padding = px(6)
  )
```

## FRA page-limit compliance

```{r}
#| label: compliance-vals
comp <- read_diag("d4_pages_compliance.csv")
compv <- function(pt, cat, col) comp[[col]][comp$process_type == pt & comp$compliance == cat]
ea_tot      <- comp$total[comp$process_type == "EA"][1]
ea_ok_pct   <- round(compv("EA", "Compliant", "pct"));     ea_ex_pct  <- round(compv("EA", "Exceeds limit", "pct"))
eis_tot     <- comp$total[comp$process_type == "EIS"][1]
eis_ok_pct  <- round(compv("EIS", "Compliant", "pct"))
eis_std_pct <- round(compv("EIS", "Exceeds standard limit", "pct"))
eis_over_pct<- round(compv("EIS", "Exceeds limit", "pct"))
eis_w300    <- compv("EIS", "Compliant", "n") + compv("EIS", "Exceeds standard limit", "n")
eis_w300_pct<- round(100 * eis_w300 / eis_tot)
```

The Fiscal Responsibility Act sets presumptive page limits — **75 pages for an EA** and **150 pages for an EIS** (300 if the agency documents extraordinary complexity). @fig-pages-compliance scores the post-FRA EA/EIS projects with a decision date against those limits (regulatory pages).

-   **EAs comply at a high rate.** `r ea_ok_pct`% of post-FRA EAs (n = `r fmt_n(ea_tot)`) fall within the 75-page limit; the `r ea_ex_pct`% that exceed it generally do so modestly rather than dramatically.
-   **EISs are more mixed, but most are within reach of the complexity ceiling.** Only `r eis_ok_pct`% of post-FRA EISs (n = `r fmt_n(eis_tot)`) land within the baseline 150-page limit, but a further `r eis_std_pct`% sit between 150 and 300 pages — i.e. within the extraordinary-complexity ceiling — so **`r eis_w300_pct`% are within the 300-page threshold overall**. The remaining `r eis_over_pct`% exceed even 300 pages, consistent with the long right tail of complex transmission, restoration, and resource-management EISs seen in the duration analysis.

Two caveats temper this. First, the **post-FRA window is short** (`r fmt_n(ea_tot)` EA / `r fmt_n(eis_tot)` EIS projects with both a regulatory page count and a decision date), so these rates are provisional and will firm up as more post-FRA decisions land. Second, the FRA limits are **presumptive, not absolute** — agencies may exceed them with senior-official approval or a documented extraordinary-complexity finding — so "exceeds limit" flags a document for review, not an automatic compliance failure.

```{r}
#| label: fig-pages-compliance
#| fig-cap: "FRA page-limit compliance among post-FRA projects (regulatory pages). EA limit 75; EIS 150 (300 if extraordinarily complex). The bracket marks the EIS share within the 300-page extraordinary-complexity threshold."
#| out-width: "90%"
include_fig("fig_d4_pages_compliance.png")
```

------------------------------------------------------------------------

# Review Timelines by CEQ Regulatory Regime {#sec-ceq-regime}

::: callout-warning
## Proposed section — awaiting your feedback

This section is a **draft proposal, not yet built.** It would test whether review durations shifted as the **CEQ NEPA implementing regulations** (40 C.F.R. §§ 1500–1508) were rewritten across administrations — a different breakpoint from the FRA *statutory* split used above. Please review the framing and planned figures and tell me whether to build it.
:::

NEPA's procedural rules were rewritten three times in four years. The natural regulatory regimes — keyed to each rule's effective date — are summarized below. (The **FRA**, June 3, 2023, is a *statutory* amendment to NEPA, not a CEQ rule; it is already the pre/post-FRA breakpoint used in the duration and page-length analyses.)

```{r}
tibble::tribble(
  ~Regime,                      ~`Effective`,    ~`What changed`,
  "1978 regulations",           "1978–2020",     "Original CEQ regulations; essentially unchanged for ~42 years.",
  "2020 Final Rule (Trump)",    "Sep 14, 2020",  "Major rewrite: presumptive 2-yr EIS / 1-yr EA limits, page limits, narrowed effects/cumulative-impacts definitions.",
  "2022 Phase 1 Rule (Biden)",  "May 20, 2022",  "Restored 1978 purpose-and-need and cumulative/indirect effects.",
  "2024 Phase 2 Rule",          "Jul 1, 2024",   "Implemented the FRA's statutory NEPA amendments (codified page/time limits, CE sharing, lead-agency roles).",
  "2025 rescission",            "early 2025",    "After EO 14154 and the Marin Audubon ruling, CEQ rescinded its regulations; agencies revert to their own NEPA procedures."
) |> navy_gt()
```

**Planned analysis.** The pipeline already carries the scaffolding for this — `08_analyze.R` computes a `reg_period` segmentation and writes `d4_duration_by_period.csv`, currently keyed to *funding* events (ARRA / BIL / IRA). We would re-point those cut dates to the CEQ rule effective dates (2020-09-14, 2022-05-20, 2024-07-01) and surface two views:

-   **Median duration by regulatory regime** — a per-process interval/bar chart comparing median review length across the 1978 / 2020 / 2022 / 2024 regimes (from the already-computed `d4_duration_by_period.csv`).
-   **Duration trend by decision year with regime markers** — the year-by-year median-duration line (`fig_d4_duration_trend.png`, already rendered) with vertical markers at the three CEQ rule dates, replacing or augmenting the current ARRA/BIL/IRA markers.

**Key caveat to flag up front.** Duration is measured at the **decision** date, but a multi-year review mostly occurs under *earlier* rules — so this reads as "reviews **decided under** a regime," not "reviews **governed by**" it. Post-2020 samples are also thin and confounded with the coverage ramp (the same provisional-sample limitation as the post-FRA split).

**Feedback requested:** (1) Is "reviews decided under a regime" the right framing, or should we anchor on the *initiation* date instead? (2) Keep the FRA statutory marker alongside the CEQ rule markers, or present them separately? (3) Which view do you want — the regime comparison, the trend-with-markers, or both?

------------------------------------------------------------------------

# BLM field-office experience {#sec-field-office}

This extension answers a request from the deliverable scope: **do BLM field offices process NEPA reviews faster as they accumulate experience** — a "learning-curve" effect? We map each BLM-led project to the field office that handled it (from the structured `DOI-BLM-<state>-<office>-<year>-<seq>` case number), order each office's reviews by the office's own cumulative count, and test whether duration falls as that count rises — **after** separating the experience signal from two confounds that would otherwise manufacture a false result: the register-anchoring artifact and secular calendar-time trends. All conclusions here are **associational, not causal**.

```{r}
#| label: fo-setup
fo_cov  <- read_diag("d4_fieldoffice_parse_coverage.csv")
fo_covv <- function(scope, col) fo_cov[[col]][fo_cov$scope == scope]
fo_off  <- arrow::read_parquet(here("phase2", "data", "analysis", "deliverable04", "blm_field_offices.parquet"))
n_offices  <- dplyr::n_distinct(fo_off$office_code)
n_case     <- sum(fo_off$parse_source == "case_number")
n_file     <- sum(fo_off$parse_source == "file_name")

fo_mod <- read_diag("d4_fieldoffice_model.csv")
mrow   <- function(proc, anch) fo_mod[fo_mod$process == proc & fo_mod$anchor == anch, ]
ce_d   <- mrow("CE", "Document")   # PRIMARY: document-anchored
ce_r   <- mrow("CE", "Register")   # sensitivity
ea_r   <- mrow("EA", "Register")   # exploratory

fo_sum      <- read_diag("d4_fieldoffice_summary.csv")
n_q_offices <- nrow(fo_sum)
med_first   <- median(fo_sum$first_med, na.rm = TRUE)
med_last    <- median(fo_sum$last_med,  na.rm = TRUE)
share_sped  <- mean(fo_sum$pct_change < 0, na.rm = TRUE)

sgn  <- function(x) sprintf("%+.1f%%", x)                                  # signed percent
ci   <- function(r) sprintf("%+.1f%% to %+.1f%%", r$pct_lo, r$pct_hi)      # signed 95% CI
```

::: callout-note
## Method & scope

-   **Field-office parse.** The validated regex `BLM-([A-Z]{2}-?[A-Z]?[0-9]{2,4})` on the case number (file name as fallback) maps **`r fmt_pct(fo_covv("ALL","pct_parsed")/100)`** of BLM-led projects (`r fmt_n(fo_covv("ALL","parsed"))` of `r fmt_n(fo_covv("ALL","blm_projects"))`) to one of **`r n_offices`** field offices (`r fmt_n(n_case)` from the structured case number, `r fmt_n(n_file)` from the file name).
-   **Document-anchored durations are primary.** BLM register "project start" dates are frequently late administrative entries (the ~40-day artifact documented in the duration analysis above), so the learning curve is computed on **document-anchored** initiation dates (`initiation_source_type ≠ metadata`); the register-anchored version is a clearly-flagged secondary view.
-   **EA and CE are analyzed separately** (CE medians run in weeks, EA in months); **EIS is excluded** (too sparse per office). Offices qualify with **≥ 30** valid durations on the analysis measure.
-   **Model.** A pooled regression with **office fixed effects**, `log(duration) ~ log(cum_count) + factor(decision_year) + project_energy_type`, fit separately for EA and CE. The `log(cum_count)` coefficient is the experience effect *net of* secular calendar trends; `factor(decision_year)` is the calendar control, `project_energy_type` the project-mix control.
:::

::: callout-important
## Finding: the apparent learning curve is a calendar-time confound, not experience

Descriptively, CE reviews look like they speed up as an office gains experience: across the **`r n_q_offices`** offices with ≥ 30 document-anchored CE reviews, the median duration of an office's *earliest* third of reviews is **`r med_first` days** versus **`r med_last` days** for its *most-experienced* third, and **`r fmt_pct(share_sped)`** of offices sped up (@fig-fo-dumbbell). **But that gradient is an artifact of calendar time.** CE durations have fallen sharply over the 2010s corpus-wide, and because an office's higher-experience reviews are also its later-calendar reviews (within-sample correlation ≈ 0.67), the raw pattern conflates the two (@fig-fo-confound). The office fixed-effects regression that adds a **decision-year control** dissolves the effect: the coefficient on `log(cumulative reviews)` is **`r sgn(ce_d$pct_per_doubling)` per doubling (95% CI `r ci(ce_d)`)** for the primary document-anchored CE measure (n = `r fmt_n(ce_d$n_reviews)` reviews, `r ce_d$n_offices` offices) — statistically indistinguishable from zero and, if anything, the *opposite* sign — even though the **same model without the calendar control** reproduces the apparent **`r sgn(ce_d$pct_per_doubling_raw)`** speed-up. The register-anchored CE sensitivity points the same way (`r sgn(ce_r$pct_per_doubling)`, CI `r ci(ce_r)`). **EA cannot support the office-level design**: no office clears the ≥ 30 document-anchored-review bar (only 144 such EA durations exist corpus-wide), and the exploratory register-anchored EA estimate (`r sgn(ea_r$pct_per_doubling)`, CI `r ci(ea_r)`) is not distinguishable from zero and rests on artifact-prone start dates. **Net of calendar time, there is no evidence that BLM field offices grow faster with experience.**
:::

@fig-fo-curve shows the raw, *uncontrolled* gradient that motivates the question — and that the regression then explains away. Document-anchored CE durations decline across the cumulative-review deciles; the register-anchored lines sit far lower (especially for EA), a direct picture of the late-administrative-entry artifact.

```{r}
#| label: fig-fo-curve
#| fig-cap: "Apparent field-office learning curve (uncontrolled): median review duration by the office's cumulative-review decile, EA and CE panels, document- vs register-anchored. Descriptive only — not net of calendar time; the regression below shows the decline is secular calendar drift, not experience."
#| out-width: "100%"
include_fig("fig_d4_fieldoffice_learning_curve.png")
```

The confound is direct in @fig-fo-confound: median CE duration falls with **both** calendar year and cumulative experience, because the two move together (later reviews are later in time). Only the regression, which holds decision-year fixed, can separate them — and once it does, the within-office experience coefficient is null/positive.

```{r}
#| label: fig-fo-confound
#| fig-cap: "Why the curve is a calendar confound: CE durations fall with both calendar year (left) and cumulative office experience (right). The two gradients look alike because they are collinear (r ≈ 0.67); the office + decision-year fixed-effects regression attributes the decline to calendar time, leaving no experience effect."
#| out-width: "100%"
include_fig("fig_d4_fieldoffice_experience_vs_calendar.png")
```

Per-office heterogeneity (@fig-fo-dumbbell) reinforces the null: some high-volume offices show shorter durations for their most-experienced third, others longer — there is no consistent within-office speed-up once each office is its own baseline. Because a later tercile is also a later calendar period, most of the leftward movement that does appear is the same secular decline, not learning.

```{r}
#| label: fig-fo-dumbbell
#| fig-cap: "First- vs last-tercile median CE duration (document-anchored) for the top 15 BLM field offices by review volume. Light dot = least-experienced third; navy dot = most-experienced third. Descriptive only; the direction is inconsistent across offices and largely tracks calendar time."
#| out-width: "100%"
include_fig("fig_d4_fieldoffice_first_vs_last.png")
```

```{r}
#| label: tbl-fo-model
#| tbl-cap: "Learning-curve regression: effect of doubling an office's cumulative review count on review duration, with and without the decision-year (calendar) control. Negative = faster with experience. The sign flip between the two columns is the calendar confound. CE document-anchored is the primary measure."
fo_mod |>
  filter(!is.na(n_reviews), n_reviews > 0) |>
  transmute(
    Process = process,
    Measure = dplyr::recode(anchor, Document = "Document-anchored", Register = "Register-anchored"),
    `Min reviews/office` = threshold,
    `N reviews` = n_reviews,
    `N offices` = n_offices,
    `%/doubling (no calendar control)` = sprintf("%+.1f%%", pct_per_doubling_raw),
    `%/doubling (calendar-controlled)` = sprintf("%+.1f%%", pct_per_doubling),
    `95% CI` = sprintf("%+.1f to %+.1f%%", pct_lo, pct_hi),
    Role = dplyr::case_when(
      process == "CE" & anchor == "Document" ~ "Primary",
      process == "CE" & anchor == "Register" ~ "Sensitivity",
      TRUE ~ "Exploratory (artifact-prone)")
  ) |>
  navy_gt() |>
  tab_footnote(footnote = "EA has no office with ≥30 document-anchored durations (144 corpus-wide), so the EA row uses the register-anchored measure at a relaxed ≥10 threshold and is exploratory only.")
```

::: callout-warning
## Caveats

-   **Calendar-time confound (the headline).** The raw "offices speed up with experience" pattern is secular calendar drift; net of decision-year there is no experience effect. This is the central caveat — the descriptive figures must not be read on their own.
-   **Register-anchoring artifact.** Durations rest partly on register/proxy start dates that are late administrative entries; the primary analysis uses document-anchored dates to limit this, but the secondary register views inherit the artifact (visible as the much shorter register lines in @fig-fo-curve).
-   **~38% of BLM projects are unparsed** (`r fmt_pct(1 - fo_covv("ALL","pct_parsed")/100)`). EISs account for most of the gap — only `r fmt_pct(fo_covv("EIS","pct_parsed")/100)` parse (programmatic/large EISs are filed under titles, not DOI-BLM case numbers), versus `r fmt_pct(fo_covv("CE","pct_parsed")/100)` of CE and `r fmt_pct(fo_covv("EA","pct_parsed")/100)` of EA. Many unparsed EAs carry an office-like code in a non-standard form (bare `C069-2023-…`, `AZ-A010-…`) lacking the `BLM-` prefix the regex anchors on, and are left out rather than guessed.
-   **Project mix.** Short well-abandonment EAs dominate some offices; `project_energy_type` is included as a covariate, but a "faster" office could reflect a shift toward simpler actions rather than learning.
-   **Office identity** may shift over time (renames, mergers, district reorganizations); a stable `office_code` is assumed.
:::

------------------------------------------------------------------------

# Requested Extensions

The following items from the deliverable request require additional analysis beyond the current timeline build. They are scoped here for the next phase:

```{r}
tibble::tibble(
  `Requested analysis` = c(
    "Geothermal timelines by BLM field office",
    "CE adoption post-FRA vs. review speed"
  ),
  `What it needs` = c(
    "Geothermal project flags (Deliverable 1/3 tech tags) × field office × duration; limited by geothermal sample size per office.",
    "CE-adoption events (which agencies adopted other agencies' CEs post-FRA, BLM/DOE/USFS) × subsequent review duration; needs the CE-adoption register."
  )
) |>
  navy_gt()
```

------------------------------------------------------------------------

# Reproduction

Run from the repository root in the `nepa` conda environment:

``` bash
# Timeline pipeline (candidates -> classify -> select -> LLM adjudicate)
conda run -n nepa python phase2/code/deliverable04/run_pipeline.py
ANTHROPIC_API_KEY=$(security find-generic-password -s nepa-anthropic -w) \
  conda run -n nepa python phase2/code/deliverable04/06_adjudicate_llm.py \
  --mode candidate_adjudication --process CE EA EIS --model claude-haiku-4-5-20251001 --workers 24

# Analysis + figures
Rscript phase2/code/deliverable04/08_analyze.R       # coverage, durations, FRA, energy
Rscript phase2/code/deliverable04/10_outliers.R      # timeline outliers (case-study candidates)

# FRA document-length analysis (regulatory pages from the full page text, all EA/EIS)
conda run -n nepa python phase2/code/deliverable04/fra/01_extract_pages.py --run
Rscript phase2/code/deliverable04/fra/02_pages_fra.R # document length / FRA page limits

# BLM field-office experience / learning-curve extension
conda run -n nepa python phase2/code/deliverable04/field_office/01_parse_offices.py --run
Rscript phase2/code/deliverable04/field_office/02_learning_curve.R
```

Outputs are written to `phase2/output/deliverable04/` (`figures/`, `diagnostics/`).

------------------------------------------------------------------------

*Draft generated `r Sys.Date()` | NEPA Decarbonization Technology Analysis — Phase 2, Deliverable 4*
