---
title: "NEPA Decarbonization Technology Analysis: Deliverable 2"
subtitle: "Determinations of Significance Across Resource Areas"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    toc-location: left
    fig-align: center
execute:
  echo: false
  warning: false
  message: false
---

<!-- Plan: phase2/plans/deliverable02.md (v2.11). Pipeline: phase2/code/deliverable02/
     00 regime -> 01 corpus+cohorts -> 02 FONSI extract -> 04 EIS extract -> 05 validate ->
     06 analyze.R (writes the CSV tables this report reads). Regenerate 06 before rendering. -->

```{r setup}
library(arrow); library(dplyr); library(tidyr); library(readr)
library(stringr); library(knitr); library(gt); library(here)
options(arrow.skip_nul = TRUE)

A   <- here("phase2", "data", "analysis", "deliverable02")
TAB <- here("phase2", "output", "deliverable02", "analysis")
rp  <- function(f) read_parquet(file.path(A, f))
rc  <- function(f) { p <- file.path(TAB, f); if (file.exists(p)) read_csv(p, show_col_types = FALSE) else tibble() }
fmt <- function(x) formatC(x, format = "d", big.mark = ",")

det    <- rp("significance_determinations.parquet")
corpus <- rp("significance_corpus.parquet")
regime <- rp("project_regime.parquet")

is_dry <- all(det$extraction_method == "regex")
```

::: {.callout-warning}
## Draft status
```{r}
#| results: asis
cat(if (is_dry)
  paste("This document outlines the deliverable's purpose and the analysis workflow; the",
        "findings themselves are still in progress. Results tables will be added once the",
        "extraction pass and the validation against the hand-labelled gold set are complete.")
  else paste("This report reflects the validated extraction; see Methods & required caveats",
             "for the applicable limitations."))
```
:::

This report delivers:

> Determinations of significance across resource areas; factors that contribute to a determination of "significant impact," starting with mitigated FONSIs.

---

## Context

When a clean-energy project — a solar farm, a transmission line — needs federal approval, the
lead agency (here, BLM or the Department of Energy) studies its environmental impact and makes a
judgment call: **is the impact "significant" or not?**

That one word decides a lot. If the agency finds the impact **not significant**, the project
moves ahead on a faster track (a document called a *Finding of No Significant Impact*, or FONSI).
If it finds the impact **significant**, that triggers a much longer, more demanding review.

Often the agency lands in between: *"this impact **would** be significant — **unless** the
developer commits to specific measures to reduce it."* That is a **mitigated finding**, and it is
the most revealing case, because it shows the exact line the agency drew and what it required to
stay on the safe side of it.

**So this deliverable answers:** across these clean-energy projects, how do agencies decide what
counts as "significant" in each environmental area (air, water, wildlife, cultural sites, and so
on) — and what mitigation do they demand to avoid that label? Answering it means reading and
coding a judgment call out of **thousands** of dense decision documents — far more than anyone
could review by hand. The workflow below is how we do that reliably.

## Workflow

The approach pairs the **scale of automation** with the **reliability of human judgment**:
software and AI do the reading at a scale no person could, while analysts define the scope, hold
the checkpoints, and grade the AI against their own answer key before any result counts.

```{=html}
<div style="max-width:660px;margin:1.25rem auto;">
  <div style="border:2px solid #0047BB;border-radius:8px;padding:.65rem 1rem;background:#f4f7fc;">
    <div style="font-weight:700;color:#012169;font-size:1.05rem;">1 · Scope</div>
    <div style="color:#1A1A2E;">Pick the comparable projects and note when each was decided</div>
  </div>
  <div style="text-align:center;color:#0047BB;font-size:1.5rem;line-height:1.1;">&#8595;</div>
  <div style="border:2px solid #0047BB;border-radius:8px;padding:.65rem 1rem;background:#f4f7fc;">
    <div style="font-weight:700;color:#012169;font-size:1.05rem;">2 · Gather documents</div>
    <div style="color:#1A1A2E;">Collect the agencies&rsquo; decision documents</div>
  </div>
  <div style="text-align:center;color:#0047BB;font-size:1.5rem;line-height:1.1;">&#8595;</div>
  <div style="border:2px solid #0047BB;border-radius:8px;padding:.65rem 1rem;background:#f4f7fc;">
    <div style="font-weight:700;color:#012169;font-size:1.05rem;">3 · Extract the judgment calls</div>
    <div style="color:#1A1A2E;">Rules flag the passages &rarr; AI classifies each &rarr; analysts hand-label a sample as the answer key</div>
  </div>
  <div style="text-align:center;color:#0047BB;font-size:1.5rem;line-height:1.1;">&#8595;</div>
  <div style="border:2px solid #0047BB;border-radius:8px;padding:.65rem 1rem;background:#f4f7fc;">
    <div style="font-weight:700;color:#012169;font-size:1.05rem;">4 · Validate</div>
    <div style="color:#1A1A2E;">Score the AI against the human answer key before anything is reported</div>
  </div>
  <div style="text-align:center;color:#0047BB;font-size:1.5rem;line-height:1.1;">&#8595;</div>
  <div style="border:2px solid #012169;border-radius:8px;padding:.65rem 1rem;background:#012169;">
    <div style="font-weight:700;color:#ffffff;font-size:1.05rem;">5 · Analyze and report</div>
    <div style="color:#ffffff;">Turn validated findings into this report and a source-linked data table</div>
  </div>
  <div style="text-align:center;color:#6b7280;font-size:.85rem;margin-top:.5rem;">
    The five steps of the analysis &mdash; each card matches a numbered step below.
  </div>
</div>
```

**Step by step:**

1. **Scope.** Fix a comparable set of projects — clean energy, led by BLM or the Department of
   Energy — and note *when* each decision was made, because the environmental rules changed over
   the years. This keeps every comparison apples-to-apples.
2. **Gather documents.** Collect the actual decision documents where the significance calls
   live — FONSIs (especially the *mitigated* ones) and the larger environmental impact statements.
3. **Extract the judgment calls.** This is the engine, in three parts: fast **rules** scan the
   documents and flag the passages that look like a significance judgment; **AI** then reads each
   flagged passage and classifies it (significant, not significant, or mitigated; which
   environmental area; any regulatory threshold); and in parallel **analysts hand-label a sample**
   by eye to serve as an *answer key*.
4. **Validate.** Compare the AI's answers to the human answer key. If the AI is accurate enough,
   the results move forward; if not, we refine and run it again. The AI is never taken on faith.
5. **Analyze and report.** Only the validated results feed the findings — the significance
   patterns by resource area, the mitigation trends, the shifts across regulatory eras and
   agencies — delivered as this report **and** a data table where every number links back to the
   exact page it came from.

**Why the results are trustworthy:** every determination is traceable to the exact document,
page, and passage; the AI is scored against human labels before it is used; and the project set
is scoped so comparisons are fair.

## Methods & required caveats

- **Agency scope.** BLM + DOE agency family. Headline denominators use only
  `agency_scope_status = 'primary_blm_doe_family'`; other-agency and manual-review rows are a
  separate context universe.
- **D4 date/regime coverage.** Many projects lack a high-confidence decision date; regime cells
  are confidence-gated and thin post-2023. EIS decision dates are frequently proxies.
- **EPA §309 letters** are out of scope — this report makes no EPA-review claim.
- **Clean-energy scope only** (`project_energy_type = 'Clean'`).
- **FONSI coverage** for the mitigated/straight tiers derives from D6 artifacts.
- **EIS retrieval limits.** The EIS tier is gated behind FONSI validation and a retrieval-recall
  spike; EIS findings are preliminary.
- **Validation sample size.** Rare class/resource/threshold cells are reported descriptively,
  not as pass/fail macro-metrics.
- **Suppressed cells.** Any cell below 5 is suppressed.

## Appendix: data dictionary

See `phase2/notes/deliverable02/data_dictionary.md` and the run manifest
`significance_run_manifest.parquet` (input+output paths, row counts, content hashes, model,
prompt/schema versions).
